### 2024-07-11 doc/manuscript.Rmd
methods

---
title: "Soil drivers of plant, fungal and bacterial community novelty in urban habitats of the city of Xixón, Spain"
output: 
  bookdown::github_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

Running title: Urban drivers of plant, fungal and bacterial novelty

Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^2^ (<https://orcid.org/0000-0002-0915-0916>),
Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)

^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain

Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain. Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}. Telephone: +34985104781.

# Author contributions

Eduardo Fernández-Pascual: Conceptualization; Methodology; Investigation; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing. Zuzana Ferencova: Conceptualization; Methodology; Investigation; Data Curation; Writing – Review & Editing. Víctor González-García: Investigation; Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.

# Acknowledgements

María García Fernández, Sara Gutiérrez Cruz and Abelardo Sigüenza Solís helped with the field surveys.

# Funding

Fundación Biodiversidad, Ministerio para la Transición Ecológica y el Reto Demográfico, Gobierno de España [XIXÓN ECORRESILIENTE, PRTR, NextGenerationEU]. Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico [SV-23-GIJON-JBA].

# Conflict of interest statement

None declared.

# Data availability statement

Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at Zenodo: https://zenodo.org/doi/10.5281/zenodo.11613571.

# Introduction

In post-industrial landscapes affected by the phenomenon of urban contraction, land located in the heart of cities remains unused for long periods (Haase et al., 2014) . When the use of any urban plot ceases, it is colonized by plants from the regional flora capable of living in urban environments (Johnson et al., 2017) . These urban plants are adapted to disturbances and can grow in degraded soils; some even accumulate heavy metals, making them nature-based solutions for bioremediation of degraded lands (Song et al., 2019) . Species traits, along with the legacy of human impact, determine which regional flora species colonize which plots (Swan et al., 2021) . When the human legacy is intense, the plant communities that develop will be novel communities, different from the historical reference communities that existed before human intervention (Gedan and Fernández-Pascual, 2019) . The novelty of the communities can be reflected in their floristic, compositional and phylogenetic composition, including their proportion of exotic species (Lososová et al., 2012) . In any case, proper management of urban communities can turn them into resources with high ecological and social potential (Anderson and Minor, 2017). (Kowarik, 2018) : in the Ibero-Atlantic ecoregion, anthropogenic habitats can host up to a third of the ecoregional flora.

The ability of plants to colonize any habitat, including urban ones, depends largely on their seed regeneration niche, that is, on their dispersal, persistence, germination and establishment functions (Grubb, 1977, Saatkamp et al., 2019) . However, although it is known that urban species often share a series of functional and ecological traits (Aronson et al., 2007, Lososová et al., 2006) , little is known about their seed regeneration niche. Especially, information on its germination traits is lacking. In species living in non-anthropogenic environments, germination phenology is regulated by environmental factors (water, temperature, light) to ensure that germination occurs at the right time to maximize subsequent seedling survival (Carta et al., 2022a) . Compared to the large amount of data on the germination of species from non-anthropogenic habitats, little is known about what factors regulate the germination of urban species. Considering other habitats affected by humans (Fernández-Pascual et al., 2022) , it could be expected that urban seeds have their germination unrelated to macroclimatic signals and show tolerance to stress due to heat, drought or heavy metals. However, this hypothesis needs to be tested through a synthesis of the seed regeneration niche in urban environments.

Cities 
1. highly fragmented landscapes
2. mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use)
3. subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems
4. comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure
5. contains a multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix

Cities are comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure. They are mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use) which are subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems (Kowarik, 2005).

The “Four Nature approach” narrows down the variety of transformational stages in urban settings to four major types that can supply wilderness in urban region (Kowarik 1992, 2005). Each of the four types relates to nature in general but results from different trajectories in human-nature interactions:

Spontaneous vegetation appears in different habitat types in the urban matrix,. Multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix. The spontaneous herbs could provide the very interesting pool of species for ecological restoration of the urban area. Their ecological value has been clear since many years because their contribution to the ecosystem services in urban areas. Although their aesthetic value has been highlighted repeatedly, their acceptance by the residents in urban areas has been questioned. Many residents associate the presence of spontaneous plants with the neglect in maintenance by public (municipal) management service. Spontaneous native urban vegetation has commonly been described as demonstrating resilience (Ignatieva et al., 2000) and exhibiting adaptations to human disturbance (Lundholm and Marlin, 2006; Sukopp, 2004).

Urban species pool is comprised of native species present before the urban development, native species developed because the new urban ecosystem and introduced species by human activity. However, just some of them can adapt to urban environment.

IGS (informal green space) consists of any urban space with a history of strong anthropogenic disturbance that is covered at least partly with non-remnant, spontaneous vegetation (Del Tredici, 2010).

# Materials and methods

## Study city

Gijón/Xixón is a medium-sized coastal city (population c. 270,000 inhabitants) located in the Bay of Biscay, within the central Asturias metropolitan area of northern Spain (population c. 800,000 inhabitants). The city belongs to the Cantabrian Mixed Forests ecoregion [@RN4667], at the border between the temperate and Mediterranean climatic zones of Europe, and has a temperate oceanic climate with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm). Most of the city is built on alluvial and estuarine deposits, sand, or limestone bedrock. The potential pre-human vegetation would be dominated by deciduous mixed forests, together with riparian forests, wetlands and coastal dunes. The area has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. The city itself has been inhabited at least since Roman times [@RN5532]. During the medieval and early modern era the urban center was limited to the original walled city, with fishing as the main economic activity. Expansion beyond the original core started in the 18th and 19th centuries. Exposure to global trade networks and the mild climate turned the region into a hotspot for biological invasions, especially from the Americas and Asia [@RN5493]. In 1852, the city was connected by railway and became a major port for the regional mining and industrial centers. During the 20th century the city experienced significant industrial development (e.g. canning, coal, steel, shipbuilding) and rapid population growth, becoming the most populated city in the Principality of Asturias [@RN5534; @RN5533]. Industrial and urban growth brought environmental problems including loss of natural wetlands and air and watercourse pollution [@RN5535]. Like the rest of its metropolitan area, the city has been affected by recent post-industrial land-use changes which have left large areas with abandoned industrial sites (i.e. brownfields) [@RN5448; @RN5449].

## Definition of urban habitats

To survey urban biodiversity across habitats, we defined six habitats types (**Figure 1A**) representing different states within the four-nature classification of urban wilderness [@RN5445]: (1) forest patches < 10 km from the city center, i.e. type-1 nature in remnants of the pre-urban vegetation; (2) hay meadows, i.e. type-2 nature in remnants of human-made ecosystems resulting from agricultural habitat transformation; (3) parks, i.e. type-3 nature in managed green spaces created after habitat destruction; (4) roadsides, i.e. type-4 nature in relatively unmanaged strips of land alongside roads, railways, etc; (5) residential lots, i.e. type-4 nature in unmanaged vacant lots of previous residential use; and (6) industrial lots, i.e. type-4 nature in unmanaged vacant lots of previous  industrial use.

## Selection of sampling plots

During June 2023, we surveyed the city to establish five 4 x 4 m sampling plots per each of the six habitat types (n = 30 plots). The plots were distributed across the city (**Figure 1B**) and represented the variation of vegetation types found within each habitat type. In two plots per habitat (**Figure 1B**), we buried a datalogger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil. Dataloggers recorded temperature once every hour and stayed on site for approximately one year, after which we retrieved them and downloaded their records to describe the thermal environment of the city and the habitats.

## Soil physicochemical sampling

From each of the 30 plots, we collected 500 g of soil for analysis of chemical composition and physical soil properties. We took the samples from 5 points in each plot (center + 4 cardinal points). We removed stones and other debris particles, sieved the soil in a 5 mm mesh and homogenized the sample in a plastic bag. We cleaned and sterilized all tools with water and ethanol between samples. We sent the samples for analysis at INEA - Escuela Universitaria de Ingeniería Agrícola, Universidad Pontificia de Comillas. Soil analysis included pH, texture, bulk density, electrical conductivity, organic matter content, total nitrogen, C/N ratio, active limestone, total carbonates, assimilable phosphorus, K, Ca, Mg, Na and heavy metals (Zn, Cr, Cu, Cd, Hg, Pb, Ni).

> Añadir detalles de los métodos de suelo (Zuzana)

## Plant, fungal and bacterial diversity sampling

In each plot, we recorded all the spontaneous vascular plant species, as well as the cover (%) of each species in the plot. For plant species names we followed Euro+Med [@RN5044]. Additionally, from each of the 30 plots, we collected soil samples for DNA metabarcoding analysis of soil fungal and bacterial communities. We performed the sampling in sterile conditions, i.e. before entering the plot for any of the other tasks. In each plot, we took samples from the four cardinal points, 0.5 m inwards from the edge, at a depth of 5 cm. Before taking the samples, we removed the vegetation with a sterile tool. Then, we collected the soil samples in a 50 ml Falcon tube, mixed them, homogenized them with a lancet and finally stored the sample into a 2 ml Eppendorf tube which we immediately froze. We sent the frozen tube to AllGenetics & Biology SL (www.allgenetics.eu) for DNA metabarcoding analysis.

## DNA metabarcoding analysis

DNA analysis by AllGenetics included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size < 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, taxonomic assignment, number of reads per taxon and generation of rarefaction curves.

*DNA isolation and quantification*. DNA was isolated from each soil sample using the DNeasy PowerSoil Pro DNA isolation kit (Qiagen) following the manufacturer’s instructions, and DNA was eluted in a final volume of 50 μL. In one sample the above-mentioned protocol did not yield a DNA extract of enough concentration, so DNA was isolated using the FastDNA Spin Kit for Soil (MPBio) following the manufacturer’s instructions, and DNA was eluted in a final volume of 70 μL. An extraction blank (Bex) was included in each round of DNA extraction, and treated as the plot samples to check for contamination. DNA concentration in each extract was quantified using the Qubit High Sensitivity dsDNA Assay (Thermo Fisher Scientific).

*DNA metabarcoding library preparation and sequencing*. For fungal library preparation, a fragment of the nuclear ITS1 region of around 280-300 bp (including the primer sequences) was amplified using the following primers: forward - ITS1f (5’ CTTGGTCATTTAGAGGAAGTAA 3’) [@RN5548]; reverse - ITS2 (5’ GCTGCGTTCTTCATCGATGC 3’) [@RN5549]. For prokaryotic library preparation, a fragment of the 16S rRNA gene of around 300 bp (including primer sequences) was amplified using the following primers: forward - 515F-Y (5’ GTGYCAGCMGCCGCGGTAA 3’) [@RN5536]; reverse - 806RB (5’ GGACTACNVGGGTWTCTAAT 3’) [@RN5537]. These primers also included the Illumina sequencing primer sequences attached to their 5’ ends. In the first amplification step, PCRs were carried out in a final volume of 12.5 μL, containing 1.25 μL of template DNA (1:2 diluted in PCR round 3; not diluted in PCR round 11), 0.5 μM of the primers, 3.13 μL of Supreme NZYTaq 2x Green Master Mix (NZYTech), and ultrapure water up to 12.5 μL. The fungal reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 35 cycles of 95 ºC for 30 s, 49 ºC for 45 s, 72 ºC for 45 s, and a final extension step at 72 ºC for 7 min. The bacterial reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 25 cycles of 95 ºC for 30 s, 47 ºC for 45 s, 72ºC for 45 s, and a final extension step at 72 ºC for 7 min. The oligonucleotide indices that are required for multiplexing different libraries in the same sequencing pool were attached in a second amplification step with identical conditions but only 5 cycles and 60 ºC as the annealing temperature. A negative control that contained no DNA (BPCR) was included in every PCR round to check for contamination during library preparation. Library size was verified by running the libraries on 2 % agarose gels stained with GreenSafe (NZYTech) and imaging them under UV light. Then, libraries were purified using the Mag-Bind RXNPure Plus magnetic beads (Omega Bio-tek), following the instructions provided by the manufacturer. Finished libraries were pooled in equimolar amounts according to the results of a Qubit dsDNA HS Assay (Thermo Fisher Scientific) quantification. The pool was sequenced in a fraction of a NovaSeq PE250 flow cell (Illumina) aiming for a total output of 2 gigabases. Any potential traces of adapter dimers were removed using Cutadapt v3.5 [@RN5538]. Quality of the sequencing data was assesed with the software FastQC [@RN5539].

*Processing of sequencing data and inference of ASVs*. Cutadapt v3.5 was used to trim off non-biological DNA sequences (primers, indices, and sequencing adapters). Then, amplicon reads were processed using the DADA2 pipeline [@RN5541] implemented in QIIME 2 (release 2023.7) [@RN5540]. The first step in the DADA2 pipeline was to remove PCR primers and filter reads according to their quality. After checking the read quality profiles, forward reads were truncated at position 175, and reverse reads at position 131. The next step was to denoise and infer Amplicon Sequence Variants (ASVs). Error rates were learned from the dataset to denoise, using the parametric error model implemented in DADA2. Dereplication of the dataset (i.e. combination of all identical reads into unique reads) was carried out to reduce computational effort. These dereplicated forward and reverse reads were used to infer ASVs with the core sample inference algorithm [@RN5541]. Subsequently, corresponding R1 and R2 reads were merged into pairs with a minimum overlap of 12 identical base pairs. The final step in the DADA2 pipeline was to reduce the impact of artefacts (chimaeras) in the dataset. The final output of the DADA2 pipeline was a table containing the count of reads of every observed ASV in each sample, as well as a table listing all ASVs and their corresponding representative sequences.

*Taxonomic assignment*. Taxonomic assignment of each ASV was performed using a pre-trained classifier of the UNITE reference database for fungi [@RN5550] (release May 2021); and of the SILVA reference database for bacteria [@RN5542] (release 138.1 August 2020). To compare the representative sequences of the ASVs to the reference database, and compute the taxonomic assignment,the algorithm sklearn [@RN5543] implemented in QIIME 2 was applied, setting a minimum confidence level of 0.8. Several filters were applied to exclude: (1) singletons (i.e. ASVs containing only one member sequence in the whole data set); (2) ASVs occurring at a frequency below 0.01 % in each sample, to reduce the mistagging bias [@RN5544]; (3) from the bacterial dataset, non-prokaryotic ASVs such as eukaryotic sequences of plastid [@RN5545] and mitochondrial [@RN5546] origin; (4) those assigned only at domain (Bacteria) or kingdom (Fungi) level; (5) ASVs present in the negative controls that were more abundant in the negative controls than in the samples, indicating contamination [@RN5547].

*Alpha rarefaction plots*. Alpha rarefaction plots showed that the sequencing depth to reach saturation was 30,000 reads for the fungi and 22,000 reads for the bacteria, above the sequencing depth of all plots for fungi (mean = 114,919 reads; min = 79,045; max = 157,110) and bacteria (mean = 113,341 reads; min = 84,094; max = 171,404).

> Rarefaction

## Statistical analysis

# Results

## Soil temperatures and properties across habitats

## Plant, fungal and bacterial diversity across habitats

> PERMANOVAs

> Analysis with species, genera and ASVs

> Analysis with presence absence

## Community novelty across habitats and taxonomic groups

## Soil drivers of novelty

# Discussion

The recorded flora represents just a subset of possible spontaneous vegetation because this study was carried on during limited time which covers just part of the growing season of the whole vegetation.

There are some other habitats in the city which were not included in the present study, e.g. wetlands, beach, cliffs.

### 2024-07-11 src/Pbtolerance.R

@@ -1,0 +1,36 @@
+library(tidyverse)
+
+c("#FFA500",  
+  "yellow", 
+  "#B3EE3A", 
+  "#40E0D0",
+  "#5CACEE", 
+  "#A020F0",
+  "#551A8B") -> colores
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/traits.csv", fileEncoding = "latin1") -> traits
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+merge(soils, species) %>%
+  filter(Pb > 150) %>%
+  merge(traits) %>%
+  filter(origin == "Native") %>%
+  group_by(taxon) %>%
+  summarise(tolerance = weighted.mean(Pb, cover)) %>%
+  arrange(-tolerance) %>%
+  data.frame %>%
+  pull(taxon) -> tolerantspp
+
+species %>%
+  filter(taxon %in% tolerantspp) %>%
+  group_by(taxon) %>%
+  summarise(n = length(taxon), cover = mean(cover)) %>%
+  filter(n > 5) %>%
+  arrange(-cover) %>%
+  data.frame

### 2024-07-11 src/bacteria.R

@@ -1,0 +1,77 @@
+library(tidyverse)
+
+read.csv("data/bacteria/Table3/table_after_filtering.csv", sep = "\t", 
+         #fileEncoding = "latin1", 
+         skip = 1) %>%
+  separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%
+  mutate(kingdom = gsub("d__", "", kingdom), 
+         phylum = gsub("p__", "", phylum), 
+         class = gsub("c__", "", class), 
+         order = gsub("o__", "", order), 
+         family = gsub("f__", "", family), 
+         genus = gsub("g__", "", genus), 
+         species = gsub("s__", "", species), 
+         species = gsub("_", " ", species)) %>%
+  rename(ASV = X.OTU.ID,
+         Xixon01 = X1.16S,
+         Xixon02 = X2.16S,
+         Xixon03 = X3.16S,
+         Xixon04 = X4.16S,
+         Xixon05 = X5.16S,
+         Xixon06 = X6.16S,
+         Xixon07 = X7.16S,
+         Xixon08 = X8.16S,
+         Xixon09 = X9.16S,
+         Xixon10 = X10.16S,
+         Xixon11 = X11.16S,
+         Xixon12 = X12.16S,
+         Xixon13 = X13.16S,
+         Xixon14 = X14.16S,
+         Xixon15 = X15.16S,
+         Xixon16 = X16.16S,
+         Xixon17 = X17.16S,
+         Xixon18 = X18.16S,
+         Xixon19 = X19.16S,
+         Xixon20 = X20.16S,
+         Xixon21 = X21.16S,
+         Xixon22 = X22.16S,
+         Xixon23 = X23FD.16S,
+         Xixon24 = X24.16S,
+         Xixon25 = X25.16S,
+         Xixon26 = X26.16S,
+         Xixon27 = X27.16S,
+         Xixon28 = X28.16S,
+         Xixon29 = X29.16S,
+         Xixon30 = X30.16S) -> bacteria
+
+bacteria %>% filter(is.na(kingdom))
+bacteria %>% filter(is.na(phylum))
+bacteria %>% filter(is.na(class))
+bacteria %>% filter(is.na(order))
+bacteria %>% filter(is.na(family))
+bacteria %>% filter(is.na(genus))
+
+bacteria %>% pull(kingdom) %>% na.omit %>% unique -> kingdoms 
[…]

### 2024-07-11 src/compositional-novelty.R

@@ -1,0 +1,210 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") %>%
+  filter(habitat != "N4D") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(species) %>%
+  select(id, taxon, cover) %>%
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 4) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus N1 and N2 together)
+
+sitescores %>%
+  filter(habitat %in% c("N1", "N2")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  gather(Reference, Centroid, N1:N2) %>%
+  arrange(id, Trait, Reference) %>%
+  mutate(Novelty = Centroid - Value) %>%
+  group_by(id, Trait) %>%
+  filter(Novelty == min(Novelty)) %>%
+  group_by(id, habitat) %>%
+  summarise(Novelty = sum(Novelty)) %>%
+  group_by(habitat) %>%
+  summarise(Novelty = mean(Novelty)) 
+
+### Novelty (versus N1)
+
+sitescores %>%
+  filter(habitat %in% c("N1")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
[…]

### 2024-07-11 src/dca.R

@@ -1,0 +1,245 @@
+library(tidyverse); library(vegan)
+
+c("#FFA500",  
+  "yellow", 
+  "#B3EE3A", 
+  "#40E0D0",
+  "#5CACEE", 
+  "#A020F0",
+  "#551A8B") -> colores
+
+read.csv("data/header.csv", fileEncoding = "latin1") %>%
+  filter(habitat != "N4D") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### PCA soils
+
+header %>% select(id, habitat) %>% merge(soils) -> datos
+datos %>% select(-c(id, habitat)) %>% FactoMineR::PCA(graph = FALSE) -> pca1
+cbind((datos %>%  dplyr::select(id, habitat)), data.frame(pca1$ind$coord[, 1:2])) -> pcaInds
+pca1$var$coord[, 1:2] %>% data.frame %>% rownames_to_column(var = "Variable") -> pcaVars
+
+ggplot(pcaInds, aes(x = Dim.1, y = Dim.2)) +
+  coord_fixed() +
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  geom_vline(xintercept = 0, linetype = "dashed") +
+  geom_segment(data = pcaVars, aes(x = 0, y = 0, xend = 3*Dim.1, yend = 3*Dim.2)) +
+  geom_point(aes(fill = habitat), color = "black", size = 7, shape = 21) +
+  ggrepel::geom_label_repel(data = pcaVars, aes(x = 3*Dim.1, y = 3*Dim.2, label = Variable),  show.legend = FALSE, size = 2) +
+  ggthemes::theme_tufte() + 
+  ggtitle(label = "PCA of soil properties") +
+  theme(text = element_text(family = "sans"),
+        legend.position = "right", 
+        legend.title = element_blank(),
+        legend.text = element_text(size = 12, color = "black"),
+        panel.background = element_rect(color = "black", fill = NULL),
+        axis.title = element_text(size = 12),
+        axis.text = element_text(size = 12, color = "black"),
[…]

### 2024-07-11 src/diversity.R

@@ -1,0 +1,145 @@
+library(tidyverse)
+
+c("#FFA500",  
+  "yellow", 
+  "#B3EE3A", 
+  "#40E0D0",
+  "#5CACEE", 
+  "#A020F0",
+  "#551A8B") -> colores
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/traits.csv", fileEncoding = "latin1") -> traits
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+# Aliens
+
+merge(species, traits) %>%
+  group_by(id, origin) %>%
+  summarise(c = sum(cover)) %>%
+  spread(origin, c, fill = 0) %>%
+  mutate(aliencover = 100 * (Alien / (Alien + Native))) %>%
+  merge(header) %>%
+  group_by(habitat) %>%
+  summarise(m = mean(aliencover))
+
+merge(species, traits) %>%
+  group_by(id, origin) %>%
+  summarise(n = length(taxon)) %>%
+  spread(origin, n, fill = 0) %>%
+  mutate(alienspp = 100 * (Alien / (Alien + Native))) %>%
+  merge(header) %>%
+  group_by(habitat) %>%
+  summarise(nalien = mean(alienspp))
+
+merge(species, traits) %>%
+  merge(header) %>%
+  select(habitat, taxon, origin) %>%
+  unique %>%
+  group_by(habitat, origin) %>%
+  tally %>%
+  spread(origin, n, fill = 0) %>%
+  mutate(aliens = 100 * (Alien / (Alien + Native))) -> alienspp
+
+# Alpha
+
+species %>% group_by(id) %>% summarise(plantR = length(taxon)) %>% data.frame -> r1
+fungi %>% group_by(id) %>% summarise(fungiR = length(taxon)) %>% data.frame-> r2
+bacteria %>% group_by(id) %>% summarise(bacteriaR = length(taxon)) %>% data.frame -> r3
+
+# Gamma
+
[…]

### 2024-07-11 src/fungi.R

@@ -1,0 +1,89 @@
+library(tidyverse)
+
+read.csv("data/fungi/Table3/table_after_filtering.csv", sep = "\t", 
+         #fileEncoding = "latin1", 
+         skip = 1) %>%
+  separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%
+  mutate(kingdom = gsub("k__", "", kingdom), 
+         phylum = gsub("p__", "", phylum), 
+         class = gsub("c__", "", class), 
+         order = gsub("o__", "", order), 
+         family = gsub("f__", "", family), 
+         genus = gsub("g__", "", genus), 
+         species = gsub("s__", "", species), 
+         species = gsub("_", " ", species)) %>%
+  rename(ASV = X.OTU.ID,
+         Xixon01 = X1.ITS,
+         Xixon02 = X2.ITS,
+         Xixon03 = X3.ITS,
+         Xixon04 = X4.ITS,
+         Xixon05 = X5.ITS,
+         Xixon06 = X6.ITS,
+         Xixon07 = X7.ITS,
+         Xixon08 = X8.ITS,
+         Xixon09 = X9.ITS,
+         Xixon10 = X10.ITS,
+         Xixon11 = X11.ITS,
+         Xixon12 = X12.ITS,
+         Xixon13 = X13.ITS,
+         Xixon14 = X14.ITS,
+         Xixon15 = X15.ITS,
+         Xixon16 = X16.ITS,
+         Xixon17 = X17.ITS,
+         Xixon18 = X18.ITS,
+         Xixon19 = X19.ITS,
+         Xixon20 = X20.ITS,
+         Xixon21 = X21.ITS,
+         Xixon22 = X22.ITS,
+         Xixon23 = X23FD.ITS,
+         Xixon24 = X24.ITS,
+         Xixon25 = X25.ITS,
+         Xixon26 = X26.ITS,
+         Xixon27 = X27.ITS,
+         Xixon28 = X28.ITS,
+         Xixon29 = X29.ITS,
+         Xixon30 = X30.ITS) -> fungi
+
+fungi %>% filter(is.na(kingdom))
+fungi %>% filter(is.na(phylum))
+fungi %>% filter(is.na(class))
+fungi %>% filter(is.na(order))
+fungi %>% filter(is.na(family))
+fungi %>% filter(is.na(genus))
+
+fungi %>% pull(kingdom) %>% na.omit %>% unique -> kingdoms 
[…]

### 2024-07-11 src/merge-novelty.R

@@ -1,0 +1,29 @@
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+header %>%
+  select(id, habitat) %>%
+  merge(tpn) %>%
+  merge(tfn) %>%
+  merge(tbn)  %>%
+  merge(cpn1) %>%
+  merge(cpn2) %>%
+  merge(cfn1) %>%
+  merge(cfn2) %>%
+  merge(cbn1) %>%
+  merge(cbn2) %>%
+  merge(soils) %>%
+  select(-c(habitat, id)) %>%
+  FactoMineR::PCA()
+  
+
+header %>%
+  select(id, habitat) %>%
+  merge(tpn) %>%
+  merge(tfn) %>%
+  merge(tbn)  %>%
+  merge(cpn1) %>%
+  merge(cfn1) %>%
+  merge(cbn1) %>%
+  merge(soils) %>%
+  select(-c(habitat, id)) %>%
+  FactoMineR::PCA()

### 2024-07-11 src/novelty.R

@@ -1,0 +1,305 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") %>%
+  filter(habitat != "N4D") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(species) %>%
+  select(id, taxon, cover) %>%
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 4) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus N1)
+
+sitescores %>%
+  filter(habitat %in% c("N1")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = N1 - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(CPN1 = sum(abs(Novelty)))  -> cpn1
+
+### Novelty (versus N2)
+
+sitescores %>%
+  filter(habitat %in% c("N2")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = N2 - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(CPN2 = sum(abs(Novelty))) -> cpn2
+
+### Compositional novelty for fungi
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(fungi) %>%
+  select(id, taxon, cover) %>%
+  spread(taxon, cover, fill = 0) %>%
[…]

### 2024-07-11 src/ordination.R

@@ -1,0 +1,248 @@
+library(tidyverse)
+
+c("forestgreen",  
+  "gold", 
+  "#40E0D0",
+  "#A020F0",
+  "#5CACEE",
+  "#551A8B") -> colores
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### NMDS plant communities
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(species) %>%
+  select(id, taxon, cover) %>%
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 2) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
+  merge(header) %>%
+  dplyr::select(id, habitat, Dim.1, Dim.2) %>%
+  mutate(habitat = fct_relevel(habitat, "N1", "N2", "N3", "N4B", "N4A", "N4C")) %>%
+  mutate(habitat = fct_recode(habitat, "Forests" = "N1", "Meadows" = "N2",
+                              "Parks" = "N3", "Roadsides" = "N4B",
+                              "Residential" = "N4A", "Industrial" = "N4C")) -> sitescores
+
+### Species scores
+
+vegan::scores(nmds, "species") %>%
+  data.frame() %>%
+  rownames_to_column("Taxon") %>%
+  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
+  mutate(Species = Taxon) %>%
+  separate(Species, into = c("G", "S")) %>%
+  mutate(Species = paste(substr(G, 1, 3), substr(S, 1, 3), sep = "")) %>%
[…]

### 2024-07-11 src/soils.R

@@ -1,0 +1,114 @@
+library(tidyverse)
+
+c("forestgreen",  
+  "gold", 
+  "#40E0D0",
+  "#A020F0",
+  "#5CACEE",
+  "#551A8B") -> colores
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### PCA soils
+
+header %>% select(id, habitat) %>% merge(soils) -> datos
+datos %>% select(-c(id, habitat)) %>% FactoMineR::PCA(graph = FALSE) -> pca1
+cbind((datos %>%  dplyr::select(id, habitat)), data.frame(pca1$ind$coord[, 1:2])) %>%
+  mutate(habitat = fct_relevel(habitat, "N1", "N2", "N3", "N4B", "N4A", "N4C")) %>%
+  mutate(habitat = fct_recode(habitat, "Forests" = "N1", "Meadows" = "N2",
+                              "Parks" = "N3", "Roadsides" = "N4B",
+                              "Residential" = "N4A", "Industrial" = "N4C")) -> pcaInds
+pca1$var$coord[, 1:2] %>% data.frame %>% rownames_to_column(var = "Variable") %>%
+  mutate(Dim.1 = ifelse(Variable == "Pb", Dim.1 + .05, Dim.1)) -> pcaVars
+
+ggplot(pcaInds, aes(x = Dim.1, y = Dim.2)) +
+  coord_fixed() +
+  geom_hline(yintercept = 0, linetype = "dashed") +
+  geom_vline(xintercept = 0, linetype = "dashed") +
+  geom_segment(data = pcaVars, aes(x = 0, y = 0, xend = 6*Dim.1, yend = 6*Dim.2)) +
+  geom_point(aes(fill = habitat), color = "black", size = 5, shape = 23) +
+  geom_label(data = pcaVars, aes(x = 6*Dim.1, y = 6*Dim.2, label = Variable),  show.legend = FALSE, size = 3) +
+  ggthemes::theme_tufte() + 
+  theme(text = element_text(family = "sans"),
+        strip.background = element_blank(),
+        legend.position = "bottom", 
+        #legend.direction = "vertical",
+        legend.title = element_blank(),
+        legend.spacing.x = unit(0, "mm"),
+        legend.spacing.y = unit(0, "mm"),
+        legend.text = element_text(size = 10), 
+        panel.background = element_rect(color = "black", fill = NULL),
[…]

### 2024-07-11 src/species-pools.R

@@ -1,0 +1,146 @@
+library(tidyverse); library(ggvenn)
+
+read.csv("data/header.csv", fileEncoding = "latin1") %>%
+  filter(habitat != "N4D") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+
+### Plant species pools
+
+list(
+  Forests = 
+    header %>% 
+    filter(habitat == "N1") %>% 
+    select(id) %>% 
+    merge(species) %>% 
+    pull(taxon) %>% 
+    unique, 
+  Meadows = 
+    header %>% 
+    filter(habitat == "N2") %>% 
+    select(id) %>% 
+    merge(species) %>% 
+    pull(taxon) %>% 
+    unique, 
+  `Urban habitats` = 
+    header %>% 
+    filter(habitat %in% c("N3", "N4A", "N4B", "N4C")) %>% 
+    select(id) %>% 
+    merge(species) %>% 
+    pull(taxon) %>% 
+    unique) -> df1
+
+### Fungal species pools
+
+list(
+  Forests = 
+    header %>% 
+    filter(habitat == "N1") %>% 
+    select(id) %>% 
+    merge(fungi) %>% 
+    pull(taxon) %>% 
+    unique, 
+  Meadows = 
+    header %>% 
+    filter(habitat == "N2") %>% 
+    select(id) %>% 
+    merge(fungi) %>% 
+    pull(taxon) %>% 
+    unique, 
+  `Urban habitats` = 
+    header %>% 
+    filter(habitat %in% c("N3", "N4A", "N4B", "N4C")) %>% 
+    select(id) %>% 
+    merge(fungi) %>% 
+    pull(taxon) %>% 
+    unique) -> df2
+
+### Bacterial species pools
+
+list(
+  Forests = 
+    header %>% 
+    filter(habitat == "N1") %>% 
+    select(id) %>% 
+    merge(bacteria) %>% 
+    pull(taxon) %>% 
+    unique, 
+  Meadows = 
+    header %>% 
+    filter(habitat == "N2") %>% 
+    select(id) %>% 
+    merge(bacteria) %>% 
[…]

### 2024-07-11 src/target spp.R

@@ -1,0 +1,34 @@
+library(tidyverse)
+
+read.csv("data/species.csv") %>%
+  group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # add header data
+  releves
+
+releves %>%
+  group_by(id) %>%
+  arrange(-cover, taxon, .by_group = TRUE) %>%
+  mutate(totalcover = sum(cover)) %>%
+  mutate(relcover = cover/totalcover) %>%
+  mutate(cumcover = cumsum(relcover)) -> relcovers
+
+read.csv("../#data/tpl/results/TPLNames.csv", fileEncoding = "latin1") %>%
+  select(Taxon, Family) %>%
+  rename(taxon = Taxon) -> families
+
+read.csv("../#data/baskin/results/dormancy.csv", fileEncoding = "latin1") %>%
+  select(TPLName, Dormancy) %>%
+  rename(taxon = TPLName) -> dormancy
+
+
+relcovers %>% filter(cumcover > 0.80) %>%
+  filter(cumcover == min(cumcover)) %>% select(id, cumcover) %>%
+  rename(threshold = cumcover) %>%
+  merge(relcovers) %>%
+  filter(cumcover <= threshold) %>%
+  select(id, taxon, cover) %>%
+  group_by(taxon) %>%
+  summarise(F = length(taxon), A = mean(cover)) %>%
+  merge(families,all.x = TRUE) %>% 
+  merge(dormancy, all.x = TRUE) %>% 
+  write.csv("results/targetspp.csv", row.names = F)

### 2024-07-11 src/taxo-vs-compo-novelty.R

@@ -1,0 +1,342 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") %>%
+  filter(habitat != "N4D") -> header
+read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
+  summarise(cover = sum(cover_percent)) -> # merge layers
+  species
+read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+
+### Plant novelty versus N1
+
+header %>%
+  filter(habitat %in% c("N1")) %>%
+  merge(species) %>%
+  pull(taxon) %>%
+  unique -> spp # Non-urban species
+
+species %>%
+  mutate(Novelspp = ifelse(! taxon %in% spp, "Yes", "No")) %>%
+  group_by(id, Novelspp) %>%
+  tally() %>%
+  spread(Novelspp, n, fill = 0) %>%
+  mutate(TPN1 = Yes / (No + Yes)) %>%
+  select(id, TPN1) -> tpn1 # taxonomic plant novelty
+  
+### Fungal novelty versus N1
+
+header %>%
+  filter(habitat %in% c("N1")) %>%
+  merge(fungi) %>%
+  pull(taxon) %>%
+  unique -> spp # Non-urban species
+
+fungi %>%
+  mutate(Novelspp = ifelse(! taxon %in% spp, "Yes", "No")) %>%
+  group_by(id, Novelspp) %>%
+  tally() %>%
+  spread(Novelspp, n, fill = 0) %>%
+  mutate(TFN1 = Yes / (No + Yes)) %>%
+  select(id, TFN1) -> tfn1 # taxonomic fungal novelty
+
+### Bacterial novelty versus N1
+
+header %>%
+  filter(habitat %in% c("N1")) %>%
+  merge(bacteria) %>%
+  pull(taxon) %>%
+  unique -> spp # Non-urban species
+
+bacteria %>%
+  mutate(Novelspp = ifelse(! taxon %in% spp, "Yes", "No")) %>%
+  group_by(id, Novelspp) %>%
+  tally() %>%
+  spread(Novelspp, n, fill = 0) %>%
+  mutate(TBN1 = Yes / (No + Yes)) %>%
+  select(id, TBN1) -> tbn1 # taxonomic bacterial novelty
+
+### Plant novelty versus N2
+
+header %>%
[…]

### 2024-07-11 src/temperatures.R

@@ -1,0 +1,51 @@
+library(tidyverse); library(lubridate)
+
+c("#FFA500",  
+  "yellow", 
+  "#B3EE3A", 
+  "#40E0D0",
+  "#5CACEE", 
+  "#A020F0",
+  "#551A8B") -> colores
+
+## Function reads and formats GP5WSHELL csv export files
+
+read_gp5 <- function(x)
+{
+  require(tidyverse)
+  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
+    dplyr::rename(temperature = X.1.oC) %>%
+    mutate(time = as.POSIXct(Time, 
+                             format =" %d.%m.%Y %H:%M:%S", 
+                             tz = "UTC"),
+           logger = gsub("\\_.*", "" , x)) %>%
+    dplyr::select(logger, time, temperature)
+}
+
+## Read logger header data
+
+read.table("data/temperature/Loggers.txt", sep = "\t", header = TRUE, fileEncoding = "latin1") %>%
+  mutate(Installed = as.POSIXct(Installed, 
+                                format ="%d/%m/%Y", 
+                                tz = "UTC")) %>%
+  rename(habitat = N.type) %>%
+  rename(logger = Logger) -> header
+
+## Read and bind all loggers, merge with logger header data
+
+wd <- getwd()
+setwd(paste(wd, "/data/temperature", sep = ""))
+do.call(rbind, lapply(list.files(pattern = "*.csv"), read_gp5)) %>% 
+  merge(header, by = "logger") %>%
+  na.omit %>% 
+  filter(time > (Installed + 60*60*24*7)) %>% # Remove records up to one week after installation
+  select(ID, plot, habitat, time, temperature) ->
+  logs
+setwd(wd)
+
+logs %>%
+  ggplot(aes(time, temperature, color = habitat)) + 
+  facet_wrap(habitat ~ plot) +
+  geom_line() +
+  scale_color_manual(values = colores) + 
+  coord_cartesian(ylim = c(0, 30))

### 2024-07-11 src/Pbtolerance.R
cleaning

@@ -1,36 +1,0 @@
-library(tidyverse)
-
-c("#FFA500",  
-  "yellow", 
-  "#B3EE3A", 
-  "#40E0D0",
-  "#5CACEE", 
-  "#A020F0",
-  "#551A8B") -> colores
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/traits.csv", fileEncoding = "latin1") -> traits
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-merge(soils, species) %>%
-  filter(Pb > 150) %>%
-  merge(traits) %>%
-  filter(origin == "Native") %>%
-  group_by(taxon) %>%
-  summarise(tolerance = weighted.mean(Pb, cover)) %>%
-  arrange(-tolerance) %>%
-  data.frame %>%
-  pull(taxon) -> tolerantspp
-
-species %>%
-  filter(taxon %in% tolerantspp) %>%
-  group_by(taxon) %>%
-  summarise(n = length(taxon), cover = mean(cover)) %>%
-  filter(n > 5) %>%
-  arrange(-cover) %>%
-  data.frame

### 2024-07-11 src/bacteria.R

@@ -1,6 +1,6 @@
 library(tidyverse)
 
-read.csv("data/bacteria/Table3/table_after_filtering.csv", sep = "\t", 
+read.csv("data/bacteria-original.csv", sep = "\t", 
          #fileEncoding = "latin1", 
          skip = 1) %>%
   separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%

### 2024-07-11 src/compositional-novelty.R

@@ -1,210 +1,0 @@
-library(tidyverse)
-
-read.csv("data/header.csv", fileEncoding = "latin1") %>%
-  filter(habitat != "N4D") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-
-### Compositional novelty for plants
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(species) %>%
-  select(id, taxon, cover) %>%
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 4) -> 
-  nmds # Ordination output
-
-### Site scores
-
-vegan::scores(nmds, "sites") %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  gather(Trait, Value, - id) %>%
-  merge(select(header, id, habitat)) -> sitescores
-
-### Novelty (versus N1 and N2 together)
-
-sitescores %>%
-  filter(habitat %in% c("N1", "N2")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  gather(Reference, Centroid, N1:N2) %>%
-  arrange(id, Trait, Reference) %>%
-  mutate(Novelty = Centroid - Value) %>%
-  group_by(id, Trait) %>%
-  filter(Novelty == min(Novelty)) %>%
-  group_by(id, habitat) %>%
-  summarise(Novelty = sum(Novelty)) %>%
-  group_by(habitat) %>%
-  summarise(Novelty = mean(Novelty)) 
-
-### Novelty (versus N1)
-
-sitescores %>%
-  filter(habitat %in% c("N1")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
[…]

### 2024-07-11 src/dca.R

@@ -1,245 +1,0 @@
-library(tidyverse); library(vegan)
-
-c("#FFA500",  
-  "yellow", 
-  "#B3EE3A", 
-  "#40E0D0",
-  "#5CACEE", 
-  "#A020F0",
-  "#551A8B") -> colores
-
-read.csv("data/header.csv", fileEncoding = "latin1") %>%
-  filter(habitat != "N4D") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-### PCA soils
-
-header %>% select(id, habitat) %>% merge(soils) -> datos
-datos %>% select(-c(id, habitat)) %>% FactoMineR::PCA(graph = FALSE) -> pca1
-cbind((datos %>%  dplyr::select(id, habitat)), data.frame(pca1$ind$coord[, 1:2])) -> pcaInds
-pca1$var$coord[, 1:2] %>% data.frame %>% rownames_to_column(var = "Variable") -> pcaVars
-
-ggplot(pcaInds, aes(x = Dim.1, y = Dim.2)) +
-  coord_fixed() +
-  geom_hline(yintercept = 0, linetype = "dashed") +
-  geom_vline(xintercept = 0, linetype = "dashed") +
-  geom_segment(data = pcaVars, aes(x = 0, y = 0, xend = 3*Dim.1, yend = 3*Dim.2)) +
-  geom_point(aes(fill = habitat), color = "black", size = 7, shape = 21) +
-  ggrepel::geom_label_repel(data = pcaVars, aes(x = 3*Dim.1, y = 3*Dim.2, label = Variable),  show.legend = FALSE, size = 2) +
-  ggthemes::theme_tufte() + 
-  ggtitle(label = "PCA of soil properties") +
-  theme(text = element_text(family = "sans"),
-        legend.position = "right", 
-        legend.title = element_blank(),
-        legend.text = element_text(size = 12, color = "black"),
-        panel.background = element_rect(color = "black", fill = NULL),
-        axis.title = element_text(size = 12),
-        axis.text = element_text(size = 12, color = "black"),
[…]

### 2024-07-11 src/diversity.R

@@ -1,145 +1,0 @@
-library(tidyverse)
-
-c("#FFA500",  
-  "yellow", 
-  "#B3EE3A", 
-  "#40E0D0",
-  "#5CACEE", 
-  "#A020F0",
-  "#551A8B") -> colores
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/traits.csv", fileEncoding = "latin1") -> traits
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-# Aliens
-
-merge(species, traits) %>%
-  group_by(id, origin) %>%
-  summarise(c = sum(cover)) %>%
-  spread(origin, c, fill = 0) %>%
-  mutate(aliencover = 100 * (Alien / (Alien + Native))) %>%
-  merge(header) %>%
-  group_by(habitat) %>%
-  summarise(m = mean(aliencover))
-
-merge(species, traits) %>%
-  group_by(id, origin) %>%
-  summarise(n = length(taxon)) %>%
-  spread(origin, n, fill = 0) %>%
-  mutate(alienspp = 100 * (Alien / (Alien + Native))) %>%
-  merge(header) %>%
-  group_by(habitat) %>%
-  summarise(nalien = mean(alienspp))
-
-merge(species, traits) %>%
-  merge(header) %>%
-  select(habitat, taxon, origin) %>%
-  unique %>%
-  group_by(habitat, origin) %>%
-  tally %>%
-  spread(origin, n, fill = 0) %>%
-  mutate(aliens = 100 * (Alien / (Alien + Native))) -> alienspp
-
-# Alpha
-
-species %>% group_by(id) %>% summarise(plantR = length(taxon)) %>% data.frame -> r1
-fungi %>% group_by(id) %>% summarise(fungiR = length(taxon)) %>% data.frame-> r2
-bacteria %>% group_by(id) %>% summarise(bacteriaR = length(taxon)) %>% data.frame -> r3
-
-# Gamma
-
[…]

### 2024-07-11 src/fungi.R

@@ -1,6 +1,6 @@
 library(tidyverse)
 
-read.csv("data/fungi/Table3/table_after_filtering.csv", sep = "\t", 
+read.csv("data/fungi-original.csv", sep = "\t", 
          #fileEncoding = "latin1", 
          skip = 1) %>%
   separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%

### 2024-07-11 src/merge-novelty.R

@@ -1,29 +1,0 @@
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-header %>%
-  select(id, habitat) %>%
-  merge(tpn) %>%
-  merge(tfn) %>%
-  merge(tbn)  %>%
-  merge(cpn1) %>%
-  merge(cpn2) %>%
-  merge(cfn1) %>%
-  merge(cfn2) %>%
-  merge(cbn1) %>%
-  merge(cbn2) %>%
-  merge(soils) %>%
-  select(-c(habitat, id)) %>%
-  FactoMineR::PCA()
-  
-
-header %>%
-  select(id, habitat) %>%
-  merge(tpn) %>%
-  merge(tfn) %>%
-  merge(tbn)  %>%
-  merge(cpn1) %>%
-  merge(cfn1) %>%
-  merge(cbn1) %>%
-  merge(soils) %>%
-  select(-c(habitat, id)) %>%
-  FactoMineR::PCA()

### 2024-07-11 src/target spp.R

@@ -1,34 +1,0 @@
-library(tidyverse)
-
-read.csv("data/species.csv") %>%
-  group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # add header data
-  releves
-
-releves %>%
-  group_by(id) %>%
-  arrange(-cover, taxon, .by_group = TRUE) %>%
-  mutate(totalcover = sum(cover)) %>%
-  mutate(relcover = cover/totalcover) %>%
-  mutate(cumcover = cumsum(relcover)) -> relcovers
-
-read.csv("../#data/tpl/results/TPLNames.csv", fileEncoding = "latin1") %>%
-  select(Taxon, Family) %>%
-  rename(taxon = Taxon) -> families
-
-read.csv("../#data/baskin/results/dormancy.csv", fileEncoding = "latin1") %>%
-  select(TPLName, Dormancy) %>%
-  rename(taxon = TPLName) -> dormancy
-
-
-relcovers %>% filter(cumcover > 0.80) %>%
-  filter(cumcover == min(cumcover)) %>% select(id, cumcover) %>%
-  rename(threshold = cumcover) %>%
-  merge(relcovers) %>%
-  filter(cumcover <= threshold) %>%
-  select(id, taxon, cover) %>%
-  group_by(taxon) %>%
-  summarise(F = length(taxon), A = mean(cover)) %>%
-  merge(families,all.x = TRUE) %>% 
-  merge(dormancy, all.x = TRUE) %>% 
-  write.csv("results/targetspp.csv", row.names = F)

### 2024-07-11 src/taxo-vs-compo-novelty.R

@@ -1,342 +1,0 @@
-library(tidyverse)
-
-read.csv("data/header.csv", fileEncoding = "latin1") %>%
-  filter(habitat != "N4D") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-
-### Plant novelty versus N1
-
-header %>%
-  filter(habitat %in% c("N1")) %>%
-  merge(species) %>%
-  pull(taxon) %>%
-  unique -> spp # Non-urban species
-
-species %>%
-  mutate(Novelspp = ifelse(! taxon %in% spp, "Yes", "No")) %>%
-  group_by(id, Novelspp) %>%
-  tally() %>%
-  spread(Novelspp, n, fill = 0) %>%
-  mutate(TPN1 = Yes / (No + Yes)) %>%
-  select(id, TPN1) -> tpn1 # taxonomic plant novelty
-  
-### Fungal novelty versus N1
-
-header %>%
-  filter(habitat %in% c("N1")) %>%
-  merge(fungi) %>%
-  pull(taxon) %>%
-  unique -> spp # Non-urban species
-
-fungi %>%
-  mutate(Novelspp = ifelse(! taxon %in% spp, "Yes", "No")) %>%
-  group_by(id, Novelspp) %>%
-  tally() %>%
-  spread(Novelspp, n, fill = 0) %>%
-  mutate(TFN1 = Yes / (No + Yes)) %>%
-  select(id, TFN1) -> tfn1 # taxonomic fungal novelty
-
-### Bacterial novelty versus N1
-
-header %>%
-  filter(habitat %in% c("N1")) %>%
-  merge(bacteria) %>%
-  pull(taxon) %>%
-  unique -> spp # Non-urban species
-
-bacteria %>%
-  mutate(Novelspp = ifelse(! taxon %in% spp, "Yes", "No")) %>%
-  group_by(id, Novelspp) %>%
-  tally() %>%
-  spread(Novelspp, n, fill = 0) %>%
-  mutate(TBN1 = Yes / (No + Yes)) %>%
-  select(id, TBN1) -> tbn1 # taxonomic bacterial novelty
-
-### Plant novelty versus N2
-
-header %>%
[…]

### 2024-07-15 doc/manuscript.Rmd
first draft of methods and results

@@
 In post-industrial landscapes affected by the phenomenon of urban contraction, land located in the heart of cities remains unused for long periods (Haase et al., 2014) .
-When the use of any urban plot ceases, it is colonized by plants from the regional flora capable of living in urban environments (Johnson et al., 2017) .
-These urban plants are adapted to disturbances and can grow in degraded soils; some even accumulate heavy metals, making them nature-based solutions for bioremediation of degraded lands (Song et al., 2019) .
-Species traits, along with the legacy of human impact, determine which regional flora species colonize which plots (Swan et al., 2021) .
-When the human legacy is intense, the plant communities that develop will be novel communities, different from the historical reference communities that existed before human intervention (Gedan and Fernández-Pascual, 2019) .
-The novelty of the communities can be reflected in their floristic, compositional and phylogenetic composition, including their proportion of exotic species (Lososová et al., 2012) .
+When the use of any urban plot ceases, it is colonized by plants, fungi or bacteria from the regional species pools that capable of living in urban environments (Johnson et al., 2017) .
+This urban biota is adapted to disturbance and can grow in degraded soils; this should make these species useful for nature-based solutions such as bioremediation of degraded lands (Song et al., 2019) .
+Species traits, along with the legacy of human impact, determine which regional pool species colonize which plots (Swan et al., 2021) .
+When the human legacy is intense, the communities that develop will be novel communities, different from the historical reference communities that existed before human intervention (Gedan and Fernández-Pascual, 2019) .
+The novelty of the communities can be reflected in their composition, including their elements such as the proportion of exotic species or the proportion of obligate synanthropic species (Lososová et al., 2012) .
 In any case, proper management of urban communities can turn them into resources with high ecological and social potential (Anderson and Minor, 2017).
-(Kowarik, 2018) : in the Ibero-Atlantic ecoregion, anthropogenic habitats can host up to a third of the ecoregional flora.
-
-The ability of plants to colonize any habitat, including urban ones, depends largely on their seed regeneration niche, that is, on their dispersal, persistence, germination and establishment functions (Grubb, 1977, Saatkamp et al., 2019) .
-However, although it is known that urban species often share a series of functional and ecological traits (Aronson et al., 2007, Lososová et al., 2006) , little is known about their seed regeneration niche.
-Especially, information on its germination traits is lacking.
-In species living in non-anthropogenic environments, germination phenology is regulated by environmental factors (water, temperature, light) to ensure that germination occurs at the right time to maximize subsequent seedling survival (Carta et al., 2022a) .
-Compared to the large amount of data on the germination of species from non-anthropogenic habitats, little is known about what factors regulate the germination of urban species.
-Considering other habitats affected by humans (Fernández-Pascual et al., 2022) , it could be expected that urban seeds have their germination unrelated to macroclimatic signals and show tolerance to stress due to heat, drought or heavy metals.
-However, this hypothesis needs to be tested through a synthesis of the seed regeneration niche in urban environments.
-
-Cities 
-1. highly fragmented landscapes
-2. mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use)
-3. subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems
-4. comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure
-5.
-contains a multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix
-
+(Kowarik, 2018).
 Cities are comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure.
 They are mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use) which are subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems (Kowarik, 2005).
-
 The “Four Nature approach” narrows down the variety of transformational stages in urban settings to four major types that can supply wilderness in urban region (Kowarik 1992, 2005).
-Each of the four types relates to nature in general but results from different trajectories in human-nature interactions:
-
-Spontaneous vegetation appears in different habitat types in the urban matrix,.
-Multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix.
-The spontaneous herbs could provide the very interesting pool of species for ecological restoration of the urban area.
-Their ecological value has been clear since many years because their contribution to the ecosystem services in urban areas.
-Although their aesthetic value has been highlighted repeatedly, their acceptance by the residents in urban areas has been questioned.
-Many residents associate the presence of spontaneous plants with the neglect in maintenance by public (municipal) management service.
-Spontaneous native urban vegetation has commonly been described as demonstrating resilience (Ignatieva et al., 2000) and exhibiting adaptations to human disturbance (Lundholm and Marlin, 2006; Sukopp, 2004).
-
-Urban species pool is comprised of native species present before the urban development, native species developed because the new urban ecosystem and introduced species by human activity.
-However, just some of them can adapt to urban environment.
-
-IGS (informal green space) consists of any urban space with a history of strong anthropogenic disturbance that is covered at least partly with non-remnant, spontaneous vegetation (Del Tredici, 2010).
 
@@
 
+We performed all analyses using R version 4.3.1 [@RN5387] and the package *tidyverse* [@RN5387] for data manipulation and visualization.
+To describe soil temperatures as recorded by the dataloggers, we calculated the mean annual temperature, the mean diurnal range and the temperature annual range following the same methdology as the one applied to calculate bioclimatic indices bio1, bio2 and bio7 in WorldClim [@RN5064].
+To assess the variability in soil physicochemical parameters, we conducted a Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166], and a correlation analysis to detect auto-correlated soil variables.
+We used PERMANOVA as implemented in *vegan* [@RN5386] to test differences between habitat types in their soil variables and their composition of plant species, fungi ASVs and bacteria ASVs.
+We used Non-metric Multi-Dimensional Scaling (NMDS) ordination as implemented in *vegan* [@RN5386] to assess variation in taxonomic composition of plants (abundance values of species), fungi (presence/absence of ASVs) and bacteria (presence/absence of ASVs).
+We used environmental fitting as implemented in *vegan* [@RN5386] to check the correlation between the community ordination axes and the soil physicochemical properties.
+We used the NMDS multidimensional compositional distance from each urban plot (i.e.
+park, roadside, residential or industrial) to the centroid of the two reference non-urban habitats (i.e.
+forests and meadows) to calculate for each plot, taxonomic group and reference habitat a value of community novelty.
+We fitted a Generalized Mixed Model using the package *nlme* [@RN2333] to test the effect on comminity novelty of taxonomic group, habitat of reference (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables.
+The models inclused sampling plot as a random factor.
+
 # Results
 
-## Soil temperatures and properties across habitats
+## Soil properties across habitats
 
-## Plant, fungal and bacterial diversity across habitats
+Soil temperature records (**Figure 2A**) indicated that forest soils were relatively buffered in their range of variation, with a mean annual temperature of 11.4 ºC, a mean diurnal range of 1.72 ºC and an annual range of 4.74 ºC.
+The meadow and urban plots had more thermal oscillation, with a mean diurnal range above 4 ºC and an annual range above 10 ºC.
+PCA ordination of the soil physicochemical properties (**Appendix S1**) indicated that the first axis of variation, explaining 23% of the variance, ordered the soils according to their concentration of heavy metals.
+The variables with the highest contribution to this axis (> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr, which accounted for 66% of the variability of the first axis.
+This first axis did not separate urban from non-urban habitats: for example, vacant industrial lots had some of the highest concentration in heavy metals, but also some of the lowest.
+Likewise, forest plots were distributed across the first axis.
+The second axis of variation explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (> 10%) being organic matter content, bulk density, N concentration and pH, which made up 59% of the axis' variability.
+This axis did separate soils according to their habitat: for example, forest sites had high organic matter and N content, while industrial vacant lots had higher bulk density and pH.
+The third axis explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (>10%) being Mg content, Mg/K ratio and conductivity, together explaining 50% of the axis.
+In this case, like for the first axis, there was no clear separation between habitats, with industrial sites having some of the highest and the lowest values.
+Correlation analysis showed strong correlations between the groups of variables contributing to the first, second and third axes (**Appendix S1**).
+Therefore, to avoid autocorrelation issues, for further analyses we kept the variables Pb content, organic matter content and Mg content (**Figure 2B**), which together provided a straightforward biological interpretation of patterns in soil composition.
+Of these values, as previously indicated by the PCA, organic matter was the only one with a clear pattern of variation: a decrease from the non-urban to the urban habitats.
+A PERMANOVA of the soil values indicated that there was not a significant difference between habitats in their soil physicochemical properties (100,000 permutations, R^2^ = 0.19, F = 1.15, p = 0.31).
 
-> PERMANOVAs
+```{r fig2, out.width = "450px", echo = FALSE, fig.cap = "Boxplot of soil properties accross habitats.
+(A) Temperatures recorded every hour in one sampling plot per habitat from 1 Jan 2024 to 31 Dec 2024.
+(B) Soil physicochemical properties.
+Values are shown for the three most explanatory and uncorrelated variables, as per a previous PCA analysis: soil Pb content, soil organic matter content, and soil Mg content."}
+knitr::include_graphics("../results/figures/Fig2.png")
+```
 
-> Analysis with species, genera and ASVs
+> Poner unidades de las variables en figura
 
-> Analysis with presence absence
+## Plant, fungal and bacterial communities across habitats
+
+The survey of plant diversity yielded 216 species, with the average number per plot being 26 (min = 10, max = 38).
+Of these species, 48% were only found in urban-habitats (parks, roadsides, residential lots and industrial lots), 27% occurred only in non-urban habitats (forests and meadows) and 25% were shared by both urban and non-urban habitats (**Fig3A**).
+On average, the habitat with the highest number of species per plot were the industrial lots (29), followed by residential lots (27), roadsides (24), meadows (23), parks (20) and forests (18).
+The most frequent species were *Holcus lanatus* (22 plots), *Plantago lanceolata* (20), *Helminthotheca echioides* (16), *Dactylis glomerata* (15) and *Lotus corniculatus* (14).
+PERMANOVA indicated that the plant species composition was significantly different accross habitats (100,000 permutations, R^2^ = 0.42, F = 3.48, p = < 0.001).
+NMDS ordination of the plant species composition (**Fig.
+3D**) indicated that the plots belonging to the same habitat were relatively grouped together, with the main exception being one of the roadside plots.
+The ordination indicated that the major gradient in vegetation composition along the first axis was a separation between the forest plots and the rest, with the meadows being the closest to the forests.
+Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that none of the soils factors had a significant correlation (R^2^ < 0.14, p > 0.05) with the composition of the plant communities.
+
+The DNA metabarcoding of soil fungal communities resulted in 4,689 ASVs.
+Of these ASVs, 46% were only found in urban-habitats (parks, roadsides, residential lots and industrial lots), 36% occurred only in non-urban habitats (forests and meadows) and 18% were shared by both urban and non-urban habitats (**Fig3B**).
+The average number of ASVs per plot was 340 (min = 174, max = 587), with the habitat-averages ranging from residential vacant lots (415) to forests (393), meadows (336), parks (323), industrial vacant lots (300) and roadsides (276).
+The majority of ASVs were assigned to Ascomycota (56%), Basidiomycota (24%), Rozellomycota (10%) and Glomeromycota (4%).
+Of the different habitats, forests had the highest proportion of Basidiomycota (25% vs.
+< 17% in the other habitats) and the lowest proportion of Ascomycota (57% vs.
+> 64%).
+The proportion of Rozellomycota was similar across habitats (from 6% in meadows to 11% in parks).
+The proportion of Glomeromycota varied between 5% in meadows to 0.5% in forests.
+A PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.33, F = 2.39, p = < 0.001).
+NMDS ordination of the fungal ASV composition (**Fig.
+3E**) behavead similarly to the ordination of plant communities, with plots tending to group by habitat, and the major compositional gradient reflecting the separation between the forest plots and the rest.
+Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated only organic matter content had a significant correlation (R^2^ = 0.46,  p < 0.001) with the composition of the fungal communities.
+Low organic matter content correlated with the communities from industrial vacant lots and roadsides (**Fig.
+3E**).
+
+For the bacterial communities, DNA metabarcoding yielded 16,306 ASVs.
+Of these ASVs, 44% were only found in urban-habitats (parks, roadsides, residential lots and industrial lots), 31% occurred only in non-urban habitats (forests and meadows) and 25% were shared by both urban and non-urban habitats (**Fig3C**).
+The average number of ASVs per plot was 1769 (min = 1478, max = 2074), with the habitat-averages ranging from roadsides (1848) to industrial lots (1794), residential lots (1792), meadows (1775), parks (1706) and forests (1697).
+The majority of ASVs were assigned to Proteobacteria (22%), Planctomycetota (15%), Actinobacteriota (15%) and Acidobacteriota (10%).
+Compared to other habitats, forests had higher proportions of Proteobacteria and Proteobacteria; a lower proportion of Actinobacteriota; and a similar proportion of Planctomycetota.
+A PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.32, F = 2.51, p = < 0.001).
+NMDS ordination of the bacterial ASV composition (**Fig.
+3F**) resulted in less grouping by habitat than for the plants and the fungi, but still reflected a main composotional gradient from forests to industrial vacant plots.
+Like in the case of the fungal communities, environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated only organic matter content had a significant correlation (R^2^ = 0.39,  p = 0.001) with the composition of the fungal communities.
+Low organic matter content correlated with the communities from industrial vacant lots and roadsides (**Fig.
+3F**).
+
+```{r fig3, out.width = "450px", echo = FALSE, fig.cap = "Plant, fungal and bacterial diversity of the sampling plots.
+Top panels show Venn plots for the shared numbers of (A) plant species, (B) fungal ASVs and (c) bacterial ASVs between urban (parks, roadsides, residential lots and industrial lots) and non-urban (forests, meadows) habitats.
+Bottom panels show  biplots of community composition ordinartion produced by Non-metric Multi-Dimensional Scaling (NMDS).
+NMDs were conducted for (A) plant survey (by species abundance), (B) the soil fungal survey (by ASV presence/absence) and (C) the soil bacterial survey (by ASV presence/absence).
+Colors indicate the habitat.
+In the biplots, arrows and labels indicate soil variables with a significant correlation with the composition of the communities.
+Only soil organic matter content and fungal-bacterial communities had a significant correlation."}
+knitr::include_graphics("../results/figures/Fig3.png")
+```
 
 ## Community novelty across habitats and taxonomic groups
+
+We used the NMDS multidimensional compositional distance (calculate from the ordinations in **Fig.
+3D-F**) from each urban plot (i.e.
+park, roadside, residential or industrial) to the centroid of the two reference non-urban habitats (i.e.
+forests and meadows) to calculate for each plot, taxonomic group and reference habitat a value of community novelty (**Fig.
+4**).
+The novelty was significantly different between taxonomic groups (F = 45.88, p < 0.001), being higher for plants (2.45 ± 0.332 S.E.) than for fungi (2.15 ± 0.396 S.E.) and bacteria (1.56 ± 0.303 S.E.).
+The novelty was also significantly different depending on the reference non-urban habitat that was used (F = 199.08, p < 0.001), being higher compared to forests (2.60 ± 0.287 S.E.) than to meadows (1.51 ± 0.300 S.E.).
+Nonetheless, for each taxonomic group, there was a high positive correlation between the novelty compared against forests or meadows (r = 0.5 for plants; 0.7 for fungi; 0.9 for bacteria).
+The different urban habitats also differed significantly in their novelty (F = 9.51, p = 0.001), being higher for industrial vacant lots (2.46 ± 0.311 S.E.) than for roadsides (2.18 ± 0.367 S.E.), parks (1.95 ± 0.440 S.E.) and residential vacant lots (1.63 ± 0.307 S.E.).
+The soil physicochemical properties (Pb, organic matter, Mg) had no significant effect on community novelty (p > 0.05).
+
+```{r fig4, out.width = "450px", echo = FALSE, fig.cap = "Community novelty for plants, fungi and bacteria.
+Community novelty was calculated as the multidimensional distance from each urban plot to the centroid of two reference non-urban habitats (forests, panels A-C-E; and meadows, panels B-D-F), as calculated by NMDS ordination.
+Novelty was calculated separately for plant (panels A-B), fungal (panels C-D) and bacterial communities (panels E-F)."}
+knitr::include_graphics("../results/figures/Fig4.png")
+```
 
@@
 There are some other habitats in the city which were not included in the present study, e.g. wetlands, beach, cliffs.
+
+# Data availability
+
+The original datasets, as well as R code for analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/disturbance.
+Upon publication, a version of record of the repository will be deposited in Zenodo.
+
+# References
+
+::: {#refs}
+:::
+
+# Supplementary
+
+**Appendix S1** Principal Component Analysis and correlation analysis of the soil physicochemical properties.

### 2024-07-15 results/supplementary/AppendixS1.Rmd
first draft of methods and results

---
title: "Appendix S1 Principal Component Analysis and correlation analysis of the soil physicochemical properties."
output:
  bookdown::pdf_document2:
    reference_docx: template.docx
    number_sections: false
    toc: false
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

```{r figS1a, out.width = "450px", echo = FALSE, fig.cap = "(A) Biplot of the first and second axes of a Principal Component Analysis of the soil physicochemical properties. Circles represent the sampling plots, coloured by habitat. Labeled arrows indicate the contribution of the soil variables to the axes."}
knitr::include_graphics("figS1a.png")
```

```{r figS1b, out.width = "450px", echo = FALSE, fig.cap = "(B) Biplot of the first and third axes of a Principal Component Analysis of the soil physicochemical properties. Circles represent the sampling plots, coloured by habitat. Labeled arrows indicate the contribution of the soil variables to the axes."}
knitr::include_graphics("figS1b.png")
```

```{r figS1c, out.width = "450px", echo = FALSE, fig.cap = "(C) Correlation heatmap of the soil variables."}
knitr::include_graphics("figS1c.png")
```

### 2024-07-15 src/bacteria.R

@@ -44,34 +44,136 @@ read.csv("data/bacteria-original.csv", sep = "\t",
          Xixon29 = X29.16S,
          Xixon30 = X30.16S) -> bacteria
 
-bacteria %>% filter(is.na(kingdom))
-bacteria %>% filter(is.na(phylum))
-bacteria %>% filter(is.na(class))
-bacteria %>% filter(is.na(order))
-bacteria %>% filter(is.na(family))
-bacteria %>% filter(is.na(genus))
+### numbers
 
-bacteria %>% pull(kingdom) %>% na.omit %>% unique -> kingdoms 
-bacteria %>% pull(phylum) %>% na.omit %>% unique -> phyla 
-bacteria %>% pull(class) %>% na.omit %>% unique -> classes 
-bacteria %>% pull(order) %>% na.omit %>% unique -> orders 
-bacteria %>% pull(family) %>% na.omit %>% unique -> families 
-bacteria %>% pull(genus) %>% na.omit %>% unique -> genera  
-bacteria %>% select(species) %>% filter(species != "unidentified") %>% arrange(species) %>% na.omit %>% unique -> spp  
+bacteria %>% group_by(ASV)
 
-bacteria %>%
-  select(kingdom:species) %>%
+bacteria %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV)
+bacteria %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV) %>% pull(species) %>% unique %>% length
+
+bacteria %>% filter(species == "unidentified" | is.na(species)) %>% 
+  filter(genus != "unidentified" & !is.na(genus)) %>% group_by(ASV)
+bacteria %>% filter(species == "unidentified" | is.na(species)) %>% 
+  filter(genus != "unidentified" & !is.na(genus)) %>% group_by(ASV) %>%
+  pull(genus) %>% unique %>% length
+
+bacteria %>% 
+  filter(species != "unidentified" & !is.na(species)) %>%
+  select(species, Xixon01:Xixon09) %>%
+  gather(id, Value, -species) %>%
+  filter(Value != 0) %>%
+  select(species, id) %>%
   unique %>%
-  arrange(kingdom, phylum, class, order, family, genus, species) %>%
[…]

### 2024-07-15 src/fig2-soils.R

@@ -1,0 +1,321 @@
+library(tidyverse); library(lubridate)
+Sys.setlocale("LC_TIME", "English")
+
+c("forestgreen",  
+  "gold", 
+  "#40E0D0",
+  "#A020F0",
+  "#5CACEE",
+  "#551A8B") -> colores
+
+### Reads and format GP5WSHELL csv export files
+
+read_gp5 <- function(x)
+{
+  require(tidyverse)
+  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
+    dplyr::rename(temperature = X.1.oC) %>%
+    mutate(time = as.POSIXct(Time, 
+                             format =" %d.%m.%Y %H:%M:%S", 
+                             tz = "UTC"),
+           logger = gsub("\\_.*", "" , x)) %>%
+    dplyr::select(logger, time, temperature)
+}
+
+read.table("data/temperature/Loggers.txt", sep = "\t", header = TRUE, fileEncoding = "latin1") %>%
+  mutate(Installed = as.POSIXct(Installed, 
+                                format ="%d/%m/%Y", 
+                                tz = "UTC")) %>%
+  rename(habitat = N.type) %>%
+  rename(logger = Logger) %>%
+  filter(plot %in% c("Xixon_24", # Bosque mixto en Porceyu con sotobosque de Laurus
+                     "Xixon_07", # Detrás del cementeriu de Xove
+                     "Xixon_25", # Parque del Lauredal
+                     "Xixon_08", # Parcela en La Calzada cerca de los feriantes
+                     "Xixon_15", # Vías del tren en La Calzada
+                     "Xixon_21")) -> # Estrella de Qatar
+  loggers
+
+## Read and bind all loggers, merge with logger header data
+
+wd <- getwd()
+setwd(paste(wd, "/data/temperature", sep = ""))
+do.call(rbind, lapply(list.files(pattern = "*.csv"), read_gp5)) %>% 
+  merge(loggers, by = "logger") %>%
+  na.omit %>% 
+  filter(time > (Installed + 60*60*24*7)) %>% # Remove records up to one week after installation
+  select(ID, plot, habitat, time, temperature) ->
+  logs
+setwd(wd)
+
+logs %>%
+  mutate(habitat = fct_relevel(habitat, "N1", "N2", "N3", "N4B", "N4A", "N4C")) %>%
+  mutate(habitat = fct_recode(habitat, "Forests" = "N1", "Meadows" = "N2",
[…]

### 2024-07-15 src/fig3-diversity.R

@@ -1,0 +1,491 @@
+library(tidyverse); library(ggvenn)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### Plant species pools
+
+list(
+  Forests = 
+    header %>% 
+    filter(habitat == "forest") %>% 
+    select(id) %>% 
+    merge(plants) %>% 
+    pull(taxon) %>% 
+    unique, 
+  Meadows = 
+    header %>% 
+    filter(habitat == "meadow") %>% 
+    select(id) %>% 
+    merge(plants) %>% 
+    pull(taxon) %>% 
+    unique, 
+  `Urban habitats` = 
+    header %>% 
+    filter(habitat %in% c("park", "residential", "roadside", "industrial")) %>% 
+    select(id) %>% 
+    merge(plants) %>% 
+    pull(taxon) %>% 
+    unique) -> df1
+
+plants %>%
+  filter(taxon %in% df1$Forests & ! taxon %in% df1$Meadows & ! taxon %in% df1$`Urban habitats`) %>%
+  group_by(taxon) %>%
+  tally() %>%
+  arrange(-n)
+
+plants %>%
+  filter(! taxon %in% df1$Forests & taxon %in% df1$Meadows & ! taxon %in% df1$`Urban habitats`) %>%
+  group_by(taxon) %>%
+  tally() %>%
+  arrange(-n)
+
+plants %>%
+  filter(! taxon %in% df1$Forests & ! taxon %in% df1$Meadows & taxon %in% df1$`Urban habitats`) %>%
+  group_by(taxon) %>%
+  tally() %>%
+  arrange(-n)
+
+### Fungal species pools
+
+list(
+  Forests = 
+    header %>% 
+    filter(habitat == "forest") %>% 
+    select(id) %>% 
+    merge(fungi) %>% 
+    pull(ASV) %>% 
+    unique, 
+  Meadows = 
+    header %>% 
+    filter(habitat == "meadow") %>% 
+    select(id) %>% 
+    merge(fungi) %>% 
+    pull(ASV) %>% 
+    unique, 
+  `Urban habitats` = 
+    header %>% 
[…]

### 2024-07-15 src/fungi.R

@@ -1,5 +1,7 @@
 library(tidyverse)
 
+### Clean fungi
+
 read.csv("data/fungi-original.csv", sep = "\t", 
          #fileEncoding = "latin1", 
          skip = 1) %>%
@@ -44,46 +46,122 @@ read.csv("data/fungi-original.csv", sep = "\t",
          Xixon29 = X29.ITS,
          Xixon30 = X30.ITS) -> fungi
 
-fungi %>% filter(is.na(kingdom))
-fungi %>% filter(is.na(phylum))
-fungi %>% filter(is.na(class))
-fungi %>% filter(is.na(order))
-fungi %>% filter(is.na(family))
-fungi %>% filter(is.na(genus))
+### Fungi numbers
 
-fungi %>% pull(kingdom) %>% na.omit %>% unique -> kingdoms 
-fungi %>% pull(phylum) %>% na.omit %>% unique -> phyla 
-fungi %>% pull(class) %>% na.omit %>% unique -> classes 
-fungi %>% pull(order) %>% na.omit %>% unique -> orders 
-fungi %>% pull(family) %>% na.omit %>% unique -> families 
-fungi %>% pull(genus) %>% na.omit %>% unique -> genera  
-fungi %>% select(species) %>% filter(species != "unidentified") %>% arrange(species) %>% na.omit %>% unique -> spp  
+fungi %>% group_by(ASV)
 
-fungi %>%
-  select(kingdom:species) %>%
+fungi %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV)
+fungi %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV) %>% pull(species) %>% unique %>% length
+
+fungi %>% filter(species == "unidentified" | is.na(species)) %>% 
+  filter(genus != "unidentified" & !is.na(genus)) %>% group_by(ASV)
+fungi %>% filter(species == "unidentified" | is.na(species)) %>% 
+  filter(genus != "unidentified" & !is.na(genus)) %>% group_by(ASV) %>%
+  pull(genus) %>% unique %>% length
+
+fungi %>% 
+  filter(species != "unidentified" & !is.na(species)) %>%
+  select(species, Xixon01:Xixon09) %>%
+  gather(id, Value, -species) %>%
+  filter(Value != 0) %>%
[…]

### 2024-07-15 src/novelty.R

@@ -1,19 +1,17 @@
 library(tidyverse)
 
-read.csv("data/header.csv", fileEncoding = "latin1") %>%
-  filter(habitat != "N4D") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
 
 ### Compositional novelty for plants
 
 header %>%
   select(habitat, id) %>%
   filter(habitat != "N4D") %>%
-  merge(species) %>%
+  merge(plants) %>%
   select(id, taxon, cover) %>%
   spread(taxon, cover, fill = 0) %>%
   column_to_rownames(var = "id") -> matrix 
@@ -30,38 +28,35 @@ vegan::scores(nmds, "sites") %>%
   gather(Trait, Value, - id) %>%
   merge(select(header, id, habitat)) -> sitescores
 
-### Novelty (versus N1)
+### Novelty (versus forest)
 
 sitescores %>%
-  filter(habitat %in% c("N1")) %>%
+  filter(habitat %in% c("forest")) %>%
   group_by(habitat, Trait) %>%
   summarise(Value = mean(Value)) %>%
   spread(habitat, Value) %>%
   merge(sitescores) %>%
-  mutate(Novelty = N1 - Value) %>%
+  mutate(Novelty = forest - Value) %>%
   group_by(id, habitat) %>%
-  summarise(CPN1 = sum(abs(Novelty)))  -> cpn1
+  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
 
-### Novelty (versus N2)
+### Novelty (versus meadow)
 
 sitescores %>%
-  filter(habitat %in% c("N2")) %>%
[…]

### 2024-07-15 src/ordination.R

@@ -1,248 +1,0 @@
-library(tidyverse)
-
-c("forestgreen",  
-  "gold", 
-  "#40E0D0",
-  "#A020F0",
-  "#5CACEE",
-  "#551A8B") -> colores
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-### NMDS plant communities
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(species) %>%
-  select(id, taxon, cover) %>%
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 2) -> 
-  nmds # Ordination output
-
-### Site scores
-
-vegan::scores(nmds, "sites") %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
-  merge(header) %>%
-  dplyr::select(id, habitat, Dim.1, Dim.2) %>%
-  mutate(habitat = fct_relevel(habitat, "N1", "N2", "N3", "N4B", "N4A", "N4C")) %>%
-  mutate(habitat = fct_recode(habitat, "Forests" = "N1", "Meadows" = "N2",
-                              "Parks" = "N3", "Roadsides" = "N4B",
-                              "Residential" = "N4A", "Industrial" = "N4C")) -> sitescores
-
-### Species scores
-
-vegan::scores(nmds, "species") %>%
-  data.frame() %>%
-  rownames_to_column("Taxon") %>%
-  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
-  mutate(Species = Taxon) %>%
-  separate(Species, into = c("G", "S")) %>%
-  mutate(Species = paste(substr(G, 1, 3), substr(S, 1, 3), sep = "")) %>%
[…]

### 2024-07-15 src/soils.R

@@ -1,114 +1,0 @@
-library(tidyverse)
-
-c("forestgreen",  
-  "gold", 
-  "#40E0D0",
-  "#A020F0",
-  "#5CACEE",
-  "#551A8B") -> colores
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-### PCA soils
-
-header %>% select(id, habitat) %>% merge(soils) -> datos
-datos %>% select(-c(id, habitat)) %>% FactoMineR::PCA(graph = FALSE) -> pca1
-cbind((datos %>%  dplyr::select(id, habitat)), data.frame(pca1$ind$coord[, 1:2])) %>%
-  mutate(habitat = fct_relevel(habitat, "N1", "N2", "N3", "N4B", "N4A", "N4C")) %>%
-  mutate(habitat = fct_recode(habitat, "Forests" = "N1", "Meadows" = "N2",
-                              "Parks" = "N3", "Roadsides" = "N4B",
-                              "Residential" = "N4A", "Industrial" = "N4C")) -> pcaInds
-pca1$var$coord[, 1:2] %>% data.frame %>% rownames_to_column(var = "Variable") %>%
-  mutate(Dim.1 = ifelse(Variable == "Pb", Dim.1 + .05, Dim.1)) -> pcaVars
-
-ggplot(pcaInds, aes(x = Dim.1, y = Dim.2)) +
-  coord_fixed() +
-  geom_hline(yintercept = 0, linetype = "dashed") +
-  geom_vline(xintercept = 0, linetype = "dashed") +
-  geom_segment(data = pcaVars, aes(x = 0, y = 0, xend = 6*Dim.1, yend = 6*Dim.2)) +
-  geom_point(aes(fill = habitat), color = "black", size = 5, shape = 23) +
-  geom_label(data = pcaVars, aes(x = 6*Dim.1, y = 6*Dim.2, label = Variable),  show.legend = FALSE, size = 3) +
-  ggthemes::theme_tufte() + 
-  theme(text = element_text(family = "sans"),
-        strip.background = element_blank(),
-        legend.position = "bottom", 
-        #legend.direction = "vertical",
-        legend.title = element_blank(),
-        legend.spacing.x = unit(0, "mm"),
-        legend.spacing.y = unit(0, "mm"),
-        legend.text = element_text(size = 10), 
-        panel.background = element_rect(color = "black", fill = NULL),
[…]

### 2024-07-15 src/species-pools.R

@@ -1,146 +1,0 @@
-library(tidyverse); library(ggvenn)
-
-read.csv("data/header.csv", fileEncoding = "latin1") %>%
-  filter(habitat != "N4D") -> header
-read.csv("data/species.csv", fileEncoding = "latin1") %>% group_by(id, taxon) %>%
-  summarise(cover = sum(cover_percent)) -> # merge layers
-  species
-read.csv("data/fungi-species.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria-species.csv", fileEncoding = "latin1") -> bacteria
-
-### Plant species pools
-
-list(
-  Forests = 
-    header %>% 
-    filter(habitat == "N1") %>% 
-    select(id) %>% 
-    merge(species) %>% 
-    pull(taxon) %>% 
-    unique, 
-  Meadows = 
-    header %>% 
-    filter(habitat == "N2") %>% 
-    select(id) %>% 
-    merge(species) %>% 
-    pull(taxon) %>% 
-    unique, 
-  `Urban habitats` = 
-    header %>% 
-    filter(habitat %in% c("N3", "N4A", "N4B", "N4C")) %>% 
-    select(id) %>% 
-    merge(species) %>% 
-    pull(taxon) %>% 
-    unique) -> df1
-
-### Fungal species pools
-
-list(
-  Forests = 
-    header %>% 
-    filter(habitat == "N1") %>% 
-    select(id) %>% 
-    merge(fungi) %>% 
-    pull(taxon) %>% 
-    unique, 
-  Meadows = 
-    header %>% 
-    filter(habitat == "N2") %>% 
-    select(id) %>% 
-    merge(fungi) %>% 
-    pull(taxon) %>% 
-    unique, 
-  `Urban habitats` = 
-    header %>% 
-    filter(habitat %in% c("N3", "N4A", "N4B", "N4C")) %>% 
-    select(id) %>% 
-    merge(fungi) %>% 
-    pull(taxon) %>% 
-    unique) -> df2
-
-### Bacterial species pools
-
-list(
-  Forests = 
-    header %>% 
-    filter(habitat == "N1") %>% 
-    select(id) %>% 
-    merge(bacteria) %>% 
-    pull(taxon) %>% 
-    unique, 
-  Meadows = 
-    header %>% 
-    filter(habitat == "N2") %>% 
-    select(id) %>% 
-    merge(bacteria) %>% 
[…]

### 2024-07-15 src/temperatures.R

@@ -1,51 +1,0 @@
-library(tidyverse); library(lubridate)
-
-c("#FFA500",  
-  "yellow", 
-  "#B3EE3A", 
-  "#40E0D0",
-  "#5CACEE", 
-  "#A020F0",
-  "#551A8B") -> colores
-
-## Function reads and formats GP5WSHELL csv export files
-
-read_gp5 <- function(x)
-{
-  require(tidyverse)
-  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
-    dplyr::rename(temperature = X.1.oC) %>%
-    mutate(time = as.POSIXct(Time, 
-                             format =" %d.%m.%Y %H:%M:%S", 
-                             tz = "UTC"),
-           logger = gsub("\\_.*", "" , x)) %>%
-    dplyr::select(logger, time, temperature)
-}
-
-## Read logger header data
-
-read.table("data/temperature/Loggers.txt", sep = "\t", header = TRUE, fileEncoding = "latin1") %>%
-  mutate(Installed = as.POSIXct(Installed, 
-                                format ="%d/%m/%Y", 
-                                tz = "UTC")) %>%
-  rename(habitat = N.type) %>%
-  rename(logger = Logger) -> header
-
-## Read and bind all loggers, merge with logger header data
-
-wd <- getwd()
-setwd(paste(wd, "/data/temperature", sep = ""))
-do.call(rbind, lapply(list.files(pattern = "*.csv"), read_gp5)) %>% 
-  merge(header, by = "logger") %>%
-  na.omit %>% 
-  filter(time > (Installed + 60*60*24*7)) %>% # Remove records up to one week after installation
-  select(ID, plot, habitat, time, temperature) ->
-  logs
-setwd(wd)
-
-logs %>%
-  ggplot(aes(time, temperature, color = habitat)) + 
-  facet_wrap(habitat ~ plot) +
-  geom_line() +
-  scale_color_manual(values = colores) + 
-  coord_cartesian(ylim = c(0, 30))

### 2024-07-16 src/fig2-soils.R
minor figure edits

@@ -83,7 +83,7 @@ logs %>%
         axis.title.y = element_text(size = 10),
         axis.text.x = element_text(size = 10, color = "black", angle = 45, vjust = 1, hjust=1),
         axis.text.y = element_text(size = 10, color = "black"),
-        plot.margin = unit(c(0.1,0.1,0.1,0.1), "cm")) +
+        plot.margin = unit(c(0.1,0.1,0.1,0.5), "cm")) +
   geom_hline(yintercept = 0, linetype = "dashed") -> F2A; F2A
 
 ### Soil bioclimatic indices

### 2025-03-17 README.Rmd
submission UFUG

---
title: "README.md"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![](xove.JPG)

# Compositional novelty of plant, fungal and bacterial communities across urban habitats

This repository stores all the information related to the manuscript *Compositional novelty of plant, fungal and bacterial communities across urban habitats*, including the datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript.

## Contents

This repository is organised following the advice of [Wilson et al. 2017](https://doi.org/10.1371/journal.pcbi.1005510) for recording and storing research projects.

The following materials are available in the folders of this repository:

* `data` Data files including the soil, plant, fungi and bacteria data.
* `src` Scripts in `R` language used to perform the analyses and create the figures.
* `results` Output of the `R` scripts, including figures and supplementary material.
* `doc` Files to create the [manuscript](https://github.com/efernandezpascual/xixon/blob/master/doc/manuscript.md) using `Rmarkdown`.

## Citation

Please cite the repository, datasets and article as:

> Fernández-Pascual E, Ferencova Z, González-García, Jiménez-Alfaro B. Compositional novelty of plant, fungal and bacterial communities across urban habitats

### 2025-03-17 doc/manuscript.Rmd
minor figure edits
submission UFUG

--- 2024-07-11 doc/manuscript.Rmd
@@
 ---
-title: "Soil drivers of plant, fungal and bacterial community novelty in urban habitats of the city of Xixón, Spain"
+title: "Compositional novelty of plant, fungal and bacterial communities across urban habitats"
 output: 
-  bookdown::github_document2:
+  bookdown::word_document2:
     reference_docx: template.docx
@@
 bibliography: refs.bib
-csl: journal-of-vegetation-science.csl
+csl: urban-forestry-and-urban-greening.csl
 link-citations: yes
@@
 
-Running title: Urban drivers of plant, fungal and bacterial novelty
+Running title: Compositional novelty across urban habitats
 
-Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^2^ (<https://orcid.org/0000-0002-0915-0916>),
-Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)
+# Abstract (250 words)
 
-^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain
+We tested the hypothesis that cities are composed of different degrees of ecological novelty by studying four urban habitats with differing degrees of management and human legacy: park lawns, roadsides, residential vacant lots, and industrial vacant lots.
+We focused on community compositional novelty, by comparing the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forest remnants and hay meadows.
+We used a compositional novelty index based on multidimensional ordination, which is straightforward to calculate and only requires  species co-occurrence data for urban and reference habitats.
+As expected, (1) plant communities displayed the highest novelty and soil bacteria the lowest; (2) urban communities were markedly different from forest communities and relatively more similar to meadow communities; and (3) the degree of compositional novelty was highest in industrial vacant lots.
+On the contrary, managed park lawns, which we had expected to be highly novel, were relatively close to hay meadows.
+The lowest novelty among urban habitats was recorded in residential vacant lots, which had biological communities that more closely resembled those of pre-urban habitats.
+Our results highlight the effect of habitat type as a major driver of urban community composition and compositional novelty.
+These results suggest that city biodiversity can be enhanced by an integrative approach to the urban landscape that favors habitat heterogeneity by passive rewilding of managed park lawns, non-intervention on residential vacant lots, direct restoration of industrial vacant lots, and conservation of natural and agricultural habitat remnants as sources of native species.
 
-Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain.
-Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}.
-Telephone: +34985104781.
+# Keywords (7)
 
-# Author contributions
+Urban biodiversity, vegetation, soil fungi, soil bacteria, DNA metabarcoding, soil pollution, novel ecosystems
 
-Eduardo Fernández-Pascual: Conceptualization; Methodology; Investigation; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing.
-Zuzana Ferencova: Conceptualization; Methodology; Investigation; Data Curation; Writing – Review & Editing.
-Víctor González-García: Investigation; Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.
+# Highlights (3 to 5, 85 words each)
 
-# Acknowledgements
-
-María García Fernández, Sara Gutiérrez Cruz and Abelardo Sigüenza Solís helped with the field surveys.
-
-# Funding
-
-Fundación Biodiversidad, Ministerio para la Transición Ecológica y el Reto Demográfico, Gobierno de España [XIXÓN ECORRESILIENTE, PRTR, NextGenerationEU].
-Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico [SV-23-GIJON-JBA].
-
-# Conflict of interest statement
-
-None declared.
-
-# Data availability statement
-
-Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at Zenodo: https://zenodo.org/doi/10.5281/zenodo.11613571.
+* A compositional novelty index based on multidimensional ordination is straightforward to calculate and only requires species co-occurrence data for urban and reference pre-urban habitats. 
+* Novelty was the highest for plant communities and the lowest for soil bacteria. 
+* Urban communities were markedly different from forest communities and relatively more like meadow communities.
+* Compositional novelty was highest in industrial vacant lots, and lowest in residential vacant lots. 
+* Park lawns were relatively close in composition to hay meadows.
 
@@
 
-In post-industrial landscapes affected by the phenomenon of urban contraction, land located in the heart of cities remains unused for long periods (Haase et al., 2014) .
-When the use of any urban plot ceases, it is colonized by plants from the regional flora capable of living in urban environments (Johnson et al., 2017) .
-These urban plants are adapted to disturbances and can grow in degraded soils; some even accumulate heavy metals, making them nature-based solutions for bioremediation of degraded lands (Song et al., 2019) .
-Species traits, along with the legacy of human impact, determine which regional flora species colonize which plots (Swan et al., 2021) .
-When the human legacy is intense, the plant communities that develop will be novel communities, different from the historical reference communities that existed before human intervention (Gedan and Fernández-Pascual, 2019) .
-The novelty of the communities can be reflected in their floristic, compositional and phylogenetic composition, including their proportion of exotic species (Lososová et al., 2012) .
-In any case, proper management of urban communities can turn them into resources with high ecological and social potential (Anderson and Minor, 2017).
-(Kowarik, 2018) : in the Ibero-Atlantic ecoregion, anthropogenic habitats can host up to a third of the ecoregional flora.
+Urban development affects landscape composition and related biodiversity patterns worldwide [@RN5627].
+Urbanization converts natural and agricultural habitats into a fine-scale mosaic of diverse urban habitats varying in size, fragmentation, past and present land uses, and degrees of human intervention [@RN5527].
+At the same time, novel urban habitats experience climatic differences from the surrounding landscape, as well as high within-city microclimatic variation [@RN5601; @RN5603].
+Urban soils likewise vary in permeability, compaction, organic matter content and human-concentrated chemicals [@RN5623; @RN5603].
 
-The ability of plants to colonize any habitat, including urban ones, depends largely on their seed regeneration niche, that is, on their dispersal, persistence, germination and establishment functions (Grubb, 1977, Saatkamp et al., 2019) .
-However, although it is known that urban species often share a series of functional and ecological traits (Aronson et al., 2007, Lososová et al., 2006) , little is known about their seed regeneration niche.
-Especially, information on its germination traits is lacking.
-In species living in non-anthropogenic environments, germination phenology is regulated by environmental factors (water, temperature, light) to ensure that germination occurs at the right time to maximize subsequent seedling survival (Carta et al., 2022a) .
-Compared to the large amount of data on the germination of species from non-anthropogenic habitats, little is known about what factors regulate the germination of urban species.
-Considering other habitats affected by humans (Fernández-Pascual et al., 2022) , it could be expected that urban seeds have their germination unrelated to macroclimatic signals and show tolerance to stress due to heat, drought or heavy metals.
-However, this hypothesis needs to be tested through a synthesis of the seed regeneration niche in urban environments.
+Urban habitats are colonized by organisms from the regional species pool that are capable of living in urban environments [@RN2300].
+The process of colonization requires species to pass through a series of hierarchical filters (biogeographical, abiotic, biotic and human) resulting in the formation of an urban species pool and the assembly of specialized urban communities [@RN5505; @RN2306].
+Urbanization promotes specialized ruderal species and traits [@RN5613; @RN5622; @RN5679], and thus urban communities are rich in species that can tolerate urban disturbance, drought and pollution [@RN5603].
+For this reason, the urban biota can be a source of nature-based solutions, such as the bioremediation of degraded soils [@RN5447; @RN5449] and ecosystem design [@RN5628; @RN5629].
+Urban communities also host a high proportion of the regional biodiversity [@RN5600; @RN5682] and make up a major part of the supply of natural elements in densely populated landscapes [@RN5445].
+Proper management of urban communities can turn them into assets with high ecological and social returns [@RN5446; @RN5676; @RN5678].
 
-Cities 
-1. highly fragmented landscapes
-2. mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use)
-3. subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems
-4. comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure
-5.
-contains a multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix
+The diversity of urban communities can be characterized along two axes: naturalness and novelty [@RN5445].
+Naturalness refers to the degree of self-organization, i.e., the extent to which community composition and functions are self-assembled or dependent on direct human inputs [@RN5445].
+Novelty refers to the degree of differentiation from the historical reference communities that existed in the area before urbanization [@RN5673; @RN5674].
+For example, urban communities under direct human management (e.g., park lawns) should have low naturalness and high novelty.
+When management stops, naturalness increases, and the community can either transition back to a state resembling the historical reference (e.g., a low novelty hay meadow) or into a novel assemblage of species (e.g., a high novelty urban grassland with non-native species) [@RN5673; @RN5674].
+The fate of a specific urban site will depend on the interaction between the legacy of human intervention and inherent site characteristics [@RN2305].
+Intense human legacies usually push communities into novel states, often dominated by non-native and otherwise unwanted taxa [@RN5603; @RN5604], which usually require more intensive and costly actions to restore the land for conservation purposes [@RN2305].
+Understanding the patterns and drivers of urban community novelty can help to predict the restoration and rewilding potential of vacant urban spaces, facilitating their integration into the management of biodiversity in cities.
 
-Cities are comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure.
-They are mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use) which are subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems (Kowarik, 2005).
-
-The “Four Nature approach” narrows down the variety of transformational stages in urban settings to four major types that can supply wilderness in urban region (Kowarik 1992, 2005).
-Each of the four types relates to nature in general but results from different trajectories in human-nature interactions:
-
-Spontaneous vegetation appears in different habitat types in the urban matrix,.
-Multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix.
-The spontaneous herbs could provide the very interesting pool of species for ecological restoration of the urban area.
-Their ecological value has been clear since many years because their contribution to the ecosystem services in urban areas.
-Although their aesthetic value has been highlighted repeatedly, their acceptance by the residents in urban areas has been questioned.
-Many residents associate the presence of spontaneous plants with the neglect in maintenance by public (municipal) management service.
-Spontaneous native urban vegetation has commonly been described as demonstrating resilience (Ignatieva et al., 2000) and exhibiting adaptations to human disturbance (Lundholm and Marlin, 2006; Sukopp, 2004).
-
-Urban species pool is comprised of native species present before the urban development, native species developed because the new urban ecosystem and introduced species by human activity.
-However, just some of them can adapt to urban environment.
-
-IGS (informal green space) consists of any urban space with a history of strong anthropogenic disturbance that is covered at least partly with non-remnant, spontaneous vegetation (Del Tredici, 2010).
+The goal of this article was to test the hypothesis that urban habitats are comprised of different degrees of ecological novelty [@RN5674].
+We addressed this hypothesis by studying four urban habitats representing different land uses and human legacies, from managed park lawns to relatively unmanaged strips of land along transportation corridors, physically altered soils in residential vacant lots, and highly contaminated industrial brownfields.
+To address community compositional novelty, we compared the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forest remnants and hay meadows.
+We expected novelty to be the highest for plants, which in many cases are the object of direct management and introduction [@RN5493]; and lowest for bacteria, which should be less affected than fungi by the usually alkaline urban soils [@RN5603].
+We expected novelty to be higher when urban habitats are compared to natural habitats (forests), and lower when compared to agricultural habitats (meadows).
+We also expected novelty to be higher in the actively managed habitat (park lawns) and in the unmanaged habitat with the highest human legacy (industrial brownfields).
+Finally, we expected that compositional novelty would be related to soil physicochemical properties, reflecting the major aspect of human legacy in urban habitats.
+To test these expectations, we tackled three main research goals: (1) to characterize the variation in soil properties across urban and pre-urban habitats; (2) to characterize the variation in plant, fungal and bacterial communities across urban and pre-urban habitats; and (3) to measure the compositional novelty across habitats and taxonomic groups.
 
@@
 
-Gijón/Xixón is a medium-sized coastal city (population c.
-270,000 inhabitants) located in the Bay of Biscay, within the central Asturias metropolitan area of northern Spain (population c.
-800,000 inhabitants).
-The city belongs to the Cantabrian Mixed Forests ecoregion [@RN4667], at the border between the temperate and Mediterranean climatic zones of Europe, and has a temperate oceanic climate with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm).
-Most of the city is built on alluvial and estuarine deposits, sand, or limestone bedrock.
-The potential pre-human vegetation would be dominated by deciduous mixed forests, together with riparian forests, wetlands and coastal dunes.
+Gijón/Xixón is a medium-sized coastal city (c.
+270,000 inhabitants; Instituto Nacional de Estadística) located in the Bay of Biscay, within the central Asturias metropolitan area (c.
+800,000 inhabitants) of northern Spain.
+The city belongs to the Cantabrian Mixed Forests ecoregion [@RN5493; @RN5600], at the border between the temperate and Mediterranean climatic zones of Europe.
+Climate is temperate oceanic with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm; Agencia Estatal de Meteorología).
+Most of the city is built on alluvial and estuarine deposits, sand or limestone bedrock.
+The potential pre-human vegetation was likely dominated by deciduous mixed forests, together with riparian forests, wetlands and coastal dunes.
 The area has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450].
 The city itself has been inhabited at least since Roman times [@RN5532].
-During the medieval and early modern era the urban center was limited to the original walled city, with fishing as the main economic activity.
+During the medieval and early modern times, the urban center was limited to the original walled city, with fishing being the main economic activity.
 Expansion beyond the original core started in the 18th and 19th centuries.
-Exposure to global trade networks and the mild climate turned the region into a hotspot for biological invasions, especially from the Americas and Asia [@RN5493].
-In 1852, the city was connected by railway and became a major port for the regional mining and industrial centers.
-During the 20th century the city experienced significant industrial development (e.g.
-canning, coal, steel, shipbuilding) and rapid population growth, becoming the most populated city in the Principality of Asturias [@RN5534; @RN5533].
+Exposure to global trade networks and the warm temperate climate turned the region into a hotspot for biological invasions, especially from the Americas and Asia [@RN5493].
+In 1852, the city was connected by railway to regional mining and industrial centers and became a major shipping port [@RN5675].
+During the following century the city experienced significant industrial development (e.g., canning, coal, steel, shipbuilding) and rapid population growth, becoming the most populated city in the Principality of Asturias [@RN5683].
 Industrial and urban growth brought environmental problems including loss of natural wetlands and air and watercourse pollution [@RN5535].
-Like the rest of its metropolitan area, the city has been affected by recent post-industrial land-use changes which have left large areas with abandoned industrial sites (i.e.
-brownfields) [@RN5448; @RN5449].
+Starting in the 1970s, the city and the metropolitan area have been affected by post-industrial land-use changes leaving large abandoned industrial sites (i.e., brownfields) [@RN5448; @RN5449].
 
@@
 
-To survey urban biodiversity across habitats, we defined six habitats types (**Figure 1A**) representing different states within the four-nature classification of urban wilderness [@RN5445]: (1) forest patches < 10 km from the city center, i.e.
-type-1 nature in remnants of the pre-urban vegetation; (2) hay meadows, i.e.
-type-2 nature in remnants of human-made ecosystems resulting from agricultural habitat transformation; (3) parks, i.e.
-type-3 nature in managed green spaces created after habitat destruction; (4) roadsides, i.e.
-type-4 nature in relatively unmanaged strips of land alongside roads, railways, etc; (5) residential lots, i.e.
-type-4 nature in unmanaged vacant lots of previous residential use; and (6) industrial lots, i.e.
-type-4 nature in unmanaged vacant lots of previous  industrial use.
+To survey urban biodiversity across habitats, we defined six habitat types (**Figure 1A**) representing different theoretical states of naturalness and novelty [@RN5445]: (1) forest patches \< 10 km from the city center, i.e., remnants of pre-urban natural habitats; (2) late mown hay meadows, i.e., remnants of pre-urban agricultural habitats; (3) park lawns, i.e., green spaces created and maintained by humans; (4) roadsides, i.e., relatively unmanaged strips of land alongside roads and train tracks; (5) residential vacant lots, i.e., unmanaged land which has suffered some degree of soil physical disturbance and is waiting to be urbanized; and (6) industrial vacant lots, i.e., unmanaged land with an intense legacy of soil disturbance and contamination.
 
-## Selection of sampling plots
+## Sampling plots
 
-During June 2023, we surveyed the city to establish five 4 x 4 m sampling plots per each of the six habitat types (n = 30 plots).
-The plots were distributed across the city (**Figure 1B**) and represented the variation of vegetation types found within each habitat type.
-In two plots per habitat (**Figure 1B**), we buried a datalogger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil.
-Dataloggers recorded temperature once every hour and stayed on site for approximately one year, after which we retrieved them and downloaded their records to describe the thermal environment of the city and the habitats.
+During June 2023, we surveyed the city to establish five sampling plots (4 x 4 m) per each of the six habitat types (n = 30 plots).
+The plots were distributed across the city (**Figure 1B**) and represent the variation of vegetation types found within each habitat type.
+In one plot per habitat (**Figure 1B**), we buried a datalogger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil.
+Dataloggers recorded temperature hourly for one year (1 Jan to 31 Dec 2024), after which we downloaded their records to describe the thermal environment of the city and the habitats.
+
+```{r fig1, echo = FALSE, fig.cap = "Pre-urban and urban habitats included in this study.
+(A) Representative pictures of each habitat, indicating their level of naturalness (i.e., degree of ecosystem self-organization) and novelty (i.e., degree of divergence from pre-human reference ecosystems).
+(B) Location of the sampling plots within the city of Gijón/Xixón, Asturias, Spain, Western Europe.
+Each plot (dots) is colored according to habitat type.
+A thermometer to the right of the sampling plot indicates that a soil temperature datalogger was installed there."}
+knitr::include_graphics("../results/figures/Fig1.png")
+```
 
@@
 
-From each of the 30 plots, we collected 500 g of soil for analysis of chemical composition and physical soil properties.
-We took the samples from 5 points in each plot (center + 4 cardinal points).
-We removed stones and other debris particles, sieved the soil in a 5 mm mesh and homogenized the sample in a plastic bag.
-We cleaned and sterilized all tools with water and ethanol between samples.
-We sent the samples for analysis at INEA - Escuela Universitaria de Ingeniería Agrícola, Universidad Pontificia de Comillas.
-Soil analysis included pH, texture, bulk density, electrical conductivity, organic matter content, total nitrogen, C/N ratio, active limestone, total carbonates, assimilable phosphorus, K, Ca, Mg, Na and heavy metals (Zn, Cr, Cu, Cd, Hg, Pb, Ni).
-
-> Añadir detalles de los métodos de suelo (Zuzana)
+From each of the 30 plots, we collected 500 g of soil to analyse soil physicochemical properties.
+We took the samples from 5 points in each plot (center + 4 cardinal points), down to a depth of 10 cm.
+We removed stones and other debris, sieved the soils in a 5 mm mesh and homogenized the samples in a plastic bag.
+Samples were analyzed at INEA Laboratory (Valladolid, Spain), following the standard methodology of the Spanish Ministry of Agriculture [@RN5599].
+Soil analysis included pH, texture, bulk density, electrical conductivity, organic matter content, total N, C/N ratio, active limestone, total carbonates, assimilable P, K, Ca, Mg, Na and heavy metals (Zn, Cr, Cu, Cd, Hg, Pb, Ni).
 
@@
 In each plot, we recorded all the spontaneous vascular plant species, as well as the cover (%) of each species in the plot.
-For plant species names we followed Euro+Med [@RN5044].
-Additionally, from each of the 30 plots, we collected soil samples for DNA metabarcoding analysis of soil fungal and bacterial communities.
-We performed the sampling in sterile conditions, i.e.
-before entering the plot for any of the other tasks.
+Plant species names follow Euro+Med [@RN5044].
+Additionally, from each of the 30 plots, we collected soil samples for DNA metabarcoding of fungal and bacterial (including Archaea) communities.
+We performed the sampling in sterile conditions, and always before entering the plot for any of the other tasks.
+We cleaned and sterilized all tools with water and ethanol between samples.
 In each plot, we took samples from the four cardinal points, 0.5 m inwards from the edge, at a depth of 5 cm.
 Before taking the samples, we removed the vegetation with a sterile tool.
-Then, we collected the soil samples in a 50 ml Falcon tube, mixed them, homogenized them with a lancet and finally stored the sample into a 2 ml Eppendorf tube which we immediately froze.
-We sent the frozen tube to AllGenetics & Biology SL (www.allgenetics.eu) for DNA metabarcoding analysis.
+Then, we collected the soil samples in a 50 ml Falcon tube, mixed them, homogenized them with a lancet and finally stored the sample in a 2 ml Eppendorf tube which we immediately froze.
+The frozen tubes were sent to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis.
+DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi, and 16S for bacteria; sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size \< 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, inference of Amplicon Sequence Variants (ASVs), taxonomic assignment, number of reads per taxon and generation of rarefaction curves (see **Appendix S1** for extended DNA metabarcoding methodology).
 
-## DNA metabarcoding analysis
+## Statistical analysis and compositional novelty index
 
-DNA analysis by AllGenetics included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size < 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, taxonomic assignment, number of reads per taxon and generation of rarefaction curves.
+We performed all analyses using R version 4.3.1 [@RN5387] and the package *tidyverse* [@RN4662] for data manipulation and visualization.
+To describe soil temperatures as recorded by the dataloggers, we calculated the mean annual temperature, the mean diurnal range and the annual range.
+To assess the variability in soil physicochemical properties, we conducted a Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] and a correlation analysis to detect auto-correlated soil variables.
+Multivariate analyses of community composition were performed in *vegan* [@RN5386]: we used PERMANOVA to test differences between habitat types in their soil variables and their composition of plant species, fungal ASVs and bacterial ASVs;  Non-metric Multi-Dimensional Scaling (NMDS) ordination to assess variation in taxonomic composition of plants (abundance values of species), fungi (presence/absence of ASVs) and bacteria (presence/absence of ASVs); and environmental fitting to test the correlation between community ordination axes and soil physicochemical properties.
 
-*DNA isolation and quantification*.
-DNA was isolated from each soil sample using the DNeasy PowerSoil Pro DNA isolation kit (Qiagen) following the manufacturer’s instructions, and DNA was eluted in a final volume of 50 μL.
-In one sample the above-mentioned protocol did not yield a DNA extract of enough concentration, so DNA was isolated using the FastDNA Spin Kit for Soil (MPBio) following the manufacturer’s instructions, and DNA was eluted in a final volume of 70 μL.
-An extraction blank (Bex) was included in each round of DNA extraction, and treated as the plot samples to check for contamination.
-DNA concentration in each extract was quantified using the Qubit High Sensitivity dsDNA Assay (Thermo Fisher Scientific).
-
-*DNA metabarcoding library preparation and sequencing*.
-For fungal library preparation, a fragment of the nuclear ITS1 region of around 280-300 bp (including the primer sequences) was amplified using the following primers: forward - ITS1f (5’ CTTGGTCATTTAGAGGAAGTAA 3’) [@RN5548]; reverse - ITS2 (5’ GCTGCGTTCTTCATCGATGC 3’) [@RN5549].
-For prokaryotic library preparation, a fragment of the 16S rRNA gene of around 300 bp (including primer sequences) was amplified using the following primers: forward - 515F-Y (5’ GTGYCAGCMGCCGCGGTAA 3’) [@RN5536]; reverse - 806RB (5’ GGACTACNVGGGTWTCTAAT 3’) [@RN5537].
-These primers also included the Illumina sequencing primer sequences attached to their 5’ ends.
-In the first amplification step, PCRs were carried out in a final volume of 12.5 μL, containing 1.25 μL of template DNA (1:2 diluted in PCR round 3; not diluted in PCR round 11), 0.5 μM of the primers, 3.13 μL of Supreme NZYTaq 2x Green Master Mix (NZYTech), and ultrapure water up to 12.5 μL.
-The fungal reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 35 cycles of 95 ºC for 30 s, 49 ºC for 45 s, 72 ºC for 45 s, and a final extension step at 72 ºC for 7 min.
-The bacterial reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 25 cycles of 95 ºC for 30 s, 47 ºC for 45 s, 72ºC for 45 s, and a final extension step at 72 ºC for 7 min.
-The oligonucleotide indices that are required for multiplexing different libraries in the same sequencing pool were attached in a second amplification step with identical conditions but only 5 cycles and 60 ºC as the annealing temperature.
-A negative control that contained no DNA (BPCR) was included in every PCR round to check for contamination during library preparation.
-Library size was verified by running the libraries on 2 % agarose gels stained with GreenSafe (NZYTech) and imaging them under UV light.
-Then, libraries were purified using the Mag-Bind RXNPure Plus magnetic beads (Omega Bio-tek), following the instructions provided by the manufacturer.
-Finished libraries were pooled in equimolar amounts according to the results of a Qubit dsDNA HS Assay (Thermo Fisher Scientific) quantification.
-The pool was sequenced in a fraction of a NovaSeq PE250 flow cell (Illumina) aiming for a total output of 2 gigabases.
-Any potential traces of adapter dimers were removed using Cutadapt v3.5 [@RN5538].
-Quality of the sequencing data was assesed with the software FastQC [@RN5539].
-
-*Processing of sequencing data and inference of ASVs*.
-Cutadapt v3.5 was used to trim off non-biological DNA sequences (primers, indices, and sequencing adapters).
-Then, amplicon reads were processed using the DADA2 pipeline [@RN5541] implemented in QIIME 2 (release 2023.7) [@RN5540].
-The first step in the DADA2 pipeline was to remove PCR primers and filter reads according to their quality.
-After checking the read quality profiles, forward reads were truncated at position 175, and reverse reads at position 131.
-The next step was to denoise and infer Amplicon Sequence Variants (ASVs).
-Error rates were learned from the dataset to denoise, using the parametric error model implemented in DADA2.
-Dereplication of the dataset (i.e.
-combination of all identical reads into unique reads) was carried out to reduce computational effort.
-These dereplicated forward and reverse reads were used to infer ASVs with the core sample inference algorithm [@RN5541].
-Subsequently, corresponding R1 and R2 reads were merged into pairs with a minimum overlap of 12 identical base pairs.
-The final step in the DADA2 pipeline was to reduce the impact of artefacts (chimaeras) in the dataset.
-The final output of the DADA2 pipeline was a table containing the count of reads of every observed ASV in each sample, as well as a table listing all ASVs and their corresponding representative sequences.
-
-*Taxonomic assignment*.
-Taxonomic assignment of each ASV was performed using a pre-trained classifier of the UNITE reference database for fungi [@RN5550] (release May 2021); and of the SILVA reference database for bacteria [@RN5542] (release 138.1 August 2020).
-To compare the representative sequences of the ASVs to the reference database, and compute the taxonomic assignment,the algorithm sklearn [@RN5543] implemented in QIIME 2 was applied, setting a minimum confidence level of 0.8.
-Several filters were applied to exclude: (1) singletons (i.e.
-ASVs containing only one member sequence in the whole data set); (2) ASVs occurring at a frequency below 0.01 % in each sample, to reduce the mistagging bias [@RN5544]; (3) from the bacterial dataset, non-prokaryotic ASVs such as eukaryotic sequences of plastid [@RN5545] and mitochondrial [@RN5546] origin; (4) those assigned only at domain (Bacteria) or kingdom (Fungi) level; (5) ASVs present in the negative controls that were more abundant in the negative controls than in the samples, indicating contamination [@RN5547].
-
-*Alpha rarefaction plots*.
-Alpha rarefaction plots showed that the sequencing depth to reach saturation was 30,000 reads for the fungi and 22,000 reads for the bacteria, above the sequencing depth of all plots for fungi (mean = 114,919 reads; min = 79,045; max = 157,110) and bacteria (mean = 113,341 reads; min = 84,094; max = 171,404).
-
-> Rarefaction
-
-## Statistical analysis
+To calculate a compositional novelty index, we used the NMDS multidimensional distance from each urban plot (i.e., park, roadside, residential or industrial) to the centroid of the two reference pre-urban habitats (i.e., forests and meadows).
+We calculated compositional novelty separately for plant, fungal and bacterial communities.
+We tested the effect of taxonomic group, pre-urban reference habitat (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables on compositional novelty, by fitting a Generalized Mixed Model with the package *nlme* [@RN2333].
+The model included sampling plot as a random factor.
+We checked and confirmed model assumptions using the package *performance* [@RN5684].
 
@@
 
-## Soil temperatures and properties across habitats
+## Soil properties across habitats
 
-## Plant, fungal and bacterial diversity across habitats
+Soil temperature records (**Figure 2A**) indicated that forest soils were relatively buffered in their thermal range of variation, with a mean annual temperature of 13.9 ºC, a mean diurnal range of 1.8 ºC and an annual range of 10.8 ºC.
+Meadow and urban plots had more thermal oscillation, with a mean annual temperature between 12.4 ºC (roadsides) and 16.8 ºC (meadows), a mean diurnal range between 3.8 ºC (roadsides) and 4.6 ºC (residential lots), and an annual range between 16.5 ºC (residential lots) and 19.5 ºC (meadows).
 
-> PERMANOVAs
+PCA ordination of the soil physicochemical properties (**Appendix S2**) indicated that the first axis of variation, explaining 23% of the variance, ordered soils according to the concentration of heavy metals.
+The variables with the highest contribution to this axis (\> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr, which accounted for 66% of the variability of the first axis.
+This first axis did not separate urban from pre-urban habitats: for example, while some industrial lots had a very high concentration of Pb and other metals, the median Pb value of the industrial habitat was below the median value of forests (**Figure 2B**; **Appendix S2**).
 
-> Analysis with species, genera and ASVs
+The second PCA axis (**Appendix S2**) explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (\> 10%) being organic matter content, bulk density, N concentration and pH, which made up 59% of the axis' variability.
+This axis separated soils according to their habitat: for example, forests had high organic matter and N content, while industrial vacant lots had higher bulk density and pH (**Figure 2B**; **Appendix S2**).
 
-> Analysis with presence absence
+The third PCA axis (**Appendix S2**) explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (\>10%) being Mg content, Mg/K ratio and conductivity, together explaining 50% of the axis.
+In this case, like for the first axis, there was no clear separation between habitats, with industrial sites having some of the highest and lowest values (**Figure 2B**; **Appendix S2**).
 
-## Community novelty across habitats and taxonomic groups
+The analysis showed strong correlations within the groups of variables contributing to the first, second and third PCA axes (**Appendix S2**).
+To avoid autocorrelation issues, for further analyses we kept three variables which taken together provide a straightforward biological interpretation of patterns in soil composition: Pb content, organic matter content and Mg content (**Figure 2B**).
+As previously indicated by PCA, organic matter was the only one with a clear pattern of variation: a decrease from pre-urban to urban habitats (**Figure 2B**).
+PERMANOVA supported that there was not a significant difference between habitats in their soil physicochemical properties (100,000 permutations, R^2^ = 0.19, F = 1.15, p = 0.31).
 
-## Soil drivers of novelty
+```{r fig2, echo = FALSE, fig.cap = "Soil properties across habitats.
+(A) Temperature recorded every hour in one sampling plot per habitat from 1 Jan 2024 to 31 Dec 2024, indicating the hourly records (shaded colors) and the daily averages (full colors).
+(B) Soil physicochemical properties.
+Values are shown for the three most explanatory and uncorrelated variables: soil Pb content, soil organic matter content, and soil Mg content."}
+knitr::include_graphics("../results/figures/Fig2.png")
+```
+
+## Plant, fungal and bacterial communities across habitats
+
+The survey of spontaneous plant diversity yielded 216 species, with the average richness per plot being 26 (min = 10, max = 38).
+Of these species, 48% were found only in urban habitats (parks, roadsides, residential lots and industrial lots), 27% occurred only in pre-urban habitats (forests and meadows) and 25% were shared by both urban and pre-urban habitats (**Fig3A**).
+On average, the habitat with the highest richness per plot was industrial lots (29 species), followed by residential lots (27), roadsides (24), meadows (23), parks (20) and forests (18).
+The most frequent species were *Holcus lanatus* (22 plots), *Plantago lanceolata* (20), *Helminthotheca echioides* (16), *Dactylis glomerata* (15) and *Lotus corniculatus* (14).
+PERMANOVA supported that the plant species composition was significantly different across habitats (100,000 permutations, R^2^ = 0.42, F = 3.48, p = \< 0.001).
+NMDS ordination of the plant species composition (**Fig.
+3D**) indicated that the plots belonging to the same habitat were relatively grouped together.
+The major gradient in vegetation composition along the first axis was a separation between forest plots and the other habitats.
+Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that none of the soil factors had a significant correlation (R^2^ \< 0.14, p \> 0.05) with the composition of the plant communities.
+
+DNA metabarcoding of soil fungal communities resulted in 4,689 amplicon sequence variants (ASVs).
+Of these, 46% were only found in urban habitats (parks, roadsides, residential lots and industrial lots), 36% occurred only in pre-urban habitats (forests and meadows) and 18% were shared by both urban and pre-urban habitats (**Fig3B**).
+The average number of ASVs per plot was 340 (min = 174, max = 587), with the habitat averages decreasing from residential vacant lots (415) to forests (393), meadows (336), parks (323), industrial vacant lots (300) and roadsides (276).
+The majority of ASVs were assigned to Ascomycota (56%), Basidiomycota (24%), Rozellomycota (10%) and Glomeromycota (4%).
+Forest habitats had the highest proportion of Basidiomycota (25% vs.
+\< 17% in the other habitats) and the lowest proportion of Ascomycota (57% vs.
+\> 64%).
+The proportion of Rozellomycota was similar across habitats (from 6% in meadows to 11% in parks).
+The proportion of Glomeromycota varied between 5% in meadows to 0.5% in forests.
+PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.33, F = 2.39, p = \< 0.001).
+NMDS ordination of the fungal ASV composition (**Fig.
+3E**) behaved similarly to the ordination of plant communities, with plots tending to group by habitat, and the major compositional gradient reflecting the separation between forest plots and the rest.
+Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only organic matter had a significant correlation (R^2^ = 0.46, p \< 0.001) with the composition of the fungal communities.
+Low organic matter content was associated with the fungal communities from industrial vacant lots and roadsides (**Fig.
+3E**).
+
+For bacterial communities, DNA metabarcoding yielded 16,306 ASVs.
+Of these, 44% were found only in urban habitats (parks, roadsides, residential lots and industrial lots), 31% occurred only in pre-urban habitats (forests and meadows) and 25% were shared by both urban and pre-urban habitats (**Fig3C**).
+The average number of ASVs per plot was 1769 (min = 1478, max = 2074), with the habitat averages decreasing from roadsides (1848) to industrial lots (1794), residential lots (1792), meadows (1775), parks (1706) and forests (1697).
+The majority of ASVs were assigned to Proteobacteria (22%), Planctomycetota (15%), Actinobacteriota (15%) and Acidobacteriota (10%).
+Compared to other habitats, forests had higher proportions of Proteobacteria and Acidobacteriota, a lower proportion of Actinobacteriota, and a similar proportion of Planctomycetota.
+PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.32, F = 2.51, p = \< 0.001).
+NMDS ordination of the bacterial ASV composition (**Fig.
+3F**) resulted in less grouping by habitat than for plants and fungi, but still reflected a main compositional gradient from forests to industrial vacant plots.
+Like in the case of the fungal communities, environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only organic matter content had a significant correlation (R^2^ = 0.39, p = 0.001) with the composition of the bacterial communities.
+Low organic matter content correlated with the communities from industrial vacant lots and roadsides (**Fig.
+3F**).
+
+```{r fig3, echo = FALSE, fig.cap = "Plant, fungal and bacterial diversity across habitats.
+Top panels show Venn plots for the shared numbers of (A) plant species, (B) fungal ASVs and (C) bacterial ASVs between urban (parks, roadsides, residential lots and industrial lots) and pre-urban (forests, meadows) habitats.
+Bottom panels show biplots of community composition ordination produced by Non-metric Multi-Dimensional Scaling (NMDS).
+NMDs were conducted for (D) the plant survey (by species abundance), (E) the soil fungal survey (by ASV presence/absence) and (F) the soil bacterial survey (by ASV presence/absence).
+Colors indicate the habitat.
+In the biplots, arrows and labels indicate soil variables with a significant correlation with the composition of the communities.
+Only soil organic matter content and fungal-bacterial communities had a significant correlation."}
+knitr::include_graphics("../results/figures/Fig3.png")
+```
+
+## Compositional novelty across habitats and taxonomic groups
+
+We used the NMDS multidimensional compositional distance (calculated from the ordinations in **Fig.
+3D-F**) from each of the urban plots (i.e., park, roadside, residential or industrial) to the centroid of the two reference pre-urban habitats (i.e., forests and meadows) to calculate a compositional novelty index (**Fig.
+4**).
+Novelty was significantly different between taxonomic groups (F = 45.88, p \< 0.001), being higher for plants (2.5 ± 0.3 S.E.) than for fungi (2.2 ± 0.4 S.E.) and bacteria (1.6 ± 0.3 S.E.).
+Novelty was also significantly different depending on the pre-urban habitat that was used as a reference (F = 199.08, p \< 0.001), being higher compared to forests (2.6 ± 0.3 S.E.) than to meadows (1.5 ± 0.3 S.E.).
+Nonetheless, for each taxonomic group, there was a high positive correlation between the novelty compared against forests and meadows (r = 0.5 for plants; 0.7 for fungi; 0.9 for bacteria).
+Different urban habitats also differed significantly in their novelty (F = 9.51, p = 0.001), being higher in industrial vacant lots (2.5 ± 0.3 S.E.) than in roadsides (2.2 ± 0.4 S.E.), parks (2.0 ± 0.4 S.E.) and residential vacant lots (1.6 ± 0.3 S.E.).
+The soil physicochemical properties (Pb, organic matter, Mg) had no significant effect on compositional novelty (p \> 0.05).
+
+```{r fig4, echo = FALSE, fig.cap = "Compositional novelty for plants, fungi and bacteria.
+Compositional novelty was calculated as the NMDS multidimensional distance from each urban plot to the centroid of two reference pre-urban habitats (forests, panels A-C-E; and meadows, panels B-D-F).
+Novelty was calculated separately for plants (panels A-B), fungi (panels C-D) and bacteria (panels E-F)."}
+knitr::include_graphics("../results/figures/Fig4.png")
+```
 
@@
 
-The recorded flora represents just a subset of possible spontaneous vegetation because this study was carried on during limited time which covers just part of the growing season of the whole vegetation.
+This study shows the compositional novelty of biological communities across urban habitats when they are compared with two reference pre-urban habitats: forest remnants and agricultural hay meadows.
+As expected, (1) novelty was the highest for plant communities and the lowest for soil bacteria; (2) urban communities were markedly different from forest communities and relatively more similar to meadow communities; and (3) the degree of compositional novelty was the highest among industrial vacant lots.
+However, managed park lawns, which we had expected to be highly novel, were relatively close in composition to hay meadows.
+Among urban habitats, the residential vacant lots had the lowest novelty, with biological communities that more closely resembled those of the reference pre-urban habitats.
 
-There are some other habitats in the city which were not included in the present study, e.g. wetlands, beach, cliffs.
+## Widespread soil contamination across urban habitats
+
+Our results partly agreed with the expected effects of urbanization on soils [@RN5623; @RN5603]: we found more microclimatic variation, less organic matter content and more compacted and alkaline soils than their pre-urban counterparts.
+But, contrary to our expectations, urban and pre-urban soils did not show a consistent pattern of contamination by heavy metals, possibly because pollution is widespread across the city.
+Indeed, soil contamination had a weak but positive correlation (r = 0.25) with the geographical proximity to the hotspot of heavy industry in the western end of the city [@RN5683].
+This geographical correlation was especially strong for the residential lots (0.89), parks (0.68) and forests (0.68); and weak for the roadsides (0.10), industrial lots (0.03) and meadows (-0.14).
+Overall, the measured soil properties had a minor effect on the composition of the biological communities, as organic matter content was the only variable to influence the composition of soil fungal and bacterial communities.
+
+## Contrasting diversity patterns in plant-bacterial vs. fungal communities
+
+Our survey found 216 vascular plant species growing in the study plots, representing 9 % of the regional flora [@RN3300].
+The highest plant species richness was recorded in vacant industrial and residential plots, while forests displayed the lowest.
+It must be considered that vacant lots, especially those of industrial origin, had a high number of non-native species, including five that are classified as highly impactful invasives in the ecoregion:  *Buddleja davidii*, *Cortaderia selloana*, *Dittrichia viscosa*, *Paspalum dilatatum* and *Sporobolus indicus* [@RN5493].
+Still, vacant lots had more species than forests even considering only native plants.
+Contrary to plants, fungi had higher richness in our forest plots than in the roadsides or industrial vacant lots, with forests being especially rich in Basidiomycota, as found in other studies [@RN5630; @RN5607].
+This agrees with previous reports of fungal communities decreasing with urbanization [@RN5609].
+One potential explanation for this pattern is the tendency of urban soils to be alkaline, favoring bacteria over fungi [@RN5603; @RN5606].
+Indeed, in our study, organic matter content (highly and inversely correlated with pH) was the soil variable that had an effect on the composition of fungal and bacterial communities; and bacterial richness was higher in urban habitats versus forests (i.e., the opposite to fungi).
+
+## Relatively low novelty in parks lawns and residential vacant lots
+
+We used a compositional novelty index based on multidimensional ordination, which is straightforward to calculate and only requires data on species co-occurrence for urban and reference habitats.
+As such, this index can help to characterize and compare novelty across urban habitats, which is a major research goal for novel ecosystem research [@RN5673; @RN5674].
+Contrary to our expectations, managed park lawns showed a relatively low novelty when compared to agricultural meadows.
+This indicates that park lawns have a high potential for low-cost urban rewilding policies, such as low-frequency and late mowing, which could convert lawns into urban grasslands rich in native biota and resembling traditional hay meadows.
+Urban grasslands are able to hold a high portion of the plant diversity found in their rural counterparts [@RN5677], especially when seminatural habitat remnants are available in the neighboring landscape to serve as a source of species [@RN5611].
+We also found that residential vacant lots have a relatively low level of novelty, including a high richness of fungi and native plants.
+These temporarily unmanaged sites are therefore a highly valuable asset to integrate biodiversity in cities [@RN5445; @RN5446].
+Additionally, identifying synanthropic and native species able to thrive in these vacant lots can help to find candidate species for urban landscaping that require low levels of watering, fertilization, weeding or mowing [@RN5614; @RN5678].
+Such native and hardy seed mixes are important tools for urban greening, given the increasing awareness of the importance of using locally sourced seed for restoration and rewilding [@RN5617].
+
+## Conclusions
+
+Our results highlight the effect of habitat type as a major driver of urban community composition and compositional novelty, supporting the findings of previous studies [@RN5607; @RN5630; @RN5608].
+The high impact of habitat types on shaping urban communities and their compositional novelty may be related to the fact that anthropogenic vegetation is highly responsive to habitat-specific disturbance regimes [@RN5600], while soil microbial communities are in turn highly conditioned by plant inputs [@RN5630].
+These results suggest that city biodiversity can be enhanced by an integrative approach to the urban landscape that favors habitat heterogeneity by passive rewilding of managed park lawns, non-intervention on residential vacant lots, direct restoration of industrial vacant lots, and conservation of natural and agricultural habitat remnants as sources of native species [@RN5678; @RN5610; @RN5612].
+In this way, the potential of urban habitats can be mobilized so they continue to support ecosystem functioning in anthropogenic city landscapes  [@RN5680; @RN5681; @RN5605].
+
+# References
+
+::: {#refs}
+:::
+
+# Supplementary
+
+**Appendix S1** Extended methodology for the soil DNA metabarcoding analysis.
+
+**Appendix S2** Principal Component Analysis and correlation analysis of the soil physicochemical properties.

### 2025-03-17 results/supplementary/AppendixS1.Rmd
submission UFUG

@@
 ---
-title: "Appendix S1 Principal Component Analysis and correlation analysis of the soil physicochemical properties."
+title: "Appendix S1 Extended methodology for the soil DNA metabarcoding analysis"
 output:
@@
     toc: false
+    latex_engine: xelatex
+bibliography: ../../doc/refs.bib
+csl: ../../doc/urban-forestry-and-urban-greening.csl
 link-citations: yes
@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
+Samples were sent to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis.
+DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size < 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, taxonomic assignment, number of reads per taxon and generation of rarefaction curves.
 
-```{r message = FALSE, echo = FALSE, warning = FALSE}
-knitr::knit_hooks$set(inline = function(x) {
-  prettyNum(x, big.mark = ",")
-})
-```
+# DNA isolation and quantification
 
-```{r figS1a, out.width = "450px", echo = FALSE, fig.cap = "(A) Biplot of the first and second axes of a Principal Component Analysis of the soil physicochemical properties.
-Circles represent the sampling plots, coloured by habitat.
-Labeled arrows indicate the contribution of the soil variables to the axes."}
-knitr::include_graphics("figS1a.png")
-```
+DNA was isolated from each soil sample using the DNeasy PowerSoil Pro DNA isolation kit (Qiagen) following the manufacturer’s instructions, and DNA was eluted in a final volume of 50 μL.
+In one sample the above-mentioned protocol did not yield a DNA extract of enough concentration, so DNA was isolated using the FastDNA Spin Kit for Soil (MPBio) following the manufacturer’s instructions, and DNA was eluted in a final volume of 70 μL.
+An extraction blank (Bex) was included in each round of DNA extraction and treated as plot samples to check for contamination.
+DNA concentration in each extract was quantified using the Qubit High Sensitivity dsDNA Assay (Thermo Fisher Scientific).
 
-```{r figS1b, out.width = "450px", echo = FALSE, fig.cap = "(B) Biplot of the first and third axes of a Principal Component Analysis of the soil physicochemical properties.
-Circles represent the sampling plots, coloured by habitat.
-Labeled arrows indicate the contribution of the soil variables to the axes."}
-knitr::include_graphics("figS1b.png")
-```
+# DNA metabarcoding library preparation and sequencing
 
-```{r figS1c, out.width = "450px", echo = FALSE, fig.cap = "(C) Correlation heatmap of the soil variables."}
-knitr::include_graphics("figS1c.png")
-```
+For fungal library preparation, a fragment of the nuclear ITS1 region of around 280-300 bp (including the primer sequences) was amplified using the following primers: forward - ITS1f (5’ CTTGGTCATTTAGAGGAAGTAA 3’) [@RN5548]; reverse - ITS2 (5’ GCTGCGTTCTTCATCGATGC 3’) [@RN5549].
+For prokaryotic library preparation, a fragment of the 16S rRNA gene of around 300 bp (including primer sequences) was amplified using the following primers: forward - 515F-Y (5’ GTGYCAGCMGCCGCGGTAA 3’) [@RN5536]; reverse - 806RB (5’ GGACTACNVGGGTWTCTAAT 3’) [@RN5537].
+These primers also included the Illumina sequencing primer sequences attached to their 5’ ends.
+In the first amplification step, PCRs were carried out in a final volume of 12.5 μL, containing 1.25 μL of template DNA (1:2 diluted in PCR round 3; not diluted in PCR round 11), 0.5 μM of the primers, 3.13 μL of Supreme NZYTaq 2x Green Master Mix (NZYTech), and ultrapure water up to 12.5 μL.
+The fungal reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 35 cycles of 95 ºC for 30 s, 49 ºC for 45 s, 72 ºC for 45 s, and a final extension step at 72 ºC for 7 min.
+The bacterial reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 25 cycles of 95 ºC for 30 s, 47 ºC for 45 s, 72ºC for 45 s, and a final extension step at 72 ºC for 7 min.
+The oligonucleotide indices that are required for multiplexing different libraries in the same sequencing pool were attached in a second amplification step with identical conditions but only 5 cycles and 60 ºC as the annealing temperature.
+A negative control that contained no DNA (BPCR) was included in every PCR round to check for contamination during library preparation.
+Library size was verified by running the libraries on 2 % agarose gels stained with GreenSafe (NZYTech) and imaging them under UV light.
+Then, libraries were purified using the Mag-Bind RXNPure Plus magnetic beads (Omega Bio-tek), following the instructions provided by the manufacturer.
+Finished libraries were pooled in equimolar amounts according to the results of a Qubit dsDNA HS Assay (Thermo Fisher Scientific) quantification.
+The pool was sequenced in a fraction of a NovaSeq PE250 flow cell (Illumina) aiming for a total output of 2 gigabases.
+Any potential traces of adapter dimers were removed using Cutadapt v3.5 [@RN5538].
+Quality of the sequencing data was assessed with the software FastQC [@RN5539].
+
+# Processing of sequencing data and inference of ASVs
+
+Cutadapt v3.5 was used to trim off non-biological DNA sequences (primers, indices, and sequencing adapters).
+Then, amplicon reads were processed using the DADA2 pipeline [@RN5541] implemented in QIIME 2 (release 2023.7) [@RN5540].
+The first step in the DADA2 pipeline was to remove PCR primers and filter reads according to their quality.
+After checking the read quality profiles, forward reads were truncated at position 175, and reverse reads at position 131.
+The next step was to denoise and infer Amplicon Sequence Variants (ASVs).
+Error rates were learned from the dataset to denoise, using the parametric error model implemented in DADA2.
+Dereplication of the dataset (i.e.
+combination of all identical reads into unique reads) was carried out to reduce computational effort.
+These dereplicated forward and reverse reads were used to infer ASVs with the core sample inference algorithm [@RN5541].
+Subsequently, corresponding R1 and R2 reads were merged into pairs with a minimum overlap of 12 identical base pairs.
+The final step in the DADA2 pipeline was to reduce the impact of artefacts (chimaeras) in the dataset.
+The final output of the DADA2 pipeline was a table containing the count of reads of every observed ASV in each sample, as well as a table listing all ASVs and their corresponding representative sequences.
+
+# Taxonomic assignment
+
+Taxonomic assignment of each ASV was performed using a pre-trained classifier of the UNITE reference database for fungi [@RN5550] (release May 2021); and of the SILVA reference database for bacteria [@RN5542] (release 138.1 August 2020).
+To compare the representative sequences of the ASVs to the reference database, and compute the taxonomic assignment, the algorithm sklearn [@RN5543] implemented in QIIME 2 was applied, setting a minimum confidence level of 0.8.
+Several filters were applied to exclude: (1) singletons (i.e.
+ASVs containing only one member sequence in the whole data set); (2) ASVs occurring at a frequency below 0.01 % in each sample, to reduce the mistagging bias [@RN5544]; (3) from the bacterial dataset, non-prokaryotic ASVs such as eukaryotic sequences of plastid [@RN5545] and mitochondrial [@RN5546] origin; (4) those assigned only at domain (Bacteria) or kingdom (Fungi) level; (5) ASVs present in the negative controls that were more abundant in the negative controls than in the samples, indicating contamination [@RN5547].
+
+# Alpha rarefaction plots
+
+Alpha rarefaction plots showed that the sequencing depth to reach saturation was 30,000 reads for the fungi and 22,000 reads for the bacteria, above the sequencing depth of all plots for fungi (mean = 114,919 reads; min = 79,045; max = 157,110) and bacteria (mean = 113,341 reads; min = 84,094; max = 171,404).
+
+# References
+
+::: {#refs}
+:::

### 2025-03-17 results/supplementary/AppendixS2.Rmd
submission UFUG

---
title: "Appendix S2 Principal Component Analysis and correlation analysis of the soil physicochemical properties"
output:
  bookdown::pdf_document2:
    reference_docx: template.docx
    number_sections: false
    toc: false
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

```{r figS2a, out.width = "450px", echo = FALSE, fig.cap = "(A) Biplot of the first and second axes of a Principal Component Analysis of the soil physicochemical properties. Circles represent the sampling plots, coloured by habitat. Labeled arrows indicate the contribution of the soil variables to the axes."}
knitr::include_graphics("figS2a.png")
```

```{r figS2b, out.width = "450px", echo = FALSE, fig.cap = "(B) Biplot of the first and third axes of a Principal Component Analysis of the soil physicochemical properties. Circles represent the sampling plots, coloured by habitat. Labeled arrows indicate the contribution of the soil variables to the axes."}
knitr::include_graphics("figS2b.png")
```

```{r figS2c, out.width = "450px", echo = FALSE, fig.cap = "(C) Correlation heatmap of the soil variables."}
knitr::include_graphics("figS2c.png")
```

### 2025-03-17 src/1-plants.R

@@ -1,0 +1,26 @@
+library(tidyverse); library(indicspecies); library(labdsv)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+
+### Plant numbers
+
+plants %>% group_by(taxon)
+plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% pull(n) %>% mean
+plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% pull(n) %>% min
+plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% pull(n) %>% max
+plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% merge(header) %>% group_by(habitat) %>% summarise(n = round(mean(n), 0)) %>% arrange(n)
+plants %>% group_by(taxon) %>% tally() %>% arrange(-n)
+
+### Permanova
+
+header %>%
+  select(id, habitat) %>%
+  merge(plants) %>%
+  spread(taxon, cover, fill = 0) -> datos
+
+datos %>% select(-c(id, habitat)) -> df1
+
+RVAideMemoire::pairwise.perm.manova(dist(df1, "euclidian"), datos$habitat, nperm = 1000000, p.method = "holm") -> pairperma
+vegan::adonis2(df1 ~ habitat, data = datos, permutations = 100000) -> perma
+anova(betadisper(vegdist(df1, method = "euclidean"), df1b$Alliance)) -> betadis

### 2025-03-17 src/2-fungi.R

@@ -1,0 +1,167 @@
+library(tidyverse)
+
+### Clean fungi
+
+read.csv("data/fungi-original.csv", sep = "\t", 
+         #fileEncoding = "latin1", 
+         skip = 1) %>%
+  separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%
+  mutate(kingdom = gsub("k__", "", kingdom), 
+         phylum = gsub("p__", "", phylum), 
+         class = gsub("c__", "", class), 
+         order = gsub("o__", "", order), 
+         family = gsub("f__", "", family), 
+         genus = gsub("g__", "", genus), 
+         species = gsub("s__", "", species), 
+         species = gsub("_", " ", species)) %>%
+  rename(ASV = X.OTU.ID,
+         Xixon01 = X1.ITS,
+         Xixon02 = X2.ITS,
+         Xixon03 = X3.ITS,
+         Xixon04 = X4.ITS,
+         Xixon05 = X5.ITS,
+         Xixon06 = X6.ITS,
+         Xixon07 = X7.ITS,
+         Xixon08 = X8.ITS,
+         Xixon09 = X9.ITS,
+         Xixon10 = X10.ITS,
+         Xixon11 = X11.ITS,
+         Xixon12 = X12.ITS,
+         Xixon13 = X13.ITS,
+         Xixon14 = X14.ITS,
+         Xixon15 = X15.ITS,
+         Xixon16 = X16.ITS,
+         Xixon17 = X17.ITS,
+         Xixon18 = X18.ITS,
+         Xixon19 = X19.ITS,
+         Xixon20 = X20.ITS,
+         Xixon21 = X21.ITS,
+         Xixon22 = X22.ITS,
+         Xixon23 = X23FD.ITS,
+         Xixon24 = X24.ITS,
+         Xixon25 = X25.ITS,
+         Xixon26 = X26.ITS,
+         Xixon27 = X27.ITS,
+         Xixon28 = X28.ITS,
+         Xixon29 = X29.ITS,
+         Xixon30 = X30.ITS) -> fungi
+
+### Fungi numbers
+
+fungi %>% group_by(ASV)
+
+fungi %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV)
[…]

### 2025-03-17 src/3-bacteria.R

@@ -1,0 +1,179 @@
+library(tidyverse)
+
+read.csv("data/bacteria-original.csv", sep = "\t", 
+         #fileEncoding = "latin1", 
+         skip = 1) %>%
+  separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%
+  mutate(kingdom = gsub("d__", "", kingdom), 
+         phylum = gsub("p__", "", phylum), 
+         class = gsub("c__", "", class), 
+         order = gsub("o__", "", order), 
+         family = gsub("f__", "", family), 
+         genus = gsub("g__", "", genus), 
+         species = gsub("s__", "", species), 
+         species = gsub("_", " ", species)) %>%
+  rename(ASV = X.OTU.ID,
+         Xixon01 = X1.16S,
+         Xixon02 = X2.16S,
+         Xixon03 = X3.16S,
+         Xixon04 = X4.16S,
+         Xixon05 = X5.16S,
+         Xixon06 = X6.16S,
+         Xixon07 = X7.16S,
+         Xixon08 = X8.16S,
+         Xixon09 = X9.16S,
+         Xixon10 = X10.16S,
+         Xixon11 = X11.16S,
+         Xixon12 = X12.16S,
+         Xixon13 = X13.16S,
+         Xixon14 = X14.16S,
+         Xixon15 = X15.16S,
+         Xixon16 = X16.16S,
+         Xixon17 = X17.16S,
+         Xixon18 = X18.16S,
+         Xixon19 = X19.16S,
+         Xixon20 = X20.16S,
+         Xixon21 = X21.16S,
+         Xixon22 = X22.16S,
+         Xixon23 = X23FD.16S,
+         Xixon24 = X24.16S,
+         Xixon25 = X25.16S,
+         Xixon26 = X26.16S,
+         Xixon27 = X27.16S,
+         Xixon28 = X28.16S,
+         Xixon29 = X29.16S,
+         Xixon30 = X30.16S) -> bacteria
+
+### numbers
+
+bacteria %>% group_by(ASV)
+
+bacteria %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV)
[…]

### 2025-03-17 src/4-soils.R

@@ -1,0 +1,400 @@
+library(tidyverse); library(lubridate)
+Sys.setlocale("LC_TIME", "English")
+
+c("forestgreen",  
+  "gold", 
+  "#40E0D0",
+  "#A020F0",
+  "#5CACEE",
+  "#551A8B") -> colores
+
+### Reads and format GP5WSHELL csv export files
+
+read_gp5 <- function(x)
+{
+  require(tidyverse)
+  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
+    dplyr::rename(temperature = X.1.oC) %>%
+    mutate(time = as.POSIXct(Time, 
+                             format =" %d.%m.%Y %H:%M:%S", 
+                             tz = "UTC"),
+           logger = gsub("\\_.*", "" , x)) %>%
+    dplyr::select(logger, time, temperature)
+}
+
+read.table("data/temperature/Loggers.txt", sep = "\t", header = TRUE, fileEncoding = "latin1") %>%
+  mutate(Installed = as.POSIXct(Installed, 
+                                format ="%d/%m/%Y", 
+                                tz = "UTC")) %>%
+  rename(habitat = N.type) %>%
+  rename(logger = Logger) %>%
+  filter(plot %in% c("Xixon_24", # Bosque mixto en Porceyu con sotobosque de Laurus
+                     "Xixon_07", # Detrás del cementeriu de Xove
+                     "Xixon_25", # Parque del Lauredal
+                     "Xixon_08", # Parcela en La Calzada cerca de los feriantes
+                     "Xixon_15", # Vías del tren en La Calzada
+                     "Xixon_21")) -> # Estrella de Qatar
+  loggers
+
+## Read and bind all loggers, merge with logger header data
+
+wd <- getwd()
+setwd(paste(wd, "/data/temperature", sep = ""))
+do.call(rbind, lapply(list.files(pattern = "*.csv"), read_gp5)) %>% 
+  merge(loggers, by = "logger") %>%
+  na.omit %>% 
+  filter(time > (Installed + 60*60*24*7)) %>% # Remove records up to one week after installation
+  filter(time >= "2024-01-01 01:00:00 UTC") %>%
+  filter(time < "2025-01-01 01:00:00 UTC") %>%
+  select(ID, plot, habitat, time, temperature) %>%
+  arrange(habitat, time) ->
+  logs
+setwd(wd)
+head(logs, n = 25)
[…]

### 2025-03-17 src/5-biodiversity.R

@@ -1,0 +1,492 @@
+library(tidyverse); library(ggvenn); library(vegan)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### Plant species pools
+
+list(
+  `Forests (5 plots)` = 
+    header %>% 
+    filter(habitat == "forest") %>% 
+    select(id) %>% 
+    merge(plants) %>% 
+    pull(taxon) %>% 
+    unique, 
+  `Meadows (5 plots)` = 
+    header %>% 
+    filter(habitat == "meadow") %>% 
+    select(id) %>% 
+    merge(plants) %>% 
+    pull(taxon) %>% 
+    unique, 
+  `Urban habitats (20 plots)` = 
+    header %>% 
+    filter(habitat %in% c("park", "residential", "roadside", "industrial")) %>% 
+    select(id) %>% 
+    merge(plants) %>% 
+    pull(taxon) %>% 
+    unique) -> df1
+
+plants %>%
+  filter(taxon %in% df1$`Forests (5 plots)` & ! taxon %in% df1$`Meadows (5 plots)` & ! taxon %in% df1$`Urban habitats (20 plots)`) %>%
+  group_by(taxon) %>%
+  tally() %>%
+  arrange(-n)
+
+plants %>%
+  filter(! taxon %in% df1$`Forests (5 plots)` & taxon %in% df1$`Meadows (5 plots)` & ! taxon %in% df1$`Urban habitats (20 plots)`) %>%
+  group_by(taxon) %>%
+  tally() %>%
+  arrange(-n)
+
+plants %>%
+  filter(! taxon %in% df1$`Forests (5 plots)` & ! taxon %in% df1$`Meadows (5 plots)` & taxon %in% df1$`Urban habitats (20 plots)`) %>%
+  group_by(taxon) %>%
+  tally() %>%
+  arrange(-n)
+
+### Fungal species pools
+
+list(
+  `Forests (5 plots)` = 
+    header %>% 
+    filter(habitat == "forest") %>% 
+    select(id) %>% 
+    merge(fungi) %>% 
+    pull(ASV) %>% 
[…]

### 2025-03-17 src/6-novelty.R

@@ -1,0 +1,302 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 4) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus forest)
+
+sitescores %>%
+  filter(habitat %in% c("forest")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = forest - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
+
+### Novelty (versus meadow)
+
+sitescores %>%
+  filter(habitat %in% c("meadow")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = meadow - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
+
+### Compositional novelty for fungi
+
+fungi %>%
+  select(id, ASV, presence) %>%
+  spread(ASV, presence, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 4) -> 
[…]

### 2025-03-17 src/bacteria.R

@@ -1,179 +1,0 @@
-library(tidyverse)
-
-read.csv("data/bacteria-original.csv", sep = "\t", 
-         #fileEncoding = "latin1", 
-         skip = 1) %>%
-  separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%
-  mutate(kingdom = gsub("d__", "", kingdom), 
-         phylum = gsub("p__", "", phylum), 
-         class = gsub("c__", "", class), 
-         order = gsub("o__", "", order), 
-         family = gsub("f__", "", family), 
-         genus = gsub("g__", "", genus), 
-         species = gsub("s__", "", species), 
-         species = gsub("_", " ", species)) %>%
-  rename(ASV = X.OTU.ID,
-         Xixon01 = X1.16S,
-         Xixon02 = X2.16S,
-         Xixon03 = X3.16S,
-         Xixon04 = X4.16S,
-         Xixon05 = X5.16S,
-         Xixon06 = X6.16S,
-         Xixon07 = X7.16S,
-         Xixon08 = X8.16S,
-         Xixon09 = X9.16S,
-         Xixon10 = X10.16S,
-         Xixon11 = X11.16S,
-         Xixon12 = X12.16S,
-         Xixon13 = X13.16S,
-         Xixon14 = X14.16S,
-         Xixon15 = X15.16S,
-         Xixon16 = X16.16S,
-         Xixon17 = X17.16S,
-         Xixon18 = X18.16S,
-         Xixon19 = X19.16S,
-         Xixon20 = X20.16S,
-         Xixon21 = X21.16S,
-         Xixon22 = X22.16S,
-         Xixon23 = X23FD.16S,
-         Xixon24 = X24.16S,
-         Xixon25 = X25.16S,
-         Xixon26 = X26.16S,
-         Xixon27 = X27.16S,
-         Xixon28 = X28.16S,
-         Xixon29 = X29.16S,
-         Xixon30 = X30.16S) -> bacteria
-
-### numbers
-
-bacteria %>% group_by(ASV)
-
-bacteria %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV)
[…]

### 2025-03-17 src/fig2-soils.R

@@ -1,321 +1,0 @@
-library(tidyverse); library(lubridate)
-Sys.setlocale("LC_TIME", "English")
-
-c("forestgreen",  
-  "gold", 
-  "#40E0D0",
-  "#A020F0",
-  "#5CACEE",
-  "#551A8B") -> colores
-
-### Reads and format GP5WSHELL csv export files
-
-read_gp5 <- function(x)
-{
-  require(tidyverse)
-  read.csv(x, skip = 2, header = TRUE) %>% # Rows to skip depend on GP5WShell version!!
-    dplyr::rename(temperature = X.1.oC) %>%
-    mutate(time = as.POSIXct(Time, 
-                             format =" %d.%m.%Y %H:%M:%S", 
-                             tz = "UTC"),
-           logger = gsub("\\_.*", "" , x)) %>%
-    dplyr::select(logger, time, temperature)
-}
-
-read.table("data/temperature/Loggers.txt", sep = "\t", header = TRUE, fileEncoding = "latin1") %>%
-  mutate(Installed = as.POSIXct(Installed, 
-                                format ="%d/%m/%Y", 
-                                tz = "UTC")) %>%
-  rename(habitat = N.type) %>%
-  rename(logger = Logger) %>%
-  filter(plot %in% c("Xixon_24", # Bosque mixto en Porceyu con sotobosque de Laurus
-                     "Xixon_07", # Detrás del cementeriu de Xove
-                     "Xixon_25", # Parque del Lauredal
-                     "Xixon_08", # Parcela en La Calzada cerca de los feriantes
-                     "Xixon_15", # Vías del tren en La Calzada
-                     "Xixon_21")) -> # Estrella de Qatar
-  loggers
-
-## Read and bind all loggers, merge with logger header data
-
-wd <- getwd()
-setwd(paste(wd, "/data/temperature", sep = ""))
-do.call(rbind, lapply(list.files(pattern = "*.csv"), read_gp5)) %>% 
-  merge(loggers, by = "logger") %>%
-  na.omit %>% 
-  filter(time > (Installed + 60*60*24*7)) %>% # Remove records up to one week after installation
-  select(ID, plot, habitat, time, temperature) ->
-  logs
-setwd(wd)
-
-logs %>%
-  mutate(habitat = fct_relevel(habitat, "N1", "N2", "N3", "N4B", "N4A", "N4C")) %>%
-  mutate(habitat = fct_recode(habitat, "Forests" = "N1", "Meadows" = "N2",
[…]

### 2025-03-17 src/fig3-diversity.R

@@ -1,491 +1,0 @@
-library(tidyverse); library(ggvenn)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-### Plant species pools
-
-list(
-  Forests = 
-    header %>% 
-    filter(habitat == "forest") %>% 
-    select(id) %>% 
-    merge(plants) %>% 
-    pull(taxon) %>% 
-    unique, 
-  Meadows = 
-    header %>% 
-    filter(habitat == "meadow") %>% 
-    select(id) %>% 
-    merge(plants) %>% 
-    pull(taxon) %>% 
-    unique, 
-  `Urban habitats` = 
-    header %>% 
-    filter(habitat %in% c("park", "residential", "roadside", "industrial")) %>% 
-    select(id) %>% 
-    merge(plants) %>% 
-    pull(taxon) %>% 
-    unique) -> df1
-
-plants %>%
-  filter(taxon %in% df1$Forests & ! taxon %in% df1$Meadows & ! taxon %in% df1$`Urban habitats`) %>%
-  group_by(taxon) %>%
-  tally() %>%
-  arrange(-n)
-
-plants %>%
-  filter(! taxon %in% df1$Forests & taxon %in% df1$Meadows & ! taxon %in% df1$`Urban habitats`) %>%
-  group_by(taxon) %>%
-  tally() %>%
-  arrange(-n)
-
-plants %>%
-  filter(! taxon %in% df1$Forests & ! taxon %in% df1$Meadows & taxon %in% df1$`Urban habitats`) %>%
-  group_by(taxon) %>%
-  tally() %>%
-  arrange(-n)
-
-### Fungal species pools
-
-list(
-  Forests = 
-    header %>% 
-    filter(habitat == "forest") %>% 
-    select(id) %>% 
-    merge(fungi) %>% 
-    pull(ASV) %>% 
-    unique, 
-  Meadows = 
-    header %>% 
-    filter(habitat == "meadow") %>% 
-    select(id) %>% 
-    merge(fungi) %>% 
-    pull(ASV) %>% 
-    unique, 
-  `Urban habitats` = 
-    header %>% 
[…]

### 2025-03-17 src/fungi.R

@@ -1,167 +1,0 @@
-library(tidyverse)
-
-### Clean fungi
-
-read.csv("data/fungi-original.csv", sep = "\t", 
-         #fileEncoding = "latin1", 
-         skip = 1) %>%
-  separate(taxonomy, into = c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = "; ") %>%
-  mutate(kingdom = gsub("k__", "", kingdom), 
-         phylum = gsub("p__", "", phylum), 
-         class = gsub("c__", "", class), 
-         order = gsub("o__", "", order), 
-         family = gsub("f__", "", family), 
-         genus = gsub("g__", "", genus), 
-         species = gsub("s__", "", species), 
-         species = gsub("_", " ", species)) %>%
-  rename(ASV = X.OTU.ID,
-         Xixon01 = X1.ITS,
-         Xixon02 = X2.ITS,
-         Xixon03 = X3.ITS,
-         Xixon04 = X4.ITS,
-         Xixon05 = X5.ITS,
-         Xixon06 = X6.ITS,
-         Xixon07 = X7.ITS,
-         Xixon08 = X8.ITS,
-         Xixon09 = X9.ITS,
-         Xixon10 = X10.ITS,
-         Xixon11 = X11.ITS,
-         Xixon12 = X12.ITS,
-         Xixon13 = X13.ITS,
-         Xixon14 = X14.ITS,
-         Xixon15 = X15.ITS,
-         Xixon16 = X16.ITS,
-         Xixon17 = X17.ITS,
-         Xixon18 = X18.ITS,
-         Xixon19 = X19.ITS,
-         Xixon20 = X20.ITS,
-         Xixon21 = X21.ITS,
-         Xixon22 = X22.ITS,
-         Xixon23 = X23FD.ITS,
-         Xixon24 = X24.ITS,
-         Xixon25 = X25.ITS,
-         Xixon26 = X26.ITS,
-         Xixon27 = X27.ITS,
-         Xixon28 = X28.ITS,
-         Xixon29 = X29.ITS,
-         Xixon30 = X30.ITS) -> fungi
-
-### Fungi numbers
-
-fungi %>% group_by(ASV)
-
-fungi %>% filter(species != "unidentified" & !is.na(species)) %>% group_by(ASV)
[…]

### 2025-03-17 src/novelty.R

@@ -1,294 +1,0 @@
-library(tidyverse)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-### Compositional novelty for plants
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(plants) %>%
-  select(id, taxon, cover) %>%
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 4) -> 
-  nmds # Ordination output
-
-### Site scores
-
-vegan::scores(nmds, "sites") %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  gather(Trait, Value, - id) %>%
-  merge(select(header, id, habitat)) -> sitescores
-
-### Novelty (versus forest)
-
-sitescores %>%
-  filter(habitat %in% c("forest")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = forest - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
-
-### Novelty (versus meadow)
-
-sitescores %>%
-  filter(habitat %in% c("meadow")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = meadow - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
-
-### Compositional novelty for fungi
-
-fungi %>%
-  select(id, ASV, presence) %>%
-  spread(ASV, presence, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 4) -> 
[…]

### 2025-03-17 src/plants.R

@@ -1,26 +1,0 @@
-library(tidyverse); library(indicspecies); library(labdsv)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-
-### Plant numbers
-
-plants %>% group_by(taxon)
-plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% pull(n) %>% mean
-plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% pull(n) %>% min
-plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% pull(n) %>% max
-plants %>% group_by(id) %>% summarise(n = length(taxon)) %>% merge(header) %>% group_by(habitat) %>% summarise(n = round(mean(n), 0)) %>% arrange(n)
-plants %>% group_by(taxon) %>% tally() %>% arrange(-n)
-
-### Permanova
-
-header %>%
-  select(id, habitat) %>%
-  merge(plants) %>%
-  spread(taxon, cover, fill = 0) -> datos
-
-datos %>% select(-c(id, habitat)) -> df1
-
-RVAideMemoire::pairwise.perm.manova(dist(df1, "euclidian"), datos$habitat, nperm = 1000000, p.method = "holm") -> pairperma
-vegan::adonis2(df1 ~ habitat, data = datos, permutations = 100000) -> perma
-anova(betadisper(vegdist(df1, method = "euclidean"), df1b$Alliance)) -> betadis

### 2025-07-07 doc/R0/title-page.Rmd
LANDUP R1

---
title: "Compositional novelty of plant, fungal and bacterial communities across urban habitats"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: urban-forestry-and-urban-greening.csl
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

Running title: Compositional novelty across urban habitats

Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^2^ (<https://orcid.org/0000-0002-0915-0916>), Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)

^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain

Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain. Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}. Telephone: +34985104781.

# Author contributions

Eduardo Fernández-Pascual: Conceptualization; Methodology; Investigation; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing. Zuzana Ferencova: Conceptualization; Methodology; Investigation; Data Curation; Writing – Review & Editing. Víctor González-García: Investigation; Visualization; Writing – Review & Editing; Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.

# Acknowledgements

María García Fernández, Sara Gutiérrez Cruz and Abelardo Sigüenza Solís helped with the field surveys. Thanks to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis and to INEA Laboratory (Valladolid, Spain) for soil physicochemical analysis. We are thankful to Fundación Biodiversidad for their support.

# Declaration of interest statement

We have nothing to declare.

# Funding

European Union - NextGenerationEU within the framework of the Recovery, Transformation and Resilience Plan (PRTR) of Ministry for the Ecological Transition and the Demographic Challenge (MITECO) of the Government of Spain —Project: Gijón Ecoresiliente (PRTR 0204103/010000). Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico [SV-23-GIJON-JBA]. Grant "Laboratorio de Vegetación y Biodiversidad" (IDE/2024/000720, Principality of Asturias-Sekuens-EU-FEDER). Víctor González-García was supported by the Government of Asturias (Programa “Severo Ochoa” PA-22-BP21-098).

# Data availability statement

Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at Zenodo: [DOI WILL BE INSERTED HERE].

### 2025-07-07 src/5-biodiversity.R

@@ -5,6 +5,7 @@ read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
 read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
 read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
 read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
 
 ### Plant species pools
 
@@ -229,6 +230,7 @@ header %>%
   select(habitat, id) %>%
   merge(plants) %>%
   select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
   spread(taxon, cover, fill = 0) %>%
   column_to_rownames(var = "id") -> matrix 
 
@@ -239,10 +241,20 @@ matrix %>%
 ### Env fit
 
 soils %>% select(Pb, OM, Mg) -> soilvar
+pca %>% select(Dim.1:Dim.3) -> soilvar
 
 vegan::envfit(nmds, soilvar, permutations = 100000, strata = NULL, 
        choices=c(1,2),  display = "sites", w  = weights(nmds, display),
-       na.rm = FALSE)
+       na.rm = FALSE) -> ef
+
+ef
+ef$vectors$arrows %>%
+  data.frame %>%
+  rownames_to_column(var = "Variable") %>%
+  filter(Variable == "Dim.1") %>%
+  mutate(Variable = as.factor(Variable)) %>%
+  mutate(Variable = fct_recode(Variable, "Pb" = "Dim.1")) %>%
+  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) -> pcaVars
 
 ### Site scores
 
@@ -274,6 +286,8 @@ ggplot(sitescores, aes(x = Dim.1, y = Dim.2)) +
   geom_hline(yintercept = 0, linetype = "dashed") +
   geom_vline(xintercept = 0, linetype = "dashed") +
   geom_point(aes(fill = habitat), color = "black", size = 3.5, shape = 21) +
+  geom_segment(data = pcaVars, aes(x = 0, y = 0, xend = Dim.1, yend = Dim.2)) +
+  geom_label(data = pcaVars, aes(x = Dim.1, y = Dim.2, label = Variable),  show.legend = FALSE, size = 4) +
[…]

### 2025-07-07 src/6-novelty-R1-PCA.R

@@ -1,0 +1,303 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") %>% 
+  decostand(method = "hellinger") -> df1
+
+df1 %>%
+  FactoMineR::PCA(graph = TRUE, scale.unit = FALSE) -> pca1
+
+### Site scores
+
+pca1$ind$coord %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus forest)
+
+sitescores %>%
+  filter(habitat %in% c("forest")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = forest - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
+
+### Novelty (versus meadow)
+
+sitescores %>%
+  filter(habitat %in% c("meadow")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = meadow - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
+
+### Compositional novelty for fungi
+
+fungi %>%
+  select(id, ASV, presence) %>%
+  spread(ASV, presence, fill = 0) %>%
+  column_to_rownames(var = "id") %>% 
+  decostand(method = "hellinger") -> df1
+
+df1 %>%
[…]

### 2025-07-07 src/6-novelty-R1.R

@@ -1,0 +1,306 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 4) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus forest)
+
+sitescores %>%
+  filter(habitat %in% c("forest")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = forest - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
+
+### Novelty (versus meadow)
+
+sitescores %>%
+  filter(habitat %in% c("meadow")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = meadow - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
+
+### Compositional novelty for fungi
+
+fungi %>%
+  select(id, ASV, presence) %>%
+  spread(ASV, presence, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
[…]

### 2025-07-07 src/7-species-area.R

@@ -1,0 +1,164 @@
+library(tidyverse); library(ggvenn); library(vegan)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
+
+
+header %>%
+  select(habitat, id) %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+library(BiodiversityR)
+
+Condit.spec <- matrix
+
+Condit.2 <- header
+site.rows <- as.character(Condit.2[, 1])
+Condit.env <- Condit.2[, 2:7]
+rownames(Condit.env) <- site.rows
+Condit.env$Age.cat <- factor(Condit.env$habitat)
+Condit.env$Area <- rep(16, nrow(Condit.env))
+
+check.datasets(Condit.spec, Condit.env)
+
+Condit.env <- Condit.env[complete.cases(Condit.env), ]
+Condit.spec <- same.sites(Condit.spec, Condit.env)
+check.datasets(Condit.spec, Condit.env)
+
+
+Accum.2 <- accumcomp(Condit.spec, y=Condit.env, factor='Age.cat', scale='Area',
+                     method='exact', conditioned=FALSE, plotit=FALSE)
+Accum.2
+
+accum.long2 <- accumcomp.long(Accum.2, ci=NA, label.freq=1)
+head(accum.long2)
+
+plotgg2 <- ggplot(data=accum.long2, aes(x = Sites, y = Richness, ymax = UPR, ymin = LWR)) + 
+  scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
+  scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
+  geom_line(aes(colour=Grouping), size=2) +
+  geom_point(data=subset(accum.long2, labelit==TRUE), 
[…]

### 2025-08-22 doc/R1/response.Rmd
LANDUP R2

---
title: "Compositional novelty of plant, fungal and bacterial communities across urban habitats"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: landscape-and-urban-planning.csl
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

## Editor

> I note that Reviewer 1 and Reviewer 2 have conflicting assessments on your analyses. I agree with Reviewer 1 in that there are issues to address in your revision, and R1 provides detailed ways to move forward. In addition to addressing Reviewer comments, please also include 1-2 sentences in your Abstract that provides the context of the study. These first sentences usually summarize the Introduction. Starting the abstract with the hypotheses does not provide the background of the study.

Thank you for reviewing our manuscript and for providing directions to improve it. We have prepared a new version based on the suggestions of both reviewers. Among other things, we have conducted additional tests to assess the issues raised by reviewer 1. We note that, overall, the additional tests did not change the output and the interpretation of the results. We have also revised the abstract following your advice.

## Reviewer 1

> This paper offers an intriguing study on assessing community novelty for multiple taxa in urban habitats. The authors use ordination methods to evaluate the differences between communities in urban habitats and those in natural and semi-natural settings, aiming to assess the novelty of urban communities. Their findings indicate that novelty is highest for plant communities and lowest for soil bacteria, with industrial vacant lots exhibiting the highest novelty among urban habitats.

Thank you for your positive evaluation and your suggestions for improving the manuscript.

> The NMDS (Non-Metric Multidimensional Scaling) method should be carefully considered for ordination analysis in your case. Since NMDS employs an iterative algorithm to seek optimal solutions, the results may vary across different computational runs, leading to inconsistencies in ordination outcomes. Therefore, it is advisable to consider alternative ordination methods based on eigenvalue decomposition.

The referee is right in his description of NMDS. Following his advice, we tested the alternative of using a method based on eigenvalue decomposition: Principal Component Analysis (PCA) with Hellinger transformation. Using one method or the other did not change the outcome and interpretation of the analysis (neither the pattern of novelty across taxa, urban habitats and reference habitats; nor the statistical significance of the LMM model). Therefore, we decided to use NMDS for the final presentation in the manuscript, since NMDS makes less assumptions about data and can be applied to a wider variety of datasets. We explain this in the methods section (lines 197-216). We also note that, although in the manuscript and figures we are reporting the output of a specific set of NMDS runs, during analysis we repeated the runs many times and the results never changed significantly; this supports that the results are robust despite the iterative nature of NMDS.

> Additionally, the community data used for different biological groups are incompatible: plant data are based on abundance, while fungal and bacterial data use presence/absence metrics, resulting in a lack of comparability across taxa.

The referee is right, and thus we have changed the analysis by transforming plant cover into presence/absence, so the three taxa can be compared. We note that the output and interpretation of results have not been altered by this change.

> Furthermore, critical details of the NMDS analysis—such as the number of axes used, the specific distance metric applied (e.g., Bray-Curtis), and the stress value (a key indicator of model fit)—were not adequately documented in the text.

We agree, and we have expanded the methods to provide a full report of NMDS including distance metric, number of axes and stress values (lines 208-216).

> The methodology mentions the use of a Generalized Mixed Model, but the corresponding results are not reported. If the authors indeed used this method, they need to provide a more detailed description of the analysis process, such as the total sample size, the distribution family used, and whether model selection was conducted.

We thank the reviewer for spotting a mistake in our text. We did not use a GLM, but a Linear Mixed Model (i.e. Gaussian family). We did this after checking that LMM assumptions were met, as described in the methods section. We did not conduct model selection. We report model results (F and p values) in the text of the results section (lines 335-350). We have included a sentence about sample size in the methods (lines 221-223).

> The paper employs PCA to reduce the dimensionality of soil physicochemical variables, but subsequently selects only three factors with the highest loadings on the first three axes for further analysis. Why not use the first three axes directly for subsequent analysis? Additionally, the explanatory rates of the first three PCA axes are not described.

Following the reviewer's advice, we now use the three axes for subsequent analysis. The only change this has caused in the results is that the first axis (corresponding to soil heavy metals) now has a significant fit with the NMDS ordination of the plant communities. We are not certain of what is meant by "explanatory rates of the first three PCA axes". We describe the contribution of each axis to the variability in the results section (lines 235, 241, 247).

> The limited sample size of 30 plots undermines the robustness and reliability of the study's findings.

We believe that model significance indicates that the sample size was robust enough to provide statistical power to test the hypotheses of this study.

> Furthermore, while natural forests were surveyed in the surrounding areas, the urban habitat sampling primarily concentrated on grassland vegetation, neglecting forest vegetation within urban contexts.

In the study city, there is no urban forest vegetation, except for natural forest remnants in the city's fringes. The level of management does not allow for succession of urban lots to proceed to any stage dominated by trees.

> The Discussion section lacks depth, merely reiterating results without offering substantive interpretation. Additionally, the section titled "Widespread Soil Contamination Across Urban Habitats" strays from the central focus of the study.

We have revised the discussion and removed the section on soil contamination, following the reviewer's suggestion. In the discussion, there is a brief recap of the results for the benefit of the reader, but this is followed by their interpretation in light of the original hypotheses and the literature. Specifically, we believe that the discussion provides substantive interpretation regarding how our results should translate into management actions for each habitat type (see for example lines 414-419).

> Several minor comments are detailed in the annotated manuscript.

We have addressed the minor comments included in the annotated PDF.

## Reviewer 2

> The paper describes the outcome of research conducted to characterize and better understand the diversity and compositional novelty of urban habitats with respect to their plant, fungal and bacterial communities. Although many studies have explored plant diversity associated with different habitats, this study is original in its investigation of the composition, diversity and relative novelty of not only plant communities, but also the soil microbiome in a range of urban habitats. The composition of these urban habitats is compared with remnant forest and meadow sites in the study region.  Although there are relatively few investigations of bacterial and fungal community composition, the authors do cite other relevant work in the field. The authors also noted the motivation and significance of the work to urban restoration efforts in the introduction and returned to discuss their results in the light of restoration and rewilding in the discussion section. Most of the methods used in data acquisition and statistical analyses are appropriate to the investigation and results are reported clearly in the text and in the figures.  The one exception is the lack of description of methods used to determine “Naturalness”, as noted below. The manuscript is quite well written.  I found very few errors and had only a few questions after reading the article, as noted below.

Thank you for your positive consideration of our work.

> Do you think that you were able to sample a sufficiently large number of replicates for each habitat type that you are approaching an understanding of the actual alpha diversity of these habitats (i.e., where are you on the species-area curve with respect to sampling the urban habitats you have chosen to study?  What might you be able to discern from your data about the potential number of species that might be found if additional sampling had been conducted?

We have checked species-area curves (see below) and our sampling is not sufficient to cover the whole species diversity of the studied habitats. The are indeed additional data to support this: an iNaturalist project on the flora of the city currently has > 2,500 research-grade observations of c. 460 vascular plant species; our survey detected 216 species. Nonetheless, we note that it was not our goal to survey the entire biodiversity of the habitats, but to compare them. With this goal in mind, we have the same number of plots per habitat, of the same size.

```{r fig1, echo = FALSE, fig.cap = ""}
knitr::include_graphics("../results/figures/plotgg2.png")
```

```{r fig2, echo = FALSE, fig.cap = ""}
knitr::include_graphics("../results/figures/plotgg2-fungi.png")
```

```{r fig3, echo = FALSE, fig.cap = ""}
knitr::include_graphics("../results/figures/plotgg2-bacteria.png")
```

> The methods used to determine the level of “Naturalness” of each habitat are not described. In Figure 1, I am confused by the assignment of Industrial habitats to “high naturalness” as well as “high novelty”. While these habitats were species rich, they were rich in exotic and invasive species as well as native species. This seems to me to be a state of lower naturalness, despite the high diversity. Please clarify the methods used to estimate the “degree of ecosystem self-organization” , at the very  least, cite other works whose methods you are using to determine “naturalness”.

We are following the terminology by Kowarik (see for example https://doi.org/10.1016/j.ufug.2017.05.017). According to this terminology, a community can be rich in introduced species ("novel") but also "natural" in the sense that the species in the assemblage have arrived with little human intervention. Of course, the species have arrived to the regional species pool by human means, but their occurrence in the local community is spontaneous. For example, the community in a park lawn is less "natural" than the community in an abandoned industrial site, because the community in the park lawn is periodically affected by seeding, watering, fertilization, etc.; while the community in the industrial site is left to follow ecological succession for several years (until the site is "cleared" out). We describe this and provide the reference in the introduction, paragraph 3 (lines 61-71). We note however that this is a conceptualization which we use for an *a priori* characterization of the habitats, rather than a methodology.

> Page 18 line 6 – Use “As” rather than “Like” to start the sentence.

Changed as indicated.

> I was taught that sentences should not begin with an acronym, such as sentences beginning with “NMDS…”  or “DNA …” .  However, I realize that in recent times, fewer journals seem to be adhering to this grammatical rule. I will defer to the editor to determine whether the authors should revise their sentences to avoid having to begin sentences with an acronym.

We have revised the text to avoid starting sentences with acronyms.

> Figure 4 A.  The color for the Residential box in the box plot seems to be incorrect, at least it does not match the color scheme in the other figures- shouldn’t it be teal blue and not purple?

The same color scheme is followed in all figures, however, in the case of Figure 4A, the residential box is so narrow that the black line which marks the median value hides the blue color of the box, and gives the impression of being purple when rendered at the final resolution.

> Line 5-6 on page 14-  To improve clarity, revise this sentence to : “As previously indicated by PCA, organic matter was the only soil property with a clear pattern of variation by habitat: a decrease from pre-urban to urban habitats (Figure 2B).

Changed as suggested.

### 2025-08-22 doc/response.Rmd
LANDUP R2

---
title: "Compositional novelty of plant, fungal and bacterial communities across urban habitats"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: landscape-and-urban-planning.csl
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

# Editor

> I have completed my evaluation of your manuscript. The reviewers recommend reconsideration of your manuscript following minor revision and modification. I invite you to resubmit your manuscript after addressing the comments below. Please resubmit your revised manuscript by Sep 12, 2025. Please address R1's comment about the suitability of using a PCA rather than PCoA, in addition to the other comments by both reviewers. When revising your manuscript, please consider all issues mentioned in the reviewers' comments carefully: please outline every change made in response to their comments and provide suitable rebuttals for any comments not addressed. Please note that your revised submission may need to be re-reviewed.

Thank you for your positive decision on our manuscript. We have addressed all minor changes, including testing the alternative of using PCoA for the ordination. Please find below our detailed response to each comment. We also uploaded a track-changes version of the manuscript.

# Reviewer 1

> I am uncertain whether PCA with Hellinger transformation is appropriate for community data, particularly presence/absence data. Principal Coordinates Analysis (PCoA) may be more suitable in this context, as it is an eigenanalysis technique (similar to PCA) and also a distance-based ordination method (like NMDS).

As suggested, we have repeated the analysis using PCoA. The alternative method did not affect the final output of the analyses in a significant way. Therefore, we have decided to keep our original choice of NMDS, which to our knowledge is the reference method in the field. We have changed the methods section to indicate that we tried PCoA in addition to PCA and NMDS.

> Regarding lines 254-255, although the author mentions referencing each axis by its most explanatory variable (e.g., PC1 = Pb, PC2 = organic matter (OM), PC3 = Mg), this can still lead to confusion for readers, as seen in lines 280-283. It would be clearer to revise this statement to: "Environmental fitting of the soil property axes PC1, PC2, and PC3 onto the ordination indicated that only PC1 (mainly explained by Pb) had a significant correlation (R² = 0.29, p = 0.008) with the fungal community composition."

We have revised the sentence as suggested, and also revised similar sentences throughout the text.

> In the Methods section, it is stated that "When using NMDS for calculating novelty, we fitted the analysis to 4 dimensions", while "when using NMDS for assessing variation in taxonomic composition, we fitted the analysis to 2 dimensions". However, Line 335-337 show novelty index calculated from the ordinations in Fig. 3D-F that based on two dimensions NMDS.

It is as explained in the Methods: we used 4 dimensions to calculate the novelty index, and 2 dimensions for biplot visualization of the variation in taxonomic composition. We agree with the reviewer that the sentence "(calculated from the ordinations in **Fig. 3D-F**)" in lines 335-337 could cause confusion. Therefore, we have removed this sentence.

> Lastly, it would be beneficial to include detailed results of the Linear Mixed Models (LMMs) in the Supplementary Material.

We have created a new Supplementary material 3 with full details of the LMM.

# Reviewer 2

> I appreciate the care the authors have taken to thoroughly address my comments, and specifically, clarifying how they define the concept of natural ness in this paper. A couple of edits would improve the grammar of the paper: Line 207-210 Replace "less" with "fewer" so the sentence reads: "Choosing NMDS or PCA did not affect in any way the final output and interpretation of the analyses, and therefore we decided to use NMDS for the final presentation of results, since it makes fewer assumptions about data and is therefore a more flexible technique that can be applied to a wider variety of datasets. " Line 232 add "the to read: "... the forest soil..." Line 254 Replace "like" with "as" Line 260-261 Revise to read, "For simplicity, we will refer to each axis by the name of the explanatory variable most strongly associated with that axis..." Once these few edits are made, I recommend acceptance of the article, and have no further comments or edits to suggest.

Once again, we would like to thank reviewer 2 for his/her constructive comments on our manuscript, and for the suggestions to improve the writing. We have implemented all suggestions.

### 2025-08-22 doc/R1/manuscript.Rmd
LANDUP R2

--- 2025-03-17 doc/manuscript.Rmd
@@
 bibliography: refs.bib
-csl: urban-forestry-and-urban-greening.csl
+csl: landscape-and-urban-planning.csl
 link-citations: yes
@@
 
+Understanding urban community novelty can help to predict the rewilding potential of vacant urban spaces, facilitating their integration into biodiverse cities.
 We tested the hypothesis that cities are composed of different degrees of ecological novelty by studying four urban habitats with differing degrees of management and human legacy: park lawns, roadsides, residential vacant lots, and industrial vacant lots.
-We focused on community compositional novelty, by comparing the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forest remnants and hay meadows.
+We focused on community compositional novelty, by comparing the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forests and meadows.
 We used a compositional novelty index based on multidimensional ordination, which is straightforward to calculate and only requires species co-occurrence data for urban and reference habitats.
-As expected, (1) plant communities displayed the highest novelty and soil bacteria the lowest; (2) urban communities were markedly different from forest communities and relatively more similar to meadow communities; and (3) the degree of compositional novelty was highest in industrial vacant lots.
-On the contrary, managed park lawns, which we had expected to be highly novel, were relatively close to hay meadows.
-The lowest novelty among urban habitats was recorded in residential vacant lots, which had biological communities that more closely resembled those of pre-urban habitats.
-Our results highlight the effect of habitat type as a major driver of urban community composition and compositional novelty.
-These results suggest that city biodiversity can be enhanced by an integrative approach to the urban landscape that favors habitat heterogeneity by passive rewilding of managed park lawns, non-intervention on residential vacant lots, direct restoration of industrial vacant lots, and conservation of natural and agricultural habitat remnants as sources of native species.
+As expected, (1) plants displayed the highest novelty and bacteria the lowest; (2) urban communities were markedly different from forests and relatively more similar to meadows; and (3) compositional novelty was highest in industrial lots.
+Managed park lawns, which we had expected to be highly novel, were relatively close to hay meadows.
+The lowest novelty was recorded in residential vacant lots, which had biological communities that more closely resembled those of pre-urban habitats.
+Our results highlight the effect of habitat type as a major driver of urban community composition and novelty.
+This suggests that city biodiversity can be enhanced by an integrative approach to the urban landscape that favors habitat heterogeneity by passive rewilding of park lawns, non-intervention on residential vacant lots, direct restoration of industrial lots, and conservation of natural and agricultural habitat remnants as sources of native species.
 
@@
 The city belongs to the Cantabrian Mixed Forests ecoregion [@RN5493; @RN5600], at the border between the temperate and Mediterranean climatic zones of Europe.
-Climate is temperate oceanic with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm; Agencia Estatal de Meteorología).
+Climate is temperate oceanic with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1062 mm; 1981-2010 data provided by Agencia Estatal de Meteorología).
 Most of the city is built on alluvial and estuarine deposits, sand or limestone bedrock.
@@
 
-To survey urban biodiversity across habitats, we defined six habitat types (**Figure 1A**) representing different theoretical states of naturalness and novelty [@RN5445]: (1) forest patches \< 10 km from the city center, i.e., remnants of pre-urban natural habitats; (2) late mown hay meadows, i.e., remnants of pre-urban agricultural habitats; (3) park lawns, i.e., green spaces created and maintained by humans; (4) roadsides, i.e., relatively unmanaged strips of land alongside roads and train tracks; (5) residential vacant lots, i.e., unmanaged land which has suffered some degree of soil physical disturbance and is waiting to be urbanized; and (6) industrial vacant lots, i.e., unmanaged land with an intense legacy of soil disturbance and contamination.
+To survey urban biodiversity across habitats, we defined six habitat types (**Figure 1A**) representing different theoretical states of naturalness and novelty [@RN5445]: (1) forest patches \< 10 km from the city center, i.e., remnants of pre-urban natural habitats; (2) late mown hay meadows, i.e., remnants of pre-urban agricultural habitats; (3) park lawns, i.e., green spaces created and maintained by humans; (4) roadsides, i.e., relatively unmanaged strips of land alongside roads and train tracks; (5) residential vacant lots, i.e., currently unmanaged land which has suffered some degree of soil physical disturbance and is waiting to be urbanized; and (6) industrial vacant lots, i.e., currently unmanaged land with an intense legacy of soil disturbance and contamination.
 
@@
 During June 2023, we surveyed the city to establish five sampling plots (4 x 4 m) per each of the six habitat types (n = 30 plots).
+The same plot size was maintained in the forest patches for comparability.
 The plots were distributed across the city (**Figure 1B**) and represent the variation of vegetation types found within each habitat type.
@@
 The frozen tubes were sent to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis.
-DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi, and 16S for bacteria; sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size \< 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, inference of Amplicon Sequence Variants (ASVs), taxonomic assignment, number of reads per taxon and generation of rarefaction curves (see **Appendix S1** for extended DNA metabarcoding methodology).
+The DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi, and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size \< 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, inference of Amplicon Sequence Variants (ASVs), taxonomic assignment, number of reads per taxon and generation of rarefaction curves (see **Appendix S1** for extended DNA metabarcoding methodology).
 
@@
 To assess the variability in soil physicochemical properties, we conducted a Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] and a correlation analysis to detect auto-correlated soil variables.
-Multivariate analyses of community composition were performed in *vegan* [@RN5386]: we used PERMANOVA to test differences between habitat types in their soil variables and their composition of plant species, fungal ASVs and bacterial ASVs;  Non-metric Multi-Dimensional Scaling (NMDS) ordination to assess variation in taxonomic composition of plants (abundance values of species), fungi (presence/absence of ASVs) and bacteria (presence/absence of ASVs); and environmental fitting to test the correlation between community ordination axes and soil physicochemical properties.
+Multivariate analyses of community composition were performed in *vegan* [@RN5386]: we used PERMANOVA to test differences between habitat types in their soil variables and their composition of plant species, fungal ASVs and bacterial ASVs; Non-metric Multi-Dimensional Scaling (NMDS, see details below) ordination to assess variation in taxonomic composition of plants (for comparison with fungal and bacterial communities, we used presence/absence of plant species, instead of cover), fungi (presence/absence of ASVs) and bacteria (presence/absence of ASVs); and environmental fitting to test the correlation between community ordination axes and soil physicochemical properties.
 
@@
 We calculated compositional novelty separately for plant, fungal and bacterial communities.
-We tested the effect of taxonomic group, pre-urban reference habitat (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables on compositional novelty, by fitting a Generalized Mixed Model with the package *nlme* [@RN2333].
-The model included sampling plot as a random factor.
+Because NMDS employs an iterative algorithm to seek optimal solutions, the results may vary across different computational runs, leading to inconsistencies in ordination outcomes.
+Therefore, we also considered an alternative ordination method based on eigenvalue decomposition (Principal Component Analysis PCA with Hellinger transformation).
+Choosing NMDS or PCA did not affect in any way the final output and interpretation of the analyses, and therefore we decided to use NMDS for the final presentation of results, since it makes less assumptions about data and is therefore a more flexible technique that can be applied to a wider variety of datasets.
+We computed NMDS using the Bray-Curtis distance.
+When using NMDS for assessing variation in taxonomic composition, we fitted the analysis to 2 dimensions, since the goal was visualization of the biplot.
+In this case, the best solution was achieved after 20 runs for plants (stress = 0.15), 20 runs for fungi (stress = 0.12) and 20 runs for bacteria (stress = 0.09).
+When using NMDS for calculating novelty, we fitted the analysis to 4 dimensions, to allow for a better fit.
+In this case, the best solution was achieved after 20 runs for plants (stress = 0.08), 20 runs for fungi (stress = 0.07) and 20 runs for bacteria (stress = 0.04).
+The plot-centroid distances (i.e., the novelty index) were calculated by summing the distances in each of the four dimensions.
+
+After calculating the novelty indices, we tested the effect of taxonomic group (plant, fungi and bacteria), pre-urban reference habitat (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables (after reducing their dimensionality by ordination analysis, see results for more details) on compositional novelty, by fitting a Linear Mixed Model with the package *nlme* [@RN2333].
+Total sample size in the model was 120 (5 plots x 4 urban habitats x 2 reference habitats x 3 taxonomic groups).
+Because each sampling plot was included in the model six times (2 reference habitats x 3 taxonomic groups), the model included plot as a random factor.
 We checked and confirmed model assumptions using the package *performance* [@RN5684].
@@
 
-Soil temperature records (**Figure 2A**) indicated that forest soils were relatively buffered in their thermal range of variation, with a mean annual temperature of 13.9 ºC, a mean diurnal range of 1.8 ºC and an annual range of 10.8 ºC.
+Soil temperature records (**Figure 2A**) indicated that forest soil was relatively buffered in its thermal range of variation, with a mean annual temperature of 13.9 ºC, a mean diurnal range of 1.8 ºC and an annual range of 10.8 ºC.
 Meadow and urban plots had more thermal oscillation, with a mean annual temperature between 12.4 ºC (roadsides) and 16.8 ºC (meadows), a mean diurnal range between 3.8 ºC (roadsides) and 4.6 ºC (residential lots), and an annual range between 16.5 ºC (residential lots) and 19.5 ºC (meadows).
 
-PCA ordination of the soil physicochemical properties (**Appendix S2**) indicated that the first axis of variation, explaining 23% of the variance, ordered soils according to the concentration of heavy metals.
-The variables with the highest contribution to this axis (\> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr, which accounted for 66% of the variability of the first axis.
+The PCA ordination of the soil physicochemical properties (**Appendix S2**) indicated that the first axis of variation, explaining 23% of the variance, ordered soils according to the concentration of heavy metals.
+The variables with the highest contribution to this axis (\> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr.
 This first axis did not separate urban from pre-urban habitats: for example, while some industrial lots had a very high concentration of Pb and other metals, the median Pb value of the industrial habitat was below the median value of forests (**Figure 2B**; **Appendix S2**).
 
-The second PCA axis (**Appendix S2**) explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (\> 10%) being organic matter content, bulk density, N concentration and pH, which made up 59% of the axis' variability.
+The second PCA axis (**Appendix S2**) explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (\> 10%) being organic matter content, bulk density, N concentration and pH.
 This axis separated soils according to their habitat: for example, forests had high organic matter and N content, while industrial vacant lots had higher bulk density and pH (**Figure 2B**; **Appendix S2**).
 
-The third PCA axis (**Appendix S2**) explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (\>10%) being Mg content, Mg/K ratio and conductivity, together explaining 50% of the axis.
+The third PCA axis (**Appendix S2**) explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (\>10%) being Mg content, Mg/K ratio and conductivity.
 In this case, like for the first axis, there was no clear separation between habitats, with industrial sites having some of the highest and lowest values (**Figure 2B**; **Appendix S2**).
@@
 The analysis showed strong correlations within the groups of variables contributing to the first, second and third PCA axes (**Appendix S2**).
-To avoid autocorrelation issues, for further analyses we kept three variables which taken together provide a straightforward biological interpretation of patterns in soil composition: Pb content, organic matter content and Mg content (**Figure 2B**).
-As previously indicated by PCA, organic matter was the only one with a clear pattern of variation: a decrease from pre-urban to urban habitats (**Figure 2B**).
+For further analyses we kept these three axes as explanatory variables.
+For simplicity, we will refer to each axis by the name of its most explanatory variable (PC1 = Pb; PC2 = organic matter (OM); PC3 = Mg).
+
+As previously indicated by PCA, organic matter was the only soil property with a clear pattern of variation by habitat: a decrease from pre-urban to urban habitats (**Figure 2B**).
 PERMANOVA supported that there was not a significant difference between habitats in their soil physicochemical properties (100,000 permutations, R^2^ = 0.19, F = 1.15, p = 0.31).
@@
 PERMANOVA supported that the plant species composition was significantly different across habitats (100,000 permutations, R^2^ = 0.42, F = 3.48, p = \< 0.001).
-NMDS ordination of the plant species composition (**Fig.
+The NMDS ordination of the plant species composition (**Fig.
 3D**) indicated that the plots belonging to the same habitat were relatively grouped together.
 The major gradient in vegetation composition along the first axis was a separation between forest plots and the other habitats.
-Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that none of the soil factors had a significant correlation (R^2^ \< 0.14, p \> 0.05) with the composition of the plant communities.
+Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only Pb had a significant correlation (R^2^ = 0.29, p = 0.008) with the composition of the fungal communities.
+High values of Pb (and other heavy metals) were associated with the plant communities from industrial vacant lots and roadsides (**Fig.
+3D**).
 
-DNA metabarcoding of soil fungal communities resulted in 4,689 amplicon sequence variants (ASVs).
+The DNA metabarcoding of soil fungal communities resulted in 4,689 amplicon sequence variants (ASVs).
 Of these, 46% were only found in urban habitats (parks, roadsides, residential lots and industrial lots), 36% occurred only in pre-urban habitats (forests and meadows) and 18% were shared by both urban and pre-urban habitats (**Fig3B**).
 The average number of ASVs per plot was 340 (min = 174, max = 587), with the habitat averages decreasing from residential vacant lots (415) to forests (393), meadows (336), parks (323), industrial vacant lots (300) and roadsides (276).
-The majority of ASVs were assigned to Ascomycota (56%), Basidiomycota (24%), Rozellomycota (10%) and Glomeromycota (4%).
+The majority of ASVs were assigned to Ascomycota (56%), followed by Basidiomycota (24%), Rozellomycota (10%) and Glomeromycota (4%).
 Forest habitats had the highest proportion of Basidiomycota (25% vs.
@@
 PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.33, F = 2.39, p = \< 0.001).
-NMDS ordination of the fungal ASV composition (**Fig.
+The NMDS ordination of the fungal ASV composition (**Fig.
 3E**) behaved similarly to the ordination of plant communities, with plots tending to group by habitat, and the major compositional gradient reflecting the separation between forest plots and the rest.
@@
 PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.32, F = 2.51, p = \< 0.001).
-NMDS ordination of the bacterial ASV composition (**Fig.
+The NMDS ordination of the bacterial ASV composition (**Fig.
 3F**) resulted in less grouping by habitat than for plants and fungi, but still reflected a main compositional gradient from forests to industrial vacant plots.
-Like in the case of the fungal communities, environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only organic matter content had a significant correlation (R^2^ = 0.39, p = 0.001) with the composition of the bacterial communities.
+As in the case of the fungal communities, environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only organic matter content had a significant correlation (R^2^ = 0.39, p = 0.001) with the composition of the bacterial communities.
 Low organic matter content correlated with the communities from industrial vacant lots and roadsides (**Fig.
@@
 ```{r fig3, echo = FALSE, fig.cap = "Plant, fungal and bacterial diversity across habitats.
-Top panels show Venn plots for the shared numbers of (A) plant species, (B) fungal ASVs and (C) bacterial ASVs between urban (parks, roadsides, residential lots and industrial lots) and pre-urban (forests, meadows) habitats.
+Top panels show Venn plots for the shared numbers of (A) plant species, (B) fungal ASVs and (C) bacterial ASVs between pre-urban (forests, meadows) and urban (parks, roadsides, residential lots and industrial lots) habitats.
 Bottom panels show biplots of community composition ordination produced by Non-metric Multi-Dimensional Scaling (NMDS).
-NMDs were conducted for (D) the plant survey (by species abundance), (E) the soil fungal survey (by ASV presence/absence) and (F) the soil bacterial survey (by ASV presence/absence).
+The NMDS were conducted for (D) the plant survey (by species presence/absence), (E) the soil fungal survey (by ASV presence/absence) and (F) the soil bacterial survey (by ASV presence/absence).
 Colors indicate the habitat.
-In the biplots, arrows and labels indicate soil variables with a significant correlation with the composition of the communities.
-Only soil organic matter content and fungal-bacterial communities had a significant correlation."}
+In the biplots, arrows and labels indicate soil variables with a significant correlation with the composition of the communities."}
 knitr::include_graphics("../results/figures/Fig3.png")
@@
 4**).
-Novelty was significantly different between taxonomic groups (F = 45.88, p \< 0.001), being higher for plants (2.5 ± 0.3 S.E.) than for fungi (2.2 ± 0.4 S.E.) and bacteria (1.6 ± 0.3 S.E.).
-Novelty was also significantly different depending on the pre-urban habitat that was used as a reference (F = 199.08, p \< 0.001), being higher compared to forests (2.6 ± 0.3 S.E.) than to meadows (1.5 ± 0.3 S.E.).
+Novelty was significantly different between taxonomic groups (F = 40.67, p \< 0.001), being higher for plants (2.37 ± 0.34 S.E.) than for fungi (2.15 ± 0.40 S.E.) and bacteria (1.56 ± 0.30 S.E.).
+Novelty was also significantly different depending on the pre-urban habitat that was used as a reference (F = 227.99, p \< 0.001), being higher compared to forests (2.60 ± 0.28 S.E.) than to meadows (1.46 ± 0.28 S.E.).
 Nonetheless, for each taxonomic group, there was a high positive correlation between the novelty compared against forests and meadows (r = 0.5 for plants; 0.7 for fungi; 0.9 for bacteria).
-Different urban habitats also differed significantly in their novelty (F = 9.51, p = 0.001), being higher in industrial vacant lots (2.5 ± 0.3 S.E.) than in roadsides (2.2 ± 0.4 S.E.), parks (2.0 ± 0.4 S.E.) and residential vacant lots (1.6 ± 0.3 S.E.).
-The soil physicochemical properties (Pb, organic matter, Mg) had no significant effect on compositional novelty (p \> 0.05).
+Different urban habitats also differed significantly in their novelty (F = 10.06, p = 0.001), being higher in industrial vacant lots (2.39 ± 0.32 S.E.) than in roadsides (2.17 ± 0.36 S.E.), parks (1.88 ± 0.42 S.E.) and residential vacant lots (1.66 ± 0.34 S.E.).
+Of the soil physicochemical properties (i.e., the Pb, organic matter, and Mg axes), only Mg had a marginally significant positive effect on compositional novelty (F = 5.63, p = 0.03).
 
@@
 
-## Widespread soil contamination across urban habitats
-
-Our results partly agreed with the expected effects of urbanization on soils [@RN5623; @RN5603]: we found more microclimatic variation, less organic matter content and more compacted and alkaline soils than their pre-urban counterparts.
-But, contrary to our expectations, urban and pre-urban soils did not show a consistent pattern of contamination by heavy metals, possibly because pollution is widespread across the city.
-Indeed, soil contamination had a weak but positive correlation (r = 0.25) with the geographical proximity to the hotspot of heavy industry in the western end of the city [@RN5683].
-This geographical correlation was especially strong for the residential lots (0.89), parks (0.68) and forests (0.68); and weak for the roadsides (0.10), industrial lots (0.03) and meadows (-0.14).
-Overall, the measured soil properties had a minor effect on the composition of the biological communities, as organic matter content was the only variable to influence the composition of soil fungal and bacterial communities.
-
 ## Contrasting diversity patterns in plant-bacterial vs. fungal communities
@@
 This agrees with previous reports of fungal communities decreasing with urbanization [@RN5609].
-One potential explanation for this pattern is the tendency of urban soils to be alkaline, favoring bacteria over fungi [@RN5603; @RN5606].
+One potential explanation for this decrease is the tendency of urban soils to be alkaline, favoring bacteria over fungi [@RN5603; @RN5606].
 Indeed, in our study, organic matter content (highly and inversely correlated with pH) was the soil variable that had an effect on the composition of fungal and bacterial communities; and bacterial richness was higher in urban habitats versus forests (i.e., the opposite to fungi).
@@
 As such, this index can help to characterize and compare novelty across urban habitats, which is a major research goal for novel ecosystem research [@RN5673; @RN5674].
-Contrary to our expectations, managed park lawns showed a relatively low novelty when compared to agricultural meadows.
+Contrary to our expectations, managed park lawns showed a relatively low novelty for all three taxonomic groups when compared to agricultural meadows.
 This indicates that park lawns have a high potential for low-cost urban rewilding policies, such as low-frequency and late mowing, which could convert lawns into urban grasslands rich in native biota and resembling traditional hay meadows.

### 2025-08-22 doc/R1/novelty-PCA.Rmd
LANDUP R2

--- 2024-07-11 doc/manuscript.Rmd
@@
 ---
-title: "Soil drivers of plant, fungal and bacterial community novelty in urban habitats of the city of Xixón, Spain"
-output: 
-  bookdown::github_document2:
-    reference_docx: template.docx
-    number_sections: false
-bibliography: refs.bib
-csl: journal-of-vegetation-science.csl
-link-citations: yes
-always_allow_html: true
+title: "Untitled"
+output: html_document
+date: "2025-06-24"
 ---
@@
 
-```{r message = FALSE, echo = FALSE, warning = FALSE}
-knitr::knit_hooks$set(inline = function(x) {
-  prettyNum(x, big.mark = ",")
-})
+## Compositional novelty across habitats and taxonomic groups
+
+We used the NMDS multidimensional compositional distance (calculated from the ordinations in **Fig.
+3D-F**) from each of the urban plots (i.e., park, roadside, residential or industrial) to the centroid of the two reference pre-urban habitats (i.e., forests and meadows) to calculate a compositional novelty index (**Fig.
+4**).
+Novelty was significantly different between taxonomic groups (F = 65.83, p \< 0.001), being higher for plants (1.55 ± 0.14 S.E.) than for fungi (1.20 ± 0.10 S.E.) and bacteria (1.01 ± 0.14 S.E.).
+Novelty was also significantly different depending on the pre-urban habitat that was used as a reference (F = 59.89, p \< 0.001), being higher compared to forests (1.40 ± 0.15 S.E.) than to meadows (1.10 ± 0.15 S.E.).
+Different urban habitats also differed significantly in their novelty (F = 8.03, p = 0.0028), being higher in industrial vacant lots (1.41 ± 0.14 S.E.) than in parks (1.26 ± 0.16 S.E.), roadsides (1.24 ± 0.17 S.E.) and residential vacant lots (1.09 ± 0.16 S.E.).
+The soil physicochemical properties (Pb, organic matter, Mg) had no significant effect on compositional novelty (p \> 0.05).
+
+```{r fig4, echo = FALSE, fig.cap = "Compositional novelty for plants, fungi and bacteria.
+Compositional novelty was calculated as the NMDS multidimensional distance from each urban plot to the centroid of two reference pre-urban habitats (forests, panels A-C-E; and meadows, panels B-D-F).
+Novelty was calculated separately for plants (panels A-B), fungi (panels C-D) and bacteria (panels E-F)."}
+knitr::include_graphics("../results/figures/Fig4-PCA.png")
 ```
-
-Running title: Urban drivers of plant, fungal and bacterial novelty
-
-Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^2^ (<https://orcid.org/0000-0002-0915-0916>),
-Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)
-
-^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain
-
-Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain.
-Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}.
-Telephone: +34985104781.
-
-# Author contributions
-
-Eduardo Fernández-Pascual: Conceptualization; Methodology; Investigation; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing.
-Zuzana Ferencova: Conceptualization; Methodology; Investigation; Data Curation; Writing – Review & Editing.
-Víctor González-García: Investigation; Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.
-
-# Acknowledgements
-
-María García Fernández, Sara Gutiérrez Cruz and Abelardo Sigüenza Solís helped with the field surveys.
-
-# Funding
-
-Fundación Biodiversidad, Ministerio para la Transición Ecológica y el Reto Demográfico, Gobierno de España [XIXÓN ECORRESILIENTE, PRTR, NextGenerationEU].
-Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico [SV-23-GIJON-JBA].
-
-# Conflict of interest statement
-
-None declared.
-
-# Data availability statement
-
-Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at Zenodo: https://zenodo.org/doi/10.5281/zenodo.11613571.
-
-# Introduction
-
-In post-industrial landscapes affected by the phenomenon of urban contraction, land located in the heart of cities remains unused for long periods (Haase et al., 2014) .
-When the use of any urban plot ceases, it is colonized by plants from the regional flora capable of living in urban environments (Johnson et al., 2017) .
-These urban plants are adapted to disturbances and can grow in degraded soils; some even accumulate heavy metals, making them nature-based solutions for bioremediation of degraded lands (Song et al., 2019) .
-Species traits, along with the legacy of human impact, determine which regional flora species colonize which plots (Swan et al., 2021) .
-When the human legacy is intense, the plant communities that develop will be novel communities, different from the historical reference communities that existed before human intervention (Gedan and Fernández-Pascual, 2019) .
-The novelty of the communities can be reflected in their floristic, compositional and phylogenetic composition, including their proportion of exotic species (Lososová et al., 2012) .
-In any case, proper management of urban communities can turn them into resources with high ecological and social potential (Anderson and Minor, 2017).
-(Kowarik, 2018) : in the Ibero-Atlantic ecoregion, anthropogenic habitats can host up to a third of the ecoregional flora.
-
-The ability of plants to colonize any habitat, including urban ones, depends largely on their seed regeneration niche, that is, on their dispersal, persistence, germination and establishment functions (Grubb, 1977, Saatkamp et al., 2019) .
-However, although it is known that urban species often share a series of functional and ecological traits (Aronson et al., 2007, Lososová et al., 2006) , little is known about their seed regeneration niche.
-Especially, information on its germination traits is lacking.
-In species living in non-anthropogenic environments, germination phenology is regulated by environmental factors (water, temperature, light) to ensure that germination occurs at the right time to maximize subsequent seedling survival (Carta et al., 2022a) .
-Compared to the large amount of data on the germination of species from non-anthropogenic habitats, little is known about what factors regulate the germination of urban species.
-Considering other habitats affected by humans (Fernández-Pascual et al., 2022) , it could be expected that urban seeds have their germination unrelated to macroclimatic signals and show tolerance to stress due to heat, drought or heavy metals.
-However, this hypothesis needs to be tested through a synthesis of the seed regeneration niche in urban environments.
-
-Cities 
-1. highly fragmented landscapes
-2. mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use)
-3. subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems
-4. comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure
-5.
-contains a multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix
-
-Cities are comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure.
-They are mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use) which are subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems (Kowarik, 2005).
-
-The “Four Nature approach” narrows down the variety of transformational stages in urban settings to four major types that can supply wilderness in urban region (Kowarik 1992, 2005).
-Each of the four types relates to nature in general but results from different trajectories in human-nature interactions:
-
-Spontaneous vegetation appears in different habitat types in the urban matrix,.
-Multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix.
-The spontaneous herbs could provide the very interesting pool of species for ecological restoration of the urban area.
-Their ecological value has been clear since many years because their contribution to the ecosystem services in urban areas.
-Although their aesthetic value has been highlighted repeatedly, their acceptance by the residents in urban areas has been questioned.
-Many residents associate the presence of spontaneous plants with the neglect in maintenance by public (municipal) management service.
-Spontaneous native urban vegetation has commonly been described as demonstrating resilience (Ignatieva et al., 2000) and exhibiting adaptations to human disturbance (Lundholm and Marlin, 2006; Sukopp, 2004).
-
-Urban species pool is comprised of native species present before the urban development, native species developed because the new urban ecosystem and introduced species by human activity.
-However, just some of them can adapt to urban environment.
-
-IGS (informal green space) consists of any urban space with a history of strong anthropogenic disturbance that is covered at least partly with non-remnant, spontaneous vegetation (Del Tredici, 2010).
-
-# Materials and methods
-
-## Study city
-
-Gijón/Xixón is a medium-sized coastal city (population c.
-270,000 inhabitants) located in the Bay of Biscay, within the central Asturias metropolitan area of northern Spain (population c.
-800,000 inhabitants).
-The city belongs to the Cantabrian Mixed Forests ecoregion [@RN4667], at the border between the temperate and Mediterranean climatic zones of Europe, and has a temperate oceanic climate with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm).
-Most of the city is built on alluvial and estuarine deposits, sand, or limestone bedrock.
-The potential pre-human vegetation would be dominated by deciduous mixed forests, together with riparian forests, wetlands and coastal dunes.
-The area has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450].
-The city itself has been inhabited at least since Roman times [@RN5532].
-During the medieval and early modern era the urban center was limited to the original walled city, with fishing as the main economic activity.
-Expansion beyond the original core started in the 18th and 19th centuries.
-Exposure to global trade networks and the mild climate turned the region into a hotspot for biological invasions, especially from the Americas and Asia [@RN5493].
-In 1852, the city was connected by railway and became a major port for the regional mining and industrial centers.
-During the 20th century the city experienced significant industrial development (e.g.
-canning, coal, steel, shipbuilding) and rapid population growth, becoming the most populated city in the Principality of Asturias [@RN5534; @RN5533].
-Industrial and urban growth brought environmental problems including loss of natural wetlands and air and watercourse pollution [@RN5535].
-Like the rest of its metropolitan area, the city has been affected by recent post-industrial land-use changes which have left large areas with abandoned industrial sites (i.e.
-brownfields) [@RN5448; @RN5449].
-
-## Definition of urban habitats
-
-To survey urban biodiversity across habitats, we defined six habitats types (**Figure 1A**) representing different states within the four-nature classification of urban wilderness [@RN5445]: (1) forest patches < 10 km from the city center, i.e.
-type-1 nature in remnants of the pre-urban vegetation; (2) hay meadows, i.e.
-type-2 nature in remnants of human-made ecosystems resulting from agricultural habitat transformation; (3) parks, i.e.
-type-3 nature in managed green spaces created after habitat destruction; (4) roadsides, i.e.
-type-4 nature in relatively unmanaged strips of land alongside roads, railways, etc; (5) residential lots, i.e.
-type-4 nature in unmanaged vacant lots of previous residential use; and (6) industrial lots, i.e.
-type-4 nature in unmanaged vacant lots of previous  industrial use.
-
-## Selection of sampling plots
-
-During June 2023, we surveyed the city to establish five 4 x 4 m sampling plots per each of the six habitat types (n = 30 plots).
-The plots were distributed across the city (**Figure 1B**) and represented the variation of vegetation types found within each habitat type.
-In two plots per habitat (**Figure 1B**), we buried a datalogger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil.
-Dataloggers recorded temperature once every hour and stayed on site for approximately one year, after which we retrieved them and downloaded their records to describe the thermal environment of the city and the habitats.
-
-## Soil physicochemical sampling
-
-From each of the 30 plots, we collected 500 g of soil for analysis of chemical composition and physical soil properties.
-We took the samples from 5 points in each plot (center + 4 cardinal points).
-We removed stones and other debris particles, sieved the soil in a 5 mm mesh and homogenized the sample in a plastic bag.
-We cleaned and sterilized all tools with water and ethanol between samples.
-We sent the samples for analysis at INEA - Escuela Universitaria de Ingeniería Agrícola, Universidad Pontificia de Comillas.
-Soil analysis included pH, texture, bulk density, electrical conductivity, organic matter content, total nitrogen, C/N ratio, active limestone, total carbonates, assimilable phosphorus, K, Ca, Mg, Na and heavy metals (Zn, Cr, Cu, Cd, Hg, Pb, Ni).
-
-> Añadir detalles de los métodos de suelo (Zuzana)
-
-## Plant, fungal and bacterial diversity sampling
-
-In each plot, we recorded all the spontaneous vascular plant species, as well as the cover (%) of each species in the plot.
-For plant species names we followed Euro+Med [@RN5044].
-Additionally, from each of the 30 plots, we collected soil samples for DNA metabarcoding analysis of soil fungal and bacterial communities.
-We performed the sampling in sterile conditions, i.e.
-before entering the plot for any of the other tasks.
-In each plot, we took samples from the four cardinal points, 0.5 m inwards from the edge, at a depth of 5 cm.
-Before taking the samples, we removed the vegetation with a sterile tool.
-Then, we collected the soil samples in a 50 ml Falcon tube, mixed them, homogenized them with a lancet and finally stored the sample into a 2 ml Eppendorf tube which we immediately froze.
-We sent the frozen tube to AllGenetics & Biology SL (www.allgenetics.eu) for DNA metabarcoding analysis.
-
-## DNA metabarcoding analysis
-
-DNA analysis by AllGenetics included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size < 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, taxonomic assignment, number of reads per taxon and generation of rarefaction curves.
-
-*DNA isolation and quantification*.
-DNA was isolated from each soil sample using the DNeasy PowerSoil Pro DNA isolation kit (Qiagen) following the manufacturer’s instructions, and DNA was eluted in a final volume of 50 μL.
-In one sample the above-mentioned protocol did not yield a DNA extract of enough concentration, so DNA was isolated using the FastDNA Spin Kit for Soil (MPBio) following the manufacturer’s instructions, and DNA was eluted in a final volume of 70 μL.
-An extraction blank (Bex) was included in each round of DNA extraction, and treated as the plot samples to check for contamination.
-DNA concentration in each extract was quantified using the Qubit High Sensitivity dsDNA Assay (Thermo Fisher Scientific).
-
-*DNA metabarcoding library preparation and sequencing*.
-For fungal library preparation, a fragment of the nuclear ITS1 region of around 280-300 bp (including the primer sequences) was amplified using the following primers: forward - ITS1f (5’ CTTGGTCATTTAGAGGAAGTAA 3’) [@RN5548]; reverse - ITS2 (5’ GCTGCGTTCTTCATCGATGC 3’) [@RN5549].
-For prokaryotic library preparation, a fragment of the 16S rRNA gene of around 300 bp (including primer sequences) was amplified using the following primers: forward - 515F-Y (5’ GTGYCAGCMGCCGCGGTAA 3’) [@RN5536]; reverse - 806RB (5’ GGACTACNVGGGTWTCTAAT 3’) [@RN5537].
-These primers also included the Illumina sequencing primer sequences attached to their 5’ ends.
-In the first amplification step, PCRs were carried out in a final volume of 12.5 μL, containing 1.25 μL of template DNA (1:2 diluted in PCR round 3; not diluted in PCR round 11), 0.5 μM of the primers, 3.13 μL of Supreme NZYTaq 2x Green Master Mix (NZYTech), and ultrapure water up to 12.5 μL.
-The fungal reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 35 cycles of 95 ºC for 30 s, 49 ºC for 45 s, 72 ºC for 45 s, and a final extension step at 72 ºC for 7 min.
-The bacterial reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 25 cycles of 95 ºC for 30 s, 47 ºC for 45 s, 72ºC for 45 s, and a final extension step at 72 ºC for 7 min.
-The oligonucleotide indices that are required for multiplexing different libraries in the same sequencing pool were attached in a second amplification step with identical conditions but only 5 cycles and 60 ºC as the annealing temperature.
-A negative control that contained no DNA (BPCR) was included in every PCR round to check for contamination during library preparation.
-Library size was verified by running the libraries on 2 % agarose gels stained with GreenSafe (NZYTech) and imaging them under UV light.
-Then, libraries were purified using the Mag-Bind RXNPure Plus magnetic beads (Omega Bio-tek), following the instructions provided by the manufacturer.
-Finished libraries were pooled in equimolar amounts according to the results of a Qubit dsDNA HS Assay (Thermo Fisher Scientific) quantification.
-The pool was sequenced in a fraction of a NovaSeq PE250 flow cell (Illumina) aiming for a total output of 2 gigabases.
-Any potential traces of adapter dimers were removed using Cutadapt v3.5 [@RN5538].
-Quality of the sequencing data was assesed with the software FastQC [@RN5539].
-
-*Processing of sequencing data and inference of ASVs*.
-Cutadapt v3.5 was used to trim off non-biological DNA sequences (primers, indices, and sequencing adapters).
-Then, amplicon reads were processed using the DADA2 pipeline [@RN5541] implemented in QIIME 2 (release 2023.7) [@RN5540].
-The first step in the DADA2 pipeline was to remove PCR primers and filter reads according to their quality.
-After checking the read quality profiles, forward reads were truncated at position 175, and reverse reads at position 131.
-The next step was to denoise and infer Amplicon Sequence Variants (ASVs).
-Error rates were learned from the dataset to denoise, using the parametric error model implemented in DADA2.
-Dereplication of the dataset (i.e.
-combination of all identical reads into unique reads) was carried out to reduce computational effort.
-These dereplicated forward and reverse reads were used to infer ASVs with the core sample inference algorithm [@RN5541].
-Subsequently, corresponding R1 and R2 reads were merged into pairs with a minimum overlap of 12 identical base pairs.
-The final step in the DADA2 pipeline was to reduce the impact of artefacts (chimaeras) in the dataset.
-The final output of the DADA2 pipeline was a table containing the count of reads of every observed ASV in each sample, as well as a table listing all ASVs and their corresponding representative sequences.
-
-*Taxonomic assignment*.
-Taxonomic assignment of each ASV was performed using a pre-trained classifier of the UNITE reference database for fungi [@RN5550] (release May 2021); and of the SILVA reference database for bacteria [@RN5542] (release 138.1 August 2020).
-To compare the representative sequences of the ASVs to the reference database, and compute the taxonomic assignment,the algorithm sklearn [@RN5543] implemented in QIIME 2 was applied, setting a minimum confidence level of 0.8.
-Several filters were applied to exclude: (1) singletons (i.e.
-ASVs containing only one member sequence in the whole data set); (2) ASVs occurring at a frequency below 0.01 % in each sample, to reduce the mistagging bias [@RN5544]; (3) from the bacterial dataset, non-prokaryotic ASVs such as eukaryotic sequences of plastid [@RN5545] and mitochondrial [@RN5546] origin; (4) those assigned only at domain (Bacteria) or kingdom (Fungi) level; (5) ASVs present in the negative controls that were more abundant in the negative controls than in the samples, indicating contamination [@RN5547].
-
-*Alpha rarefaction plots*.
-Alpha rarefaction plots showed that the sequencing depth to reach saturation was 30,000 reads for the fungi and 22,000 reads for the bacteria, above the sequencing depth of all plots for fungi (mean = 114,919 reads; min = 79,045; max = 157,110) and bacteria (mean = 113,341 reads; min = 84,094; max = 171,404).
-
-> Rarefaction
-
-## Statistical analysis
-
-# Results
-
-## Soil temperatures and properties across habitats
-
-## Plant, fungal and bacterial diversity across habitats
-
-> PERMANOVAs
-
-> Analysis with species, genera and ASVs
-
-> Analysis with presence absence
-
-## Community novelty across habitats and taxonomic groups
-
-## Soil drivers of novelty
-
-# Discussion
-
-The recorded flora represents just a subset of possible spontaneous vegetation because this study was carried on during limited time which covers just part of the growing season of the whole vegetation.
-
-There are some other habitats in the city which were not included in the present study, e.g. wetlands, beach, cliffs.

### 2025-08-22 doc/R1/title-page.Rmd
LANDUP R2

--- 2025-07-07 doc/R0/title-page.Rmd
@@
 
-Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^2^ (<https://orcid.org/0000-0002-0915-0916>), Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)
+Eduardo Fernández-Pascual^1,2,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^3^ (<https://orcid.org/0000-0002-0915-0916>), Víctor González-García^1,2^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1,2^ (<https://orcid.org/0000-0001-6601-9597>)
 
-^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain
+^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Asturias, Spain; ^2^ Department of Organismal and Systems Biology, University of Oviedo, Oviedo/Uviéu, Asturias, Spain; ^3^ Jardín Botánico Atlántico, Gijón/Xixón, Spain
 
@@
 Thanks to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis and to INEA Laboratory (Valladolid, Spain) for soil physicochemical analysis.
-We are thankful to Fundación Biodiversidad for their support.
 
@@
 
-European Union - NextGenerationEU within the framework of the Recovery, Transformation and Resilience Plan (PRTR) of Ministry for the Ecological Transition and the Demographic Challenge (MITECO) of the Government of Spain —Project: Gijón Ecoresiliente (PRTR 0204103/010000).
-Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico [SV-23-GIJON-JBA].
+The project is supported by the Biodiversity Foundation of the Ministry for the Ecological Transition and the Demographic Challenge (MITECO) of the Government of Spain, within the framework of the Recovery, Transformation and Resilience Plan (PRTR), funded by the European Union - NextGenerationEU.
+Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico (SV-23-GIJON-JBA).
 Grant "Laboratorio de Vegetación y Biodiversidad" (IDE/2024/000720, Principality of Asturias-Sekuens-EU-FEDER).

### 2025-08-22 doc/manuscript.Rmd
LANDUP R2

@@
 The frozen tubes were sent to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis.
-DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi, and 16S for bacteria; sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size \< 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, inference of Amplicon Sequence Variants (ASVs), taxonomic assignment, number of reads per taxon and generation of rarefaction curves (see **Appendix S1** for extended DNA metabarcoding methodology).
+The DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi, and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size \< 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, inference of Amplicon Sequence Variants (ASVs), taxonomic assignment, number of reads per taxon and generation of rarefaction curves (see **Supplementary material 1** for extended DNA metabarcoding methodology).
 
@@
 We calculated compositional novelty separately for plant, fungal and bacterial communities.
-We tested the effect of taxonomic group, pre-urban reference habitat (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables on compositional novelty, by fitting a Generalized Mixed Model with the package *nlme* [@RN2333].
-The model included sampling plot as a random factor.
+Because NMDS employs an iterative algorithm to seek optimal solutions, the results may vary across different computational runs, leading to inconsistencies in ordination outcomes.
+Therefore, we also considered alternative ordination methods based on eigenvalue decomposition (Principal Component Analysis PCA with Hellinger transformation, and Principal Coordinates Analysis PCoA of a Bray–Curtis dissimilarity matrix).
+The final output and interpretation of the analyses remained stable independently of the ordination method used, and therefore we decided to use NMDS for the final presentation of results, since it makes fewer assumptions about data and is therefore a more flexible technique that can be applied to a wider variety of datasets.
+We computed NMDS using the Bray-Curtis distance.
+When using NMDS for assessing variation in taxonomic composition, we fitted the analysis to 2 dimensions, since the goal was visualization of the biplot.
+In this case, the best solution was achieved after 20 runs for plants (stress = 0.15), 20 runs for fungi (stress = 0.12) and 20 runs for bacteria (stress = 0.09).
+When using NMDS for calculating novelty, we fitted the analysis to 4 dimensions, to allow for a better fit.
+In this case, the best solution was achieved after 20 runs for plants (stress = 0.08), 20 runs for fungi (stress = 0.07) and 20 runs for bacteria (stress = 0.04).
+The plot-centroid distances (i.e., the novelty index) were calculated by summing the distances in each of the four dimensions.
+
+After calculating the novelty indices, we tested the effect of taxonomic group (plant, fungi and bacteria), pre-urban reference habitat (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables (after reducing their dimensionality by ordination analysis, see results for more details) on compositional novelty, by fitting a Linear Mixed Model with the package *nlme* [@RN2333].
+Total sample size in the model was 120 (5 plots x 4 urban habitats x 2 reference habitats x 3 taxonomic groups).
+Because each sampling plot was included in the model six times (2 reference habitats x 3 taxonomic groups), the model included plot as a random factor.
 We checked and confirmed model assumptions using the package *performance* [@RN5684].
@@
 
-Soil temperature records (**Figure 2A**) indicated that forest soils were relatively buffered in their thermal range of variation, with a mean annual temperature of 13.9 ºC, a mean diurnal range of 1.8 ºC and an annual range of 10.8 ºC.
+Soil temperature records (**Figure 2A**) indicated that the forest soil was relatively buffered in its thermal range of variation, with a mean annual temperature of 13.9 ºC, a mean diurnal range of 1.8 ºC and an annual range of 10.8 ºC.
 Meadow and urban plots had more thermal oscillation, with a mean annual temperature between 12.4 ºC (roadsides) and 16.8 ºC (meadows), a mean diurnal range between 3.8 ºC (roadsides) and 4.6 ºC (residential lots), and an annual range between 16.5 ºC (residential lots) and 19.5 ºC (meadows).
 
-PCA ordination of the soil physicochemical properties (**Appendix S2**) indicated that the first axis of variation, explaining 23% of the variance, ordered soils according to the concentration of heavy metals.
-The variables with the highest contribution to this axis (\> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr, which accounted for 66% of the variability of the first axis.
-This first axis did not separate urban from pre-urban habitats: for example, while some industrial lots had a very high concentration of Pb and other metals, the median Pb value of the industrial habitat was below the median value of forests (**Figure 2B**; **Appendix S2**).
+The PCA ordination of the soil physicochemical properties (**Supplementary material 2**) indicated that the first axis of variation, explaining 23% of the variance, ordered soils according to the concentration of heavy metals.
+The variables with the highest contribution to this axis (\> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr.
+This first axis did not separate urban from pre-urban habitats: for example, while some industrial lots had a very high concentration of Pb and other metals, the median Pb value of the industrial habitat was below the median value of forests (**Figure 2B**; **Supplementary material 2**).
 
-The second PCA axis (**Appendix S2**) explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (\> 10%) being organic matter content, bulk density, N concentration and pH, which made up 59% of the axis' variability.
-This axis separated soils according to their habitat: for example, forests had high organic matter and N content, while industrial vacant lots had higher bulk density and pH (**Figure 2B**; **Appendix S2**).
+The second PCA axis (**Supplementary material 2**) explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (\> 10%) being organic matter content, bulk density, N concentration and pH.
+This axis separated soils according to their habitat: for example, forests had high organic matter and N content, while industrial vacant lots had higher bulk density and pH (**Figure 2B**; **Supplementary material 2**).
 
-The third PCA axis (**Appendix S2**) explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (\>10%) being Mg content, Mg/K ratio and conductivity, together explaining 50% of the axis.
-In this case, like for the first axis, there was no clear separation between habitats, with industrial sites having some of the highest and lowest values (**Figure 2B**; **Appendix S2**).
+The third PCA axis (**Supplementary material 2**) explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (\>10%) being Mg content, Mg/K ratio and conductivity.
+In this case, as for the first axis, there was no clear separation between habitats, with industrial sites having some of the highest and lowest values (**Figure 2B**; **Supplementary material 2**).
 
-The analysis showed strong correlations within the groups of variables contributing to the first, second and third PCA axes (**Appendix S2**).
-To avoid autocorrelation issues, for further analyses we kept three variables which taken together provide a straightforward biological interpretation of patterns in soil composition: Pb content, organic matter content and Mg content (**Figure 2B**).
-As previously indicated by PCA, organic matter was the only one with a clear pattern of variation: a decrease from pre-urban to urban habitats (**Figure 2B**).
+The analysis showed strong correlations within the groups of variables contributing to the first, second and third PCA axes (**Supplementary material 2**).
+For further analyses we kept these three axes as explanatory variables.
+For simplicity, we will refer to each axis by the name of the explanatory variable most strongly associated with that axis (PC1 = Pb; PC2 = organic matter (OM); PC3 = Mg).
+
+As previously indicated by PCA, organic matter was the only soil property with a clear pattern of variation by habitat: a decrease from pre-urban to urban habitats (**Figure 2B**).
 PERMANOVA supported that there was not a significant difference between habitats in their soil physicochemical properties (100,000 permutations, R^2^ = 0.19, F = 1.15, p = 0.31).
@@
 PERMANOVA supported that the plant species composition was significantly different across habitats (100,000 permutations, R^2^ = 0.42, F = 3.48, p = \< 0.001).
-NMDS ordination of the plant species composition (**Fig.
+The NMDS ordination of the plant species composition (**Fig.
 3D**) indicated that the plots belonging to the same habitat were relatively grouped together.
 The major gradient in vegetation composition along the first axis was a separation between forest plots and the other habitats.
-Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that none of the soil factors had a significant correlation (R^2^ \< 0.14, p \> 0.05) with the composition of the plant communities.
+Environmental fitting of the soil property axes PC1, PC2, and PC3 onto the ordination indicated that only PC1 (mainly explained by Pb) had a significant correlation (R^2^ = 0.29, p = 0.008) with the plant community composition.
+High values of Pb (and other heavy metals) were associated with the plant communities from industrial vacant lots and roadsides (**Fig.
+3D**).
 
-DNA metabarcoding of soil fungal communities resulted in 4,689 amplicon sequence variants (ASVs).
+The DNA metabarcoding of soil fungal communities resulted in 4,689 amplicon sequence variants (ASVs).
 Of these, 46% were only found in urban habitats (parks, roadsides, residential lots and industrial lots), 36% occurred only in pre-urban habitats (forests and meadows) and 18% were shared by both urban and pre-urban habitats (**Fig3B**).
 The average number of ASVs per plot was 340 (min = 174, max = 587), with the habitat averages decreasing from residential vacant lots (415) to forests (393), meadows (336), parks (323), industrial vacant lots (300) and roadsides (276).
-The majority of ASVs were assigned to Ascomycota (56%), Basidiomycota (24%), Rozellomycota (10%) and Glomeromycota (4%).
+The majority of ASVs were assigned to Ascomycota (56%), followed by Basidiomycota (24%), Rozellomycota (10%) and Glomeromycota (4%).
 Forest habitats had the highest proportion of Basidiomycota (25% vs.
@@
 PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.33, F = 2.39, p = \< 0.001).
-NMDS ordination of the fungal ASV composition (**Fig.
+The NMDS ordination of the fungal ASV composition (**Fig.
 3E**) behaved similarly to the ordination of plant communities, with plots tending to group by habitat, and the major compositional gradient reflecting the separation between forest plots and the rest.
-Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only organic matter had a significant correlation (R^2^ = 0.46, p \< 0.001) with the composition of the fungal communities.
+Environmental fitting of the soil property axes PC1, PC2, and PC3 onto the ordination indicated that only PC2 (mainly explained by organic matter) had a significant correlation (R^2^ = 0.46, p \< 0.001) with the fungal community composition.
 Low organic matter content was associated with the fungal communities from industrial vacant lots and roadsides (**Fig.
@@
 PERMANOVA performed on the presence/absence of ASVs indicated significant differences in composition between habitats (100,000 permutations, R^2^ = 0.32, F = 2.51, p = \< 0.001).
-NMDS ordination of the bacterial ASV composition (**Fig.
+The NMDS ordination of the bacterial ASV composition (**Fig.
 3F**) resulted in less grouping by habitat than for plants and fungi, but still reflected a main compositional gradient from forests to industrial vacant plots.
-Like in the case of the fungal communities, environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that only organic matter content had a significant correlation (R^2^ = 0.39, p = 0.001) with the composition of the bacterial communities.
+As in the case of the fungal communities, environmental fitting of the soil property axes PC1, PC2, and PC3 onto the ordination indicated that only PC2 (mainly explained by organic matter) had a significant correlation (R^2^ = 0.39, p = 0.001) with the bacterial community composition.
 Low organic matter content correlated with the communities from industrial vacant lots and roadsides (**Fig.
@@
 
-We used the NMDS multidimensional compositional distance (calculated from the ordinations in **Fig.
-3D-F**) from each of the urban plots (i.e., park, roadside, residential or industrial) to the centroid of the two reference pre-urban habitats (i.e., forests and meadows) to calculate a compositional novelty index (**Fig.
+We used the NMDS multidimensional compositional distance from each of the urban plots (i.e., park, roadside, residential or industrial) to the centroid of the two reference pre-urban habitats (i.e., forests and meadows) to calculate a compositional novelty index (**Fig.
 4**).
-Novelty was significantly different between taxonomic groups (F = 45.88, p \< 0.001), being higher for plants (2.5 ± 0.3 S.E.) than for fungi (2.2 ± 0.4 S.E.) and bacteria (1.6 ± 0.3 S.E.).
-Novelty was also significantly different depending on the pre-urban habitat that was used as a reference (F = 199.08, p \< 0.001), being higher compared to forests (2.6 ± 0.3 S.E.) than to meadows (1.5 ± 0.3 S.E.).
+Novelty was significantly different between taxonomic groups (F = 40.67, p \< 0.001; see **Supplementary material 3** for full details of the Linear Mixed Model), being higher for plants (2.37 ± 0.34 S.E.) than for fungi (2.15 ± 0.40 S.E.) and bacteria (1.56 ± 0.30 S.E.).
+Novelty was also significantly different depending on the pre-urban habitat that was used as a reference (F = 227.99, p \< 0.001), being higher compared to forests (2.60 ± 0.28 S.E.) than to meadows (1.46 ± 0.28 S.E.).
 Nonetheless, for each taxonomic group, there was a high positive correlation between the novelty compared against forests and meadows (r = 0.5 for plants; 0.7 for fungi; 0.9 for bacteria).
-Different urban habitats also differed significantly in their novelty (F = 9.51, p = 0.001), being higher in industrial vacant lots (2.5 ± 0.3 S.E.) than in roadsides (2.2 ± 0.4 S.E.), parks (2.0 ± 0.4 S.E.) and residential vacant lots (1.6 ± 0.3 S.E.).
-The soil physicochemical properties (Pb, organic matter, Mg) had no significant effect on compositional novelty (p \> 0.05).
+Different urban habitats also differed significantly in their novelty (F = 10.06, p = 0.001), being higher in industrial vacant lots (2.39 ± 0.32 S.E.) than in roadsides (2.17 ± 0.36 S.E.), parks (1.88 ± 0.42 S.E.) and residential vacant lots (1.66 ± 0.34 S.E.).
+Of the soil physicochemical properties (i.e., the Pb, organic matter, and Mg axes), only Mg had a marginally significant positive effect on compositional novelty (F = 5.63, p = 0.03).
 
@@
 
-**Appendix S1** Extended methodology for the soil DNA metabarcoding analysis.
+**Supplementary material 1** Extended methodology for the soil DNA metabarcoding analysis.
 
-**Appendix S2** Principal Component Analysis and correlation analysis of the soil physicochemical properties.
+**Supplementary material 2** Principal Component Analysis and correlation analysis of the soil physicochemical properties.
+
+**Supplementary material 3** Full results of the Linear Mixed Model.

### 2025-08-22 results/supplementary/Supplementary-material-1.Rmd
LANDUP R2

--- 2025-03-17 results/supplementary/AppendixS1.Rmd
@@
 ---
-title: "Appendix S1 Extended methodology for the soil DNA metabarcoding analysis"
+title: "Supplementary material 1 Extended methodology for the soil DNA metabarcoding analysis"
 output:
@@
 bibliography: ../../doc/refs.bib
-csl: ../../doc/urban-forestry-and-urban-greening.csl
+csl: ../../doc/landscape-and-urban-planning.csl
 link-citations: yes

### 2025-08-22 results/supplementary/Supplementary-material-2.Rmd
LANDUP R2

--- 2025-03-17 results/supplementary/AppendixS2.Rmd
@@
 ---
-title: "Appendix S2 Principal Component Analysis and correlation analysis of the soil physicochemical properties"
+title: "Supplementary material 2 Principal Component Analysis and correlation analysis of the soil physicochemical properties"
 output:

### 2025-08-22 results/supplementary/Supplementary-material-3.Rmd
LANDUP R2

---
title: "Supplementary material 3 Full results of the Linear Mixed Model"
output:
  bookdown::pdf_document2:
    reference_docx: template.docx
    number_sections: false
    toc: false
link-citations: yes
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark = ",")
})
```

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse)
load(file = "../model/m1.Rdata")
```

## Model summary

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(nlme)
summary(m1)
```

## ANOVA table

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(nlme)
anova.lme(m1)
```

## Diagnostic plots

```{r message = FALSE, echo = FALSE, warning = FALSE}
library(performance)
library(see)
check_model(m1)
```

### 2025-08-22 src/6-novelty-R2-PCoA.R

@@ -1,0 +1,318 @@
+library(tidyverse); library(vegan)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+### Distance matrix
+
+dist <- vegdist(matrix,  method = "bray")
+
+### PCoA
+
+PCoA <- wcmdscale(d = dist)
+
+### Site scores
+
+PCoA %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus forest)
+
+sitescores %>%
+  filter(habitat %in% c("forest")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = forest - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
+
+### Novelty (versus meadow)
+
+sitescores %>%
+  filter(habitat %in% c("meadow")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = meadow - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
+
+### Compositional novelty for fungi
+
+fungi %>%
+  select(id, ASV, presence) %>%
+  spread(ASV, presence, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
[…]

### 2025-08-22 src/6-novelty.R

@@ -270,6 +270,7 @@ df1 %>%
   summarise(m = mean(Value), se = sd(Value)/sqrt(5))
 
 nlme::lme(Value ~ Group + Reference + habitat + OM + Pb + Mg,  random=~1 | id, data = df1) -> m1
+save(m1, file = "results/model/m1.RData")
 summary(m1)
 nlme::anova.lme(m1)
 

### 2025-09-30 src/5-biodiversity-SEBOT.R
accepted submission

@@ -1,0 +1,242 @@
+library(tidyverse); library(ggvenn); library(vegan)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
+
+### NMDS
+
+c("forestgreen",  
+  "gold", 
+  "#40E0D0",
+  "#A020F0",
+  "#5CACEE",
+  "#551A8B") -> colores
+
+### NMDS plant communities
+
+header %>%
+  select(habitat, id) %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 2) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
+  merge(header) %>%
+  dplyr::select(id, habitat, Dim.1, Dim.2) %>%
+  mutate(habitat = fct_relevel(habitat, "forest", "meadow", "park", "roadside", "residential", "industrial")) %>%
+  mutate(habitat = fct_recode(habitat, "BOSQUES" = "forest", "PRADOS" = "meadow",
+                              "PARQUES" = "park", "CUNETAS" = "roadside",
+                              "RESIDENCIALES" = "residential", "INDUSTRIALES" = "industrial")) -> sitescores
+
+### Species scores
+
+vegan::scores(nmds, "species") %>%
+  data.frame() %>%
+  rownames_to_column("Taxon") %>%
+  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
+  mutate(Species = Taxon) %>%
+  separate(Species, into = c("G", "S")) %>%
[…]

### 2025-09-30 src/6-novelty-R1-SEBOT.R

@@ -1,0 +1,323 @@
+library(tidyverse)
+
+read.csv("data/header.csv", fileEncoding = "latin1") -> header
+read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
+read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
+read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
+read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
+
+### Compositional novelty for plants
+
+header %>%
+  select(habitat, id) %>%
+  filter(habitat != "N4D") %>%
+  merge(plants) %>%
+  select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
+  spread(taxon, cover, fill = 0) %>%
+  column_to_rownames(var = "id") -> matrix 
+
+matrix %>%
+  vegan::metaMDS(trymax = 100000, k = 4) -> 
+  nmds # Ordination output
+
+### Site scores
+
+vegan::scores(nmds, "sites") %>%
+  data.frame() %>%
+  rownames_to_column("id") %>%
+  gather(Trait, Value, - id) %>%
+  merge(select(header, id, habitat)) -> sitescores
+
+### Novelty (versus forest)
+
+sitescores %>%
+  filter(habitat %in% c("forest")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = forest - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
+
+### Novelty (versus meadow)
+
+sitescores %>%
+  filter(habitat %in% c("meadow")) %>%
+  group_by(habitat, Trait) %>%
+  summarise(Value = mean(Value)) %>%
+  spread(habitat, Value) %>%
+  merge(sitescores) %>%
+  mutate(Novelty = meadow - Value) %>%
+  group_by(id, habitat) %>%
+  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
+
+### Compositional novelty for fungi
+
+fungi %>%
+  select(id, ASV, presence) %>%
+  spread(ASV, presence, fill = 0) %>%
[…]

### 2025-09-30 doc/manuscript.Rmd
cleanup after acceptance

@@
 output: 
-  bookdown::word_document2:
+  bookdown::github_document2:
     reference_docx: template.docx
@@
 Running title: Compositional novelty across urban habitats
+
+Eduardo Fernández-Pascual^1,2,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^3^ (<https://orcid.org/0000-0002-0915-0916>), Víctor González-García^1,2^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1,2^ (<https://orcid.org/0000-0001-6601-9597>)
+
+^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Asturias, Spain; ^2^ Department of Organismal and Systems Biology, University of Oviedo, Oviedo/Uviéu, Asturias, Spain; ^3^ Jardín Botánico Atlántico, Gijón/Xixón, Spain
+
+Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain.
+Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}.
+Telephone: +34985104781.
+
+# Author contributions
+
+Eduardo Fernández-Pascual: Conceptualization; Methodology; Investigation; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing.
+Zuzana Ferencova: Conceptualization; Methodology; Investigation; Data Curation; Writing – Review & Editing.
+Víctor González-García: Investigation; Visualization; Writing – Review & Editing; Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.
+
+# Acknowledgements
+
+María García Fernández, Sara Gutiérrez Cruz and Abelardo Sigüenza Solís helped with the field surveys.
+Thanks to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis and to INEA Laboratory (Valladolid, Spain) for soil physicochemical analysis.
+
+# Declaration of interest statement
+
+We have nothing to declare.
+
+# Funding
+
+The project is supported by the Biodiversity Foundation of the Ministry for the Ecological Transition and the Demographic Challenge (MITECO) of the Government of Spain, within the framework of the Recovery, Transformation and Resilience Plan (PRTR), funded by the European Union - NextGenerationEU.
+Project PID2024-156390NB-I00 funded by MICIU/ AEI / 10.13039/501100011033/ FEDER, UE.
+Project SV-23-GIJON-JBA funded by Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico.
+Grant "Laboratorio de Vegetación y Biodiversidad" (IDE/2024/000720, Principality of Asturias-Sekuens-EU-FEDER).
+Víctor González-García was supported by the Government of Asturias (Programa “Severo Ochoa” PA-22-BP21-098).
+
+# Data availability statement
+
+Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at Zenodo: [DOI WILL BE INSERTED HERE].
 

### 2025-09-30 src/5-biodiversity-SEBOT.R

@@ -1,242 +1,0 @@
-library(tidyverse); library(ggvenn); library(vegan)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
-
-### NMDS
-
-c("forestgreen",  
-  "gold", 
-  "#40E0D0",
-  "#A020F0",
-  "#5CACEE",
-  "#551A8B") -> colores
-
-### NMDS plant communities
-
-header %>%
-  select(habitat, id) %>%
-  merge(plants) %>%
-  select(id, taxon, cover) %>%
-  mutate(cover = 1) %>% # Transform cover to presence
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 2) -> 
-  nmds # Ordination output
-
-### Site scores
-
-vegan::scores(nmds, "sites") %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
-  merge(header) %>%
-  dplyr::select(id, habitat, Dim.1, Dim.2) %>%
-  mutate(habitat = fct_relevel(habitat, "forest", "meadow", "park", "roadside", "residential", "industrial")) %>%
-  mutate(habitat = fct_recode(habitat, "BOSQUES" = "forest", "PRADOS" = "meadow",
-                              "PARQUES" = "park", "CUNETAS" = "roadside",
-                              "RESIDENCIALES" = "residential", "INDUSTRIALES" = "industrial")) -> sitescores
-
-### Species scores
-
-vegan::scores(nmds, "species") %>%
-  data.frame() %>%
-  rownames_to_column("Taxon") %>%
-  rename(Dim.1 = NMDS1, Dim.2 = NMDS2) %>%
-  mutate(Species = Taxon) %>%
-  separate(Species, into = c("G", "S")) %>%
[…]

### 2025-09-30 src/6-novelty-R1-PCA.R

@@ -1,303 +1,0 @@
-library(tidyverse)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-
-### Compositional novelty for plants
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(plants) %>%
-  select(id, taxon, cover) %>%
-  mutate(cover = 1) %>% # Transform cover to presence
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") %>% 
-  decostand(method = "hellinger") -> df1
-
-df1 %>%
-  FactoMineR::PCA(graph = TRUE, scale.unit = FALSE) -> pca1
-
-### Site scores
-
-pca1$ind$coord %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  gather(Trait, Value, - id) %>%
-  merge(select(header, id, habitat)) -> sitescores
-
-### Novelty (versus forest)
-
-sitescores %>%
-  filter(habitat %in% c("forest")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = forest - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
-
-### Novelty (versus meadow)
-
-sitescores %>%
-  filter(habitat %in% c("meadow")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = meadow - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
-
-### Compositional novelty for fungi
-
-fungi %>%
-  select(id, ASV, presence) %>%
-  spread(ASV, presence, fill = 0) %>%
-  column_to_rownames(var = "id") %>% 
-  decostand(method = "hellinger") -> df1
-
-df1 %>%
[…]

### 2025-09-30 src/6-novelty-R1-SEBOT.R

@@ -1,323 +1,0 @@
-library(tidyverse)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
-
-### Compositional novelty for plants
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(plants) %>%
-  select(id, taxon, cover) %>%
-  mutate(cover = 1) %>% # Transform cover to presence
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 4) -> 
-  nmds # Ordination output
-
-### Site scores
-
-vegan::scores(nmds, "sites") %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  gather(Trait, Value, - id) %>%
-  merge(select(header, id, habitat)) -> sitescores
-
-### Novelty (versus forest)
-
-sitescores %>%
-  filter(habitat %in% c("forest")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = forest - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
-
-### Novelty (versus meadow)
-
-sitescores %>%
-  filter(habitat %in% c("meadow")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = meadow - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
-
-### Compositional novelty for fungi
-
-fungi %>%
-  select(id, ASV, presence) %>%
-  spread(ASV, presence, fill = 0) %>%
[…]

### 2025-09-30 src/6-novelty-R1.R

@@ -1,306 +1,0 @@
-library(tidyverse)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
-
-### Compositional novelty for plants
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(plants) %>%
-  select(id, taxon, cover) %>%
-  mutate(cover = 1) %>% # Transform cover to presence
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
-  vegan::metaMDS(trymax = 100000, k = 4) -> 
-  nmds # Ordination output
-
-### Site scores
-
-vegan::scores(nmds, "sites") %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  gather(Trait, Value, - id) %>%
-  merge(select(header, id, habitat)) -> sitescores
-
-### Novelty (versus forest)
-
-sitescores %>%
-  filter(habitat %in% c("forest")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = forest - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
-
-### Novelty (versus meadow)
-
-sitescores %>%
-  filter(habitat %in% c("meadow")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = meadow - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
-
-### Compositional novelty for fungi
-
-fungi %>%
-  select(id, ASV, presence) %>%
-  spread(ASV, presence, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-matrix %>%
[…]

### 2025-09-30 src/6-novelty-R2-PCoA.R

@@ -1,318 +1,0 @@
-library(tidyverse); library(vegan)
-
-read.csv("data/header.csv", fileEncoding = "latin1") -> header
-read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
-read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
-read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
-read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
-read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
-
-### Compositional novelty for plants
-
-header %>%
-  select(habitat, id) %>%
-  filter(habitat != "N4D") %>%
-  merge(plants) %>%
-  select(id, taxon, cover) %>%
-  mutate(cover = 1) %>% # Transform cover to presence
-  spread(taxon, cover, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
-### Distance matrix
-
-dist <- vegdist(matrix,  method = "bray")
-
-### PCoA
-
-PCoA <- wcmdscale(d = dist)
-
-### Site scores
-
-PCoA %>%
-  data.frame() %>%
-  rownames_to_column("id") %>%
-  gather(Trait, Value, - id) %>%
-  merge(select(header, id, habitat)) -> sitescores
-
-### Novelty (versus forest)
-
-sitescores %>%
-  filter(habitat %in% c("forest")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = forest - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.forest = sum(abs(Novelty)))  -> plant.forest
-
-### Novelty (versus meadow)
-
-sitescores %>%
-  filter(habitat %in% c("meadow")) %>%
-  group_by(habitat, Trait) %>%
-  summarise(Value = mean(Value)) %>%
-  spread(habitat, Value) %>%
-  merge(sitescores) %>%
-  mutate(Novelty = meadow - Value) %>%
-  group_by(id, habitat) %>%
-  summarise(plant.meadow = sum(abs(Novelty))) -> plant.meadow
-
-### Compositional novelty for fungi
-
-fungi %>%
-  select(id, ASV, presence) %>%
-  spread(ASV, presence, fill = 0) %>%
-  column_to_rownames(var = "id") -> matrix 
-
[…]

### 2025-09-30 src/6-novelty.R

@@ -5,6 +5,7 @@ read.csv("data/plants.csv", fileEncoding = "latin1") -> plants
 read.csv("data/fungi.csv", fileEncoding = "latin1") -> fungi
 read.csv("data/bacteria.csv", fileEncoding = "latin1") -> bacteria
 read.csv("data/soils.csv", fileEncoding = "latin1") -> soils
+read.csv("data/soil-pca.csv", fileEncoding = "latin1") -> pca
 
 ### Compositional novelty for plants
 
@@ -13,6 +14,7 @@ header %>%
   filter(habitat != "N4D") %>%
   merge(plants) %>%
   select(id, taxon, cover) %>%
+  mutate(cover = 1) %>% # Transform cover to presence
   spread(taxon, cover, fill = 0) %>%
   column_to_rownames(var = "id") -> matrix 
 
@@ -157,11 +159,11 @@ novelty %>%
   gather(Trait, Novelty, -c(id, habitat)) %>%
   mutate(Trait = fct_relevel(Trait, "plant.forest", "plant.meadow", "fungi.forest", "fungi.meadow", "bacterial.forest", "bacterial.meadow")) %>%
   mutate(Trait = fct_recode(Trait, "(E) Bacterial\n      novelty\n      vs. forests" = "bacterial.forest",
-                           "(C) Fungal\n      novelty\n      vs. forests" = "fungi.forest",
-                           "(A) Plant\n      novelty\n      vs. forests" = "plant.forest",
-                           "(F) Bacterial\n      novelty\n      vs. meadows" = "bacterial.meadow",
-                           "(D) Fungal\n      novelty\n      vs. meadows" = "fungi.meadow",
-                           "(B) Plant\n      novelty\n      vs. meadows" = "plant.meadow",)) %>%
+                            "(C) Fungal\n      novelty\n      vs. forests" = "fungi.forest",
+                            "(A) Plant\n      novelty\n      vs. forests" = "plant.forest",
+                            "(F) Bacterial\n      novelty\n      vs. meadows" = "bacterial.meadow",
+                            "(D) Fungal\n      novelty\n      vs. meadows" = "fungi.meadow",
+                            "(B) Plant\n      novelty\n      vs. meadows" = "plant.meadow",)) %>%
[…]

### 2025-09-30 README.Rmd
revised readme

@@
 This repository stores all the information related to the manuscript *Compositional novelty of plant, fungal and bacterial communities across urban habitats*, including the datasets, the scripts to perform data cleaning and analysis, and the `Rmarkdown` files to create the manuscript.
+
+# Abstract
+
+Understanding urban community novelty can help to predict the rewilding potential of vacant urban spaces, facilitating their integration into biodiverse cities.
+We tested the hypothesis that cities are composed of different degrees of ecological novelty by studying four urban habitats with differing degrees of management and human legacy: park lawns, roadsides, residential vacant lots, and industrial vacant lots.
+We focused on community compositional novelty, by comparing the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forests and meadows.
+We used a compositional novelty index based on multidimensional ordination, which is straightforward to calculate and only requires species co-occurrence data for urban and reference habitats.
+As expected, (1) plants displayed the highest novelty and bacteria the lowest; (2) urban communities were markedly different from forests and relatively more similar to meadows; and (3) compositional novelty was highest in industrial lots.
+Managed park lawns, which we had expected to be highly novel, were relatively close to hay meadows.
+The lowest novelty was recorded in residential vacant lots, which had biological communities that more closely resembled those of pre-urban habitats.
+Our results highlight the effect of habitat type as a major driver of urban community composition and novelty.
+This suggests that city biodiversity can be enhanced by an integrative approach to the urban landscape that favors habitat heterogeneity by passive rewilding of park lawns, non-intervention on residential vacant lots, direct restoration of industrial lots, and conservation of natural and agricultural habitat remnants as sources of native species.
+
+# Keywords
+
+Urban biodiversity, vegetation, soil fungi, soil bacteria, DNA metabarcoding, soil pollution, novel ecosystems
 
